### 2023-11-02 paper/main.tex
Paper, first commit.

%% Beginning of file 'sample631.tex'
%%
%% Modified 2021 March
%%
%% This is a sample manuscript marked up using the
%% AASTeX v6.31 LaTeX 2e macros.
%%
%% AASTeX is now based on Alexey Vikhlinin's emulateapj.cls 
%% (Copyright 2000-2015).  See the classfile for details.

%% AASTeX requires revtex4-1.cls and other external packages such as
%% latexsym, graphicx, amssymb, longtable, and epsf.  Note that as of 
%% Oct 2020, APS now uses revtex4.2e for its journals but remember that 
%% AASTeX v6+ still uses v4.1. All of these external packages should 
%% already be present in the modern TeX distributions but not always.
%% For example, revtex4.1 seems to be missing in the linux version of
%% TexLive 2020. One should be able to get all packages from www.ctan.org.
%% In particular, revtex v4.1 can be found at 
%% https://www.ctan.org/pkg/revtex4-1.

%% The first piece of markup in an AASTeX v6.x document is the \documentclass
%% command. LaTeX will ignore any data that comes before this command. The 
%% documentclass can take an optional argument to modify the output style.
%% The command below calls the preprint style which will produce a tightly 
%% typeset, one-column, single-spaced document.  It is the default and thus
%% does not need to be explicitly stated.
%%
%% using aastex version 6.3

%\documentclass[linenumbers]{aastex631}
\documentclass[linenumbers, twocolumn, trackchanges]{aastex631}

%% The default is a single spaced, 10 point font, single spaced article.
%% There are 5 other style options available via an optional argument. They
%% can be invoked like this:
%%
%% \documentclass[arguments]{aastex631}
%% 
%% where the layout options are:
%%
%%  twocolumn   : two text columns, 10 point font, single spaced article.
%%                This is the most compact and represent the final published
%%                derived PDF copy of the accepted manuscript from the publisher
%%  manuscript  : one text column, 12 point font, double spaced article.
%%  preprint    : one text column, 12 point font, single spaced article.  
%%  preprint2   : two text columns, 12 point font, single spaced article.
%%  modern      : a stylish, single text column, 12 point font, article with
%% 		  wider left and right margins. This uses the Daniel
%% 		  Foreman-Mackey and David Hogg design.
%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts 
%%                but now that abstracts are required this is obsolete for
%%                AAS Journals. Authors might need it for other reasons. DO NOT
%%                use \begin{abstract} and \end{abstract} with this style.
%%
%% Note that you can submit to the AAS Journals in any of these 6 styles.
%%
%% There are other optional arguments one can invoke to allow other stylistic
%% actions. The available options are:
%%
%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
%%   tighten      : Makes baselineskip slightly smaller, only works with 
%%                  the twocolumn substyle.
%%   times        : uses times font instead of the default
%%   linenumbers  : turn on lineno package.
%%   trackchanges : required to see the revision mark up and print its output
%%   longauthor   : Do not use the more compressed footnote style (default) for 
%%                  the author/collaboration/affiliations. Instead print all
%%                  affiliation information after each name. Creates a much 
%%                  longer author list but may be desirable for short 
%%                  author papers.
%% twocolappendix : make 2 column appendix.
%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
%%                  for dual anonymous review.
%%
%% these can be used in any combination, e.g.
%%
%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
%%
%% AASTeX v6.* now includes \hyperref support. While we have built in specific
%% defaults into the classfile you can manually override them with the
%% \hypersetup command. For example,
%%
%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
%%
%% will change the color of the internal links to red, the links to the
%% bibliography to green, the file links to cyan, and the external links to
%% magenta. Additional information on \hyperref options can be found here:
%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
%%
%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
%% to improve the accessibility of the compiled pdf file.
%%
%% If you want to create your own macros, you can do so
%% using \newcommand. Your macros should appear before
%% the \begin{document} command.
%%
\newcommand{\vdag}{(v)^\dagger}
\newcommand\aastex{AAS\TeX}
\newcommand\latex{La\TeX}
\newcommand{\magcom}[1]{\textcolor{magenta}{#1}} % {\textbf{#1}}}
\newcommand{\redcom}[1]{\textcolor{red}{#1}} % {\textbf{#1}}}
\newcommand{\yelcom}[1]{\colorbox{yellow}{#1}} % {\textbf{#1}}}
\newcommand{\pd}{\texttt{photoD}}
\newcommand{\mh}{[M/H]}
\newcommand{\Mb}{\textit{M\textsubscript{b}}}
\newcommand{\logg}{\textit{log(g)}}
\newcommand{\teff}{\textit{T\textsubscript{eff}}}

%% Reintroduced the \received and \accepted commands from AASTeX v5.2
%\received{March 1, 2021}
%\revised{April 1, 2021}
%\accepted{\today}

%% Command to document which AAS Journal the manuscript was submitted to.
%% Adds "Submitted to " the argument.
%\submitjournal{PSJ}

%% The following command can be used to set the latex table counters.  It
%% is needed in this document because it uses a mix of latex tabular and
%% AASTeX deluxetables.  In general it should not be needed.
%\setcounter{table}{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%% The following section outlines numerous optional output that
%% can be displayed in the front matter or as running meta-data.
%%
%% If you wish, you may supply running head information, although
%% this information may be modified by the editorial offices.
\shorttitle{PhotoD}
\shortauthors{Schwarz et al.}
%%
%% You can add a light gray and diagonal water-mark to the first page 
%% with this command:
%% \watermark{text}
%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
%% long you can control the water-mark size with:
%% \setwatermarkfontsize{dimension}
%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\graphicspath{{./}{figures/}}

\begin{document}

\title{PhotoD: LSST photometric distances out to 100 kpc.}

%% LaTeX will automatically break titles if they run longer than
%% one line. However, you may use \\ to force a line break if
%% you desire. In v6.31 you can include a footnote in the title.

%%
%% The \author command is the same as before except it now takes an optional
%% argument which is the 16 digit ORCID. The syntax is:
%% \author[xxxx-xxxx-xxxx-xxxx]{Author Name}

%% Use \affiliation for affiliation information. The old \affil is now aliased
%% to \affiliation. AASTeX v6.31 will automatically index these in the header.
%% When a duplicate is found its index will be the same as its previous entry.
%%
%% Note that \altaffilmark and \altaffiltext have been removed and thus 
%% can not be used to document secondary affiliations. If they are used latex
%% will issue a specific error message and quit. Please use multiple 
%% \affiliation calls for to document more than one affiliation.
%%
%% The new \altaffiliation can be used to indicate some secondary information
%% such as fellowships. This command produces a non-numeric footnote that is
%% set away from the numeric \affiliation footnotes.  NOTE that if an
%% \altaffiliation command is used it must come BEFORE the \affiliation call,
%% right after the \author command, in order to place the footnotes in
%% the proper location.
%%
%% Use \email to set provide email addresses. Each \email will appear on its
%% own line so you can put multiple email address in one \email call. A new
%% \correspondingauthor command is available in V6.31 to identify the
%% corresponding author of the manuscript. It is the author's responsibility
%% to make sure this name is also in the author list.
%%
%% While authors can be grouped inside the same \author and \affiliation
%% commands it is better to have a single author for each. This allows for
%% one to exploit all the new benefits and should make book-keeping easier.
%%
%% If done correctly the peer review system will be able to
%% automatically put the author and affiliation information from the manuscript
%% and save the corresponding author the trouble of entering it by hand.

\correspondingauthor{August Muench}
\email{greg.schwarz@aas.org, gus.muench@aas.org}

\author[0000-0002-0786-7307]{Greg J. Schwarz}
\affiliation{American Astronomical Society \\
1667 K Street NW, Suite 800 \\
Washington, DC 20006, USA}

\author{August Muench}
\affiliation{American Astronomical Society \\
1667 K Street NW, Suite 800 \\
Washington, DC 20006, USA}

\collaboration{6}{(AAS Journals Data Editors)}

\author{Butler Burton}
\affiliation{Leiden University}
\affiliation{AAS Journals Associate Editor-in-Chief}

\author{Amy Hendrickson}
\altaffiliation{AASTeX v6+ programmer}
\affiliation{TeXnology Inc.}

\author{Julie Steffen}
\affiliation{AAS Director of Publishing}
\affiliation{American Astronomical Society \\
1667 K Street NW, Suite 800 \\
Washington, DC 20006, USA}

\author{Magaret Donnelly}
\affiliation{IOP Publishing, Washington, DC 20005}

%% Mark off the abstract in the ``abstract'' environment. 
\begin{abstract}

This example manuscript is intended to serve as a tutorial and template for
authors to use when writing their own AAS Journal articles. The manuscript
includes a history of \aastex\ and documents the new features in the
previous versions as well as the bug fixes in version 6.31. This
manuscript includes many figure and table examples to illustrate these new
features.  Information on features not explicitly mentioned in the article
can be viewed in the manuscript comments or more extensive online
documentation. Authors are welcome replace the text, tables, figures, and
bibliography with their own and submit the resulting manuscript to the AAS
Journals peer review system.  The first lesson in the tutorial is to remind
authors that the AAS Journals, the Astrophysical Journal (ApJ), the
Astrophysical Journal Letters (ApJL), the Astronomical Journal (AJ), and
the Planetary Science Journal (PSJ) all have a 250 word limit for the 
abstract\footnote{Abstracts for Research Notes of the American Astronomical 
Society (RNAAS) are limited to 150 words}.  If you exceed this length the
Editorial office will ask you to shorten it. This abstract has 182 words.

\end{abstract}

%% Keywords should appear after the \end{abstract} command. 
%% The AAS Journals now uses Unified Astronomy Thesaurus concepts:
%% https://astrothesaurus.org
%% You will be asked to selected these concepts during the submission process
%% but this old "keyword" functionality is maintained in case authors want
%% to include these concepts in their preprints.
\keywords{Distance measure (395) --- Interstellar extinction (841) --- Photometry (1234) --- Stellar distance (1595) --- Two-color diagrams (1724)}

%% From the front matter, we move on to the body of the paper.
%% Sections are demarcated by \section and \subsection, respectively.
%% Observe the use of the LaTeX \label
%% command after the \subsection to give a symbolic KEY to the
%% subsection for cross-referencing in a \ref command.
%% You can use LaTeX's \ref and \label commands to keep track of
%% cross-references to sections, equations, tables, and figures.
%% That way, if you change the order of any elements, LaTeX will
%% automatically renumber them.
%%
%% We recommend that authors also use the natbib \citep
%% and \citet commands to identify citations.  The citations are
%% tied to the reference list via symbolic KEYs. The KEY corresponds
%% to the KEY in the \bibitem in the reference list below.

\section{Introduction} \label{sec:intro}

Thanks to the \textit{Vera C. Rubin} observatory's \textit{Legacy Survey of Space and Time} (LSST), for the first time in history, an astronomical catalog will contain more Milky Way stars than there are living people on Earth -- of the order 10-20 billion, depending on model assumptions. In order to map the Milky Way in three dimensions, distances to these stars must be accurately  estimated. In this paper we describe a method that will deliver LSST-based stellar distance estimations complementary to \textit{Gaia's} state-of-the-art trigonometric parallaxes and reach about 10 times further, to approximately 100 kpc. These results will be transformative for the studies of the Milky Way in general, and the stellar and the dark matter halo in particular as never before was there a survey that simultaneously observed roughly two thirds of the sky, to the co-added depth of \textit{r}$\approx$26 mag.

\magcom{A bit about the importance of the distance estimation in the MW, dust implications (for extragalactic science too).}

There are a variety of astronomical methods to estimate distances to stars, ranging from direct geometric (trigonometric) methods for nearby stars to indirect methods based on astrophysics for more distant stars.

\magcom{Mention \citet{bailer-jones_estimating_2021}, \cite{green_measuring_2014,green_three-dimensional_2015,green_3d_2019}, \citet{juric_milky_2008} and \citet{lallement_3d_2014}, \citet{queiroz_starhorse_2018}}.

Layout of the paper is...

\section{Method} \label{sec:method}

The photometric distance estimation method (hereafter \pd) is conceptually quite simple and relies on the strong correlations between the stellar colors and spectral energy distributions (SED) for dominant stellar populations. The stellar spectral energy distributions, and consequently colors, are determined by the effective temperature (\textit{T\textsubscript{eff}}), the surface gravity (\textit{g}), and the metallicity ([M/H]), or alternatively, by the absolute magnitude in band \textit{b} (\textit{M\textsubscript{b}}), [M/H] and age.

The distributions that describe these correlations are obtained either from models or from observations. For example, the distribution of stellar SEDs in the color-color diagram in Figure \ref{fig:2c_example} provides key insights in stellar evolution and classification of different stellar populations such as main-sequence stars, giant stars, white dwarf stars, a majority of unresolved binary stars and even extragalactic objects. Analogous distributions are responsible for the abundant structure seen in the Hertzsprung-Russell diagram (HRD).

\begin{figure}[ht!]
	\plotone{GOOZvV.png}
	\caption{Color-color diagram. \magcom{Add a version with overlaid evolutionary tracks in order to emphasize the required precision of the photometry required to disentangle the dwarfs and giants and different metallicities. \textbf{Are ev. tracks with different metallicities overlapping}? Can there be a situation where there is confusion between giants and dwarfs with different metallicities?} \label{fig:2c_example}}
\end{figure}

Metallicity is an important factor in these correlations, as it has a strong effect on the luminosity of the stars. This is reflected in the width of the main stellar loci of the color-magnitude diagrams (CMD) of the stellar populations observed at the same distance and the two-color diagrams (\magcom{quantify}), as seen in Figure \ref{fig:metallicity}. The best photometric estimators of metallicity are colors whose shorter-wavelength component includes the metal absorbtion bands at near-UV wavelengths, short of Balmer break. Therefore, the LSST has a comparative advantage over the surveys lacking \textit{u}-band measurements, and could provide accurate distances within the range of 5-10\%. \magcom{A plot of model spectra, fixed, \textit{log(g)} and \textit{T\textsubscript{eff}}, several different metallicities?}

\begin{figure}[ht!]
	\plotone{Fdtez8.png}
	\caption{Yadayada. \label{fig:metallicity}}
\end{figure}

Extinction is a major source of systematic errors in the process of luminosity and distance determination. The fact that the extinction vector is nearly parallel to the main stellar locus in the two-color diagrams gives rise to degeneracies that complicate the determination of the stellar type. An example is displayed in Fig. \ref{fig:degeneracies}, where in the left panel any of the different star types designated as 1,2 and 3 can have the same observed colors as the star marked as "Obs". This degeneracy is a result of the combination of colors chosen for the two-color diagram and depends on the adopted extinction curve \textit{R\textsubscript{V}} and the position on the stellar locus. The degeneracy can be broken if several different colors are used, particularly those towards the infrared, where the stellar locus is not kinked and the extinction vector is not parallel to it (as shown in the right panel of the Figure \ref{fig:degeneracies}).

\begin{figure}[ht!]
	\plotone{BYiR7o.png}
	\caption{Extinction \& degeneracies. \label{fig:degeneracies}}
\end{figure}

$L=4\pi\sigma R^2 T^4$\\
$g = \frac{GM}{R^2}$\\
$L=4\pi \sigma \frac{GM}{g} T^4$\\
$\mu = m_b - M_b {\ }(- A_b)$\\
$log(d) = 1 + \frac{\mu}{5}$\\

Given the observed colors, and assuming small errors, the estimation of stellar parameters is relatively straight forward and follows one of the two possible approaches: empirical or model-based. In the former case, the case the underlying distribution to which the observations are matched is obtained from (ideally) spectroscopic and photometric observations, while in the latter case the distributions are generated from population synthesis models.

\magcom{Advantages \& disadvantages of the model-based and empirical approaches, how model based approaches can be improved by adding empirical information for specific cases}.

% Here we provide a layout of our method that simultaneously provides distance estimates as well as the three-dimensional estimation of the dust distribution. The distance modulus of a star can be estimated given its absolute and apparent magnitudes and the extinction along the line of sight. Therefore, given the observed colors it is necessary to simultaneously estimate the effective temperature, metallicity and extinction.

\appendix

\bibliography{PhotoD}{}
\bibliographystyle{aasjournal}

%% Include this line if you are using the \added, \replaced, \deleted
%% commands to see a summary list of all changes at the end of the article.
%\listofchanges

\end{document}

### 2023-11-02 paper/main.tex
Update

@@
-%% Beginning of file 'sample631.tex'
-%%
-%% Modified 2021 March
-%%
-%% This is a sample manuscript marked up using the
-%% AASTeX v6.31 LaTeX 2e macros.
-%%
-%% AASTeX is now based on Alexey Vikhlinin's emulateapj.cls 
-%% (Copyright 2000-2015).  See the classfile for details.
-
-%% AASTeX requires revtex4-1.cls and other external packages such as
-%% latexsym, graphicx, amssymb, longtable, and epsf.  Note that as of 
-%% Oct 2020, APS now uses revtex4.2e for its journals but remember that 
-%% AASTeX v6+ still uses v4.1. All of these external packages should 
-%% already be present in the modern TeX distributions but not always.
-%% For example, revtex4.1 seems to be missing in the linux version of
-%% TexLive 2020. One should be able to get all packages from www.ctan.org.
-%% In particular, revtex v4.1 can be found at 
-%% https://www.ctan.org/pkg/revtex4-1.
-
-%% The first piece of markup in an AASTeX v6.x document is the \documentclass
-%% command. LaTeX will ignore any data that comes before this command. The 
-%% documentclass can take an optional argument to modify the output style.
-%% The command below calls the preprint style which will produce a tightly 
-%% typeset, one-column, single-spaced document.  It is the default and thus
-%% does not need to be explicitly stated.
-%%
-%% using aastex version 6.3
-
-%\documentclass[linenumbers]{aastex631}
 \documentclass[linenumbers, twocolumn, trackchanges]{aastex631}
@@
 %%
+
+\usepackage{xspace}
+
 \newcommand{\vdag}{(v)^\dagger}
@@
 \newcommand{\yelcom}[1]{\colorbox{yellow}{#1}} % {\textbf{#1}}}
-\newcommand{\pd}{\texttt{photoD}}
-\newcommand{\mh}{[M/H]}
-\newcommand{\Mb}{\textit{M\textsubscript{b}}}
-\newcommand{\logg}{\textit{log(g)}}
-\newcommand{\teff}{\textit{T\textsubscript{eff}}}
+\newcommand{\pd}{\texttt{photoD}\xspace}
+\newcommand{\mh}{[M/H]\xspace}
+\newcommand{\Mb}{\textit{M\textsubscript{b}}\xspace}
+\newcommand{\logg}{\textit{log(g)}\xspace}
+\newcommand{\teff}{\textit{T\textsubscript{eff}}\xspace}
+\newcommand{\RV}{\textit{R\textsubscript{V}}\xspace}
 
@@
 
-\magcom{Mention \citet{bailer-jones_estimating_2021}, \cite{green_measuring_2014,green_three-dimensional_2015,green_3d_2019}, \citet{juric_milky_2008} and \citet{lallement_3d_2014}, \citet{queiroz_starhorse_2018}}.
+\magcom{Mention \cite{bailer-jones_estimating_2021}, \cite{gordon_panchromatic_2016}, \cite{green_measuring_2014,green_three-dimensional_2015,green_3d_2019}, \cite{juric_milky_2008} and \cite{lallement_3d_2014}, \cite{queiroz_starhorse_2018}}.
 
@@
 This is reflected in the width of the main stellar loci of the color-magnitude diagrams (CMD) of the stellar populations observed at the same distance and the two-color diagrams (\magcom{quantify}), as seen in Figure \ref{fig:metallicity}.
-The best photometric estimators of metallicity are colors whose shorter-wavelength component includes the metal absorbtion bands at near-UV wavelengths, short of Balmer break.
+The best photometric estimators of metallicity are colors whose shorter-wavelength component includes the metal absorbtion bands at near-UV wavelengths, short of Balmer break (300 $\lessapprox\lambda$ [nm] $\lessapprox$ 400).
 Therefore, the LSST has a comparative advantage over the surveys lacking \textit{u}-band measurements, and could provide accurate distances within the range of 5-10\%.
@@
 
+\magcom{Explain the choice of \RV.}
+
 \begin{figure}[ht!]
@@
 
+Another important degeneracy arises from the fact that even for a fixed \teff and \mh, the \logg thus luminosity are not uniquely determined.
+A way to mitigate this issue is to adopt a Bayesian framework
+
+overlap in the color-color diagrams (e.g. Figure \ref{fig:2c_example}). This effectively means that for a fixed \teff proxy ($g-r$),
+
 Given the observed colors, and assuming small errors, the estimation of stellar parameters is relatively straight forward and follows one of the two possible approaches: empirical or model-based.
@@
 \magcom{Advantages \& disadvantages of the model-based and empirical approaches, how model based approaches can be improved by adding empirical information for specific cases}.
+
+\magcom{\cite{gordon_panchromatic_2016} na \href{https://beast.readthedocs.io/en/latest/beast\_graphical\_model.html}{BEAST webu} imaju zgodne diagrame; možda bi i mi mogli nešto tog tipa napraviti, bar za draft, primjer }
+
+\begin{figure}[ht!]
+	\plottwo{beast-graphic-text.png}{beast-graphic-math.png}
+	\caption{BEAST. \label{fig:BEAST}}
+\end{figure}
+
+% -- modeli za boje moraju biti savršeni
+% -- novi prior iz trilegala
+% -- robustni framework koji može obraditi 10B zvijezda, i da nije spor
+% -- probali smo s MCMC, exhaustive locus search, i neuronske mreže koje su bile najbrže (drugi članak)
+
+\section{Testing}
+
+% -- aginst Rich et al. BDSD survey ugriz podaci
+% -- Bailer-Jones i S82
+% -- Trilegal simulacije s SDSS SED modelima (validacija)
+% -- kuglasti skupovi
+
+\section{Discussion}
+
+% -- popraviti modele
+% -- dovući model s mladim zvijezdama
+% -- dodati modele za WD, BHB, unresolved binaries, RRL i druge populacije
 

### 2023-11-04 src/BayesTools.py
added dumpPriorMaps code

@@ -2,6 +2,7 @@ import numpy as np
 from scipy.interpolate import griddata
 import matplotlib.pyplot as plt
 from matplotlib.colors import LogNorm
+from scipy.stats import gaussian_kde
 import LocusTools as lt 
 import PlotTools as pt
 
@@ -10,7 +11,7 @@ import PlotTools as pt
 def getBayesConstants():
     
     BayesConst = {}
-    # parameters for stellar locus (main sequence and red giants) table
+    # parameters for stellar locus (main sequence and red giants) table, as well as priors 
     BayesConst['FeHmin'] = -2.5       # the range of considered [Fe/H], in priors and elsewhere
     BayesConst['FeHmax'] = 1.0
     BayesConst['FeHNpts'] = 36        # defines the grid step for [Fe/H]
@@ -35,7 +36,7 @@ def getBayesConstants():
 ### IMPLEMENTATION OF MAP-BASED PRIORS
 def getMetadataPriors(priorMap=""):
     if (priorMap==""):
-        bc = bt.getBayesConstants()
+        bc = getBayesConstants()
         FeHmin = bc['FeHmin']
         FeHmin = bc['FeHmin']
         FeHNpts = bc['FeHNpts']
@@ -69,11 +70,15 @@ def get2Dmap(sample, labels, metadata):
     Z = kde.evaluate(np.vstack([Xgrid.ravel(), Ygrid.ravel()]))
     return (Xgrid, Ygrid, Z)
 
-def dumpPriorMaps(sample, rmagMin, rmagMax, rmagNsteps, fileRootname):
+def dumpPriorMaps(sample, fileRootname):
+
+    ## data frame called "sample" here must have the following columns: 'FeH', 'Mr', 'rmag'
+    labels = ['FeH', 'Mr', 'rmag']
+
     ## numerical model specifications and constants for Bayesian PhotoD method
-    bc = bt.getBayesConstants()
+    bc = getBayesConstants()
     FeHmin = bc['FeHmin']
-    FeHmin = bc['FeHmin']
+    FeHmax = bc['FeHmax']
     FeHNpts = bc['FeHNpts']
     MrFaint = bc['MrFaint']
     MrBright = bc['MrBright'] 
@@ -84,7 +89,6 @@ def dumpPriorMaps(sample, rmagMin, rmagMax, rmagNsteps, fileRootname):
     rmagNsteps = bc['rmagNsteps']  
 
     # -------
[…]

### 2023-11-06 paper/main.tex
Added text to Method section.

@@
 \usepackage{xspace}
+\usepackage{amsmath}
 
@@
 \newcommand{\mh}{[M/H]\xspace}
-\newcommand{\Mb}{\textit{M\textsubscript{b}}\xspace}
+\newcommand{\mhm}{[\mathrm{M} / \mathrm{H}]\xspace}
+\newcommand{\Mb}{\ensuremath{M{\mathrm{_b}}}\xspace}
+\newcommand{\Ar}{\ensuremath{A{\mathrm{_r}}}\xspace}
+
+% \newcommand{\Mb}{\textit{M\textsubscript{b}}\xspace}
 \newcommand{\logg}{\textit{log(g)}\xspace}
@@
 \newcommand{\RV}{\textit{R\textsubscript{V}}\xspace}
+%\newcommand{\AV}{\textit{A\textsubscript{V}}\xspace}
+
+\newcommand{\ul}{\ensuremath{u}\xspace}
+\newcommand{\gl}{\ensuremath{g}\xspace}
+\newcommand{\rl}{\ensuremath{r}\xspace}
+\newcommand{\il}{\ensuremath{i}\xspace}
+\newcommand{\zl}{\ensuremath{z}\xspace}
+\newcommand{\yl}{\ensuremath{y}\xspace}
+
+\renewcommand{\arg}[1]{\! \left( #1 \right)}
+\newcommand{\Pcond}[2]{P \arg{ #1 \, | \, #2 }}
+% \newcommand{\yps}{\ensuremath{y_{\mathrm{P1}}}\xspace}
+%\newcommand{\EBV}{\ensuremath{E \left( B \! - \! V \right)}\xspace}
+%\newcommand{\Egr}{\ensuremath{E \left( g \! - \! r \right)}\xspace}
+%\newcommand{\EBPRP}{\ensuremath{E \left( BP \! - \! RP \right)}\xspace}
 
@@
 
-Extinction is a major source of systematic errors in the process of luminosity and distance determination.
+Extinction is another major source of systematic errors in the process of luminosity and distance determination.
 The fact that the extinction vector is nearly parallel to the main stellar locus in the two-color diagrams gives rise to degeneracies that complicate the determination of the stellar type.
@@
 \ref{fig:degeneracies}, where in the left panel any of the different star types designated as 1,2 and 3 can have the same observed colors as the star marked as "Obs".
-This degeneracy is a result of the combination of colors chosen for the two-color diagram and depends on the adopted extinction curve \textit{R\textsubscript{V}} and the position on the stellar locus.
-The degeneracy can be broken if several different colors are used, particularly those towards the infrared, where the stellar locus is not kinked and the extinction vector is not parallel to it (as shown in the right panel of the Figure \ref{fig:degeneracies}).
+This degeneracy is a result of the combination of colors chosen for the two-color diagram and depends on the position on the stellar locus and the adopted extinction curve parametrized by a single parameter \textit{R\textsubscript{V}}
+
+\begin{equation}
+	R_V = \frac{A_V}{E(B-V)},
+\end{equation}
+
+where $A_V$ stands for extinction in \textit{V}-band and \textit{E(B-V)} is the color excess. This relationship can be extended to an arbitrary photometric bandpass $\lambda$:
+
+\begin{equation}
+	A_{\lambda} = C_{\lambda}(R_V)A_r,
+\end{equation}
+
+with $A_r$ designating extinction in \textit{r}-band and $C_{\lambda}(R_V)$ describing the shape of the extinction curve.
+The degeneracy from the left panel in Fig.
+\ref{fig:degeneracies} can be broken if several different colors are used, particularly those towards the infrared, where the stellar locus is not kinked and the extinction vector is not parallel to it (as shown in the right panel of the Figure \ref{fig:degeneracies}), where $r-i$ and $i-z$ colors are used, and assuming a fixed extinction law a unique solution for the extinction is possible.
 
@@
 
-$L=4\pi\sigma R^2 T^4$\\
-$g = \frac{GM}{R^2}$\\
-$L=4\pi \sigma \frac{GM}{g} T^4$\\
-$\mu = m_b - M_b {\ }(- A_b)$\\
-$log(d) = 1 + \frac{\mu}{5}$\\
+Another important degeneracy arises from the fact that even for a fixed \teff and \mh, the \logg and thus the luminosity are not uniquely determined by the colors: a degeneracy may exist between the giant branch and the main sequence as the colors constructed from \textit{ugrizy} bands are not sensitive to \logg (\magcom{As evident from ŽI's gr-ri plot.}).
+We treat this issue by adopting a prior based on bins in apparent magnitude.
 
-Another important degeneracy arises from the fact that even for a fixed \teff and \mh, the \logg thus luminosity are not uniquely determined.
-A way to mitigate this issue is to adopt a Bayesian framework
+We adopt a Bayesian framework in which we simultaneously fit for \Mb, \mh and \Ar, assuming a fixed \RV value of 3.1\footnote{In principle \RV could be also fitted for.} The posterior for each individual star with LSST photometry is then given as:
 
-overlap in the color-color diagrams (e.g. Figure \ref{fig:2c_example}). This effectively means that for a fixed \teff proxy ($g-r$),
+\begin{equation} \label{eq:likelihood}
+	\begin{split}
+		    &\Pcond{\Mb, \, \mh, \, \Ar}{\vec{c}} = \\
+		& \frac{\Pcond{\vec{c}}{\Mb, \, \mh, \, \Ar} P\arg{\Mb, \, \mh, \, \Ar}}{P\arg{\vec{c}}} \,
+	\end{split}
+\end{equation}
 
-Given the observed colors, and assuming small errors, the estimation of stellar parameters is relatively straight forward and follows one of the two possible approaches: empirical or model-based.
-In the former case, the case the underlying distribution to which the observations are matched is obtained from (ideally) spectroscopic and photometric observations, while in the latter case the distributions are generated from population synthesis models.
+with \ensuremath{\vec{c}} standing for the vector of input colors (\ul-\gl, \gl-\rl and so on). The log-likelihood is given by:
+
+\begin{equation}
+	ln \mathcal{L} = {-1 \over \sqrt{2\pi}} \,\sum_{i=1}^N \left({ c^{obs}_i - c^{mod}_i  \over \sigma_i } \right)^2 \,
+\end{equation}
+
+where \ensuremath{c^{obs}_i} are the observed colors and \ensuremath{c^{mod}_i} model colors.
+The values of model colors and the priors are extracted from from TRILEGAL (\citet{dal_tio_simulating_2022}).
+In order to extract the priors (i.e.
+prior maps), we divide TRILEGAL data in healpix bins, and further subdivide them in one-magnitude wide bins in apparent magnitude.
+The latter subdivision is helpful in breaking the degeneracies between the giant and dwarf stars, as intrinsically luminous stars become strongly disfavored at faint magnitudes\footnote{In other words, an apparently faint giant star would imply a very large distance.
+For example a moderately bright giant star with \ensuremath{M_r}=0 mag and \ensuremath{r}=22 mag would imply a distance modulus of 22 mag, or distance of approximately 251 kpc.}.
+
+\magcom{Add plots describing the method, and go through one example like in Željko's slides. }
+
+Our method is basically brute-force fitting with some intelligent tricks leveraged to obtain faster execution that will be required for 10B LSST stars.
+We use \cite{schlegel_maps_1998} (SFD98) maps in order to limit the range of the extinction.
+This is usually a valid assumption because the SFD98 maps provide \textit{total} extinction along a line of sight.
+Our fitting procedure is also executed on an adaptive grid, a coarse search over the parameter space is performed first in order to establish the layout of the manifold.
+However, care is taken that any possible local minima are not missed by appropriately adjusting the step size \magcom{how?}.
+The located maxima are then explored with a smaller step size (\magcom{adjusted how?}).
+
+In addition to the approach described here, we also tested Markov Chain Monte Carlo and neural network approaches that will be/are described in forthcoming/published papers.
+
+%Under the assumptions that the locus of the color-color diagram is representative of the stellar SEDs, that the shape of the extinction curve can be parametrized with a single parameter \RV
 
@@
 % -- probali smo s MCMC, exhaustive locus search, i neuronske mreže koje su bile najbrže (drugi članak)
+
+%$L=4\pi\sigma R^2 T^4=4\pi \sigma \frac{GM}{g} T^4$\\
+%$g = \frac{GM}{R^2}$\\
+%$\mu = m_b - M_b {\ }(+ A_b)$\\
+%$log(d) = 1 + \frac{\mu}{5}$\\
 

### 2023-11-06 paper/analysis.tex
individual files

--- 2023-11-02 paper/main.tex
@@
-\documentclass[linenumbers, twocolumn, trackchanges]{aastex631}
-
-%% The default is a single spaced, 10 point font, single spaced article.
-%% There are 5 other style options available via an optional argument. They
-%% can be invoked like this:
-%%
-%% \documentclass[arguments]{aastex631}
-%% 
-%% where the layout options are:
-%%
-%%  twocolumn   : two text columns, 10 point font, single spaced article.
-%%                This is the most compact and represent the final published
-%%                derived PDF copy of the accepted manuscript from the publisher
-%%  manuscript  : one text column, 12 point font, double spaced article.
-%%  preprint    : one text column, 12 point font, single spaced article.  
-%%  preprint2   : two text columns, 12 point font, single spaced article.
-%%  modern      : a stylish, single text column, 12 point font, article with
-%% 		  wider left and right margins. This uses the Daniel
-%% 		  Foreman-Mackey and David Hogg design.
-%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts 
-%%                but now that abstracts are required this is obsolete for
-%%                AAS Journals. Authors might need it for other reasons. DO NOT
-%%                use \begin{abstract} and \end{abstract} with this style.
-%%
-%% Note that you can submit to the AAS Journals in any of these 6 styles.
-%%
-%% There are other optional arguments one can invoke to allow other stylistic
-%% actions. The available options are:
-%%
-%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
-%%   tighten      : Makes baselineskip slightly smaller, only works with 
-%%                  the twocolumn substyle.
-%%   times        : uses times font instead of the default
-%%   linenumbers  : turn on lineno package.
-%%   trackchanges : required to see the revision mark up and print its output
-%%   longauthor   : Do not use the more compressed footnote style (default) for 
-%%                  the author/collaboration/affiliations. Instead print all
-%%                  affiliation information after each name. Creates a much 
-%%                  longer author list but may be desirable for short 
-%%                  author papers.
-%% twocolappendix : make 2 column appendix.
-%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
-%%                  for dual anonymous review.
-%%
-%% these can be used in any combination, e.g.
-%%
-%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
-%%
-%% AASTeX v6.* now includes \hyperref support. While we have built in specific
-%% defaults into the classfile you can manually override them with the
-%% \hypersetup command. For example,
-%%
-%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
-%%
-%% will change the color of the internal links to red, the links to the
-%% bibliography to green, the file links to cyan, and the external links to
-%% magenta. Additional information on \hyperref options can be found here:
-%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
-%%
-%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
-%% to improve the accessibility of the compiled pdf file.
-%%
-%% If you want to create your own macros, you can do so
-%% using \newcommand. Your macros should appear before
-%% the \begin{document} command.
-%%
-
-\usepackage{xspace}
-
-\newcommand{\vdag}{(v)^\dagger}
-\newcommand\aastex{AAS\TeX}
-\newcommand\latex{La\TeX}
-\newcommand{\magcom}[1]{\textcolor{magenta}{#1}} % {\textbf{#1}}}
-\newcommand{\redcom}[1]{\textcolor{red}{#1}} % {\textbf{#1}}}
-\newcommand{\yelcom}[1]{\colorbox{yellow}{#1}} % {\textbf{#1}}}
-\newcommand{\pd}{\texttt{photoD}\xspace}
-\newcommand{\mh}{[M/H]\xspace}
-\newcommand{\Mb}{\textit{M\textsubscript{b}}\xspace}
-\newcommand{\logg}{\textit{log(g)}\xspace}
-\newcommand{\teff}{\textit{T\textsubscript{eff}}\xspace}
-\newcommand{\RV}{\textit{R\textsubscript{V}}\xspace}
-
-%% Reintroduced the \received and \accepted commands from AASTeX v5.2
-%\received{March 1, 2021}
-%\revised{April 1, 2021}
-%\accepted{\today}
-
-%% Command to document which AAS Journal the manuscript was submitted to.
-%% Adds "Submitted to " the argument.
-%\submitjournal{PSJ}
-
-%% The following command can be used to set the latex table counters.  It
-%% is needed in this document because it uses a mix of latex tabular and
-%% AASTeX deluxetables.  In general it should not be needed.
-%\setcounter{table}{1}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%
-%% The following section outlines numerous optional output that
-%% can be displayed in the front matter or as running meta-data.
-%%
-%% If you wish, you may supply running head information, although
-%% this information may be modified by the editorial offices.
-\shorttitle{PhotoD}
-\shortauthors{Schwarz et al.}
-%%
-%% You can add a light gray and diagonal water-mark to the first page 
-%% with this command:
-%% \watermark{text}
-%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
-%% long you can control the water-mark size with:
-%% \setwatermarkfontsize{dimension}
-%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
-%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\graphicspath{{./}{figures/}}
-
-\begin{document}
-
-\title{PhotoD: LSST photometric distances out to 100 kpc.}
-
-%% LaTeX will automatically break titles if they run longer than
-%% one line. However, you may use \\ to force a line break if
-%% you desire. In v6.31 you can include a footnote in the title.
-
-%%
-%% The \author command is the same as before except it now takes an optional
-%% argument which is the 16 digit ORCID. The syntax is:
-%% \author[xxxx-xxxx-xxxx-xxxx]{Author Name}
-
-%% Use \affiliation for affiliation information. The old \affil is now aliased
-%% to \affiliation. AASTeX v6.31 will automatically index these in the header.
-%% When a duplicate is found its index will be the same as its previous entry.
-%%
-%% Note that \altaffilmark and \altaffiltext have been removed and thus 
-%% can not be used to document secondary affiliations. If they are used latex
-%% will issue a specific error message and quit. Please use multiple 
-%% \affiliation calls for to document more than one affiliation.
-%%
-%% The new \altaffiliation can be used to indicate some secondary information
-%% such as fellowships. This command produces a non-numeric footnote that is
-%% set away from the numeric \affiliation footnotes.  NOTE that if an
-%% \altaffiliation command is used it must come BEFORE the \affiliation call,
-%% right after the \author command, in order to place the footnotes in
-%% the proper location.
-%%
-%% Use \email to set provide email addresses. Each \email will appear on its
-%% own line so you can put multiple email address in one \email call. A new
-%% \correspondingauthor command is available in V6.31 to identify the
-%% corresponding author of the manuscript. It is the author's responsibility
-%% to make sure this name is also in the author list.
-%%
-%% While authors can be grouped inside the same \author and \affiliation
-%% commands it is better to have a single author for each. This allows for
-%% one to exploit all the new benefits and should make book-keeping easier.
-%%
-%% If done correctly the peer review system will be able to
-%% automatically put the author and affiliation information from the manuscript
-%% and save the corresponding author the trouble of entering it by hand.
-
-\correspondingauthor{August Muench}
-\email{greg.schwarz@aas.org, gus.muench@aas.org}
-
-\author[0000-0002-0786-7307]{Greg J. Schwarz}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\author{August Muench}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\collaboration{6}{(AAS Journals Data Editors)}
-
-\author{Butler Burton}
-\affiliation{Leiden University}
-\affiliation{AAS Journals Associate Editor-in-Chief}
-
-\author{Amy Hendrickson}
-\altaffiliation{AASTeX v6+ programmer}
-\affiliation{TeXnology Inc.}
-
-\author{Julie Steffen}
-\affiliation{AAS Director of Publishing}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\author{Magaret Donnelly}
-\affiliation{IOP Publishing, Washington, DC 20005}
-
-%% Mark off the abstract in the ``abstract'' environment. 
-\begin{abstract}
-
-This example manuscript is intended to serve as a tutorial and template for
-authors to use when writing their own AAS Journal articles. The manuscript
-includes a history of \aastex\ and documents the new features in the
-previous versions as well as the bug fixes in version 6.31. This
-manuscript includes many figure and table examples to illustrate these new
-features.  Information on features not explicitly mentioned in the article
-can be viewed in the manuscript comments or more extensive online
-documentation. Authors are welcome replace the text, tables, figures, and
-bibliography with their own and submit the resulting manuscript to the AAS
-Journals peer review system.  The first lesson in the tutorial is to remind
-authors that the AAS Journals, the Astrophysical Journal (ApJ), the
-Astrophysical Journal Letters (ApJL), the Astronomical Journal (AJ), and
-the Planetary Science Journal (PSJ) all have a 250 word limit for the 
-abstract\footnote{Abstracts for Research Notes of the American Astronomical 
-Society (RNAAS) are limited to 150 words}.  If you exceed this length the
-Editorial office will ask you to shorten it. This abstract has 182 words.
-
-\end{abstract}
-
-%% Keywords should appear after the \end{abstract} command. 
-%% The AAS Journals now uses Unified Astronomy Thesaurus concepts:
-%% https://astrothesaurus.org
-%% You will be asked to selected these concepts during the submission process
-%% but this old "keyword" functionality is maintained in case authors want
-%% to include these concepts in their preprints.
-\keywords{Distance measure (395) --- Interstellar extinction (841) --- Photometry (1234) --- Stellar distance (1595) --- Two-color diagrams (1724)}
-
-%% From the front matter, we move on to the body of the paper.
-%% Sections are demarcated by \section and \subsection, respectively.
-%% Observe the use of the LaTeX \label
-%% command after the \subsection to give a symbolic KEY to the
-%% subsection for cross-referencing in a \ref command.
-%% You can use LaTeX's \ref and \label commands to keep track of
-%% cross-references to sections, equations, tables, and figures.
-%% That way, if you change the order of any elements, LaTeX will
-%% automatically renumber them.
-%%
-%% We recommend that authors also use the natbib \citep
-%% and \citet commands to identify citations.  The citations are
-%% tied to the reference list via symbolic KEYs. The KEY corresponds
-%% to the KEY in the \bibitem in the reference list below.
-
-\section{Introduction} \label{sec:intro}
-
-Thanks to the \textit{Vera C.
-Rubin} observatory's \textit{Legacy Survey of Space and Time} (LSST), for the first time in history, an astronomical catalog will contain more Milky Way stars than there are living people on Earth -- of the order 10-20 billion, depending on model assumptions.
-In order to map the Milky Way in three dimensions, distances to these stars must be accurately  estimated.
-In this paper we describe a method that will deliver LSST-based stellar distance estimations complementary to \textit{Gaia's} state-of-the-art trigonometric parallaxes and reach about 10 times further, to approximately 100 kpc.
-These results will be transformative for the studies of the Milky Way in general, and the stellar and the dark matter halo in particular as never before was there a survey that simultaneously observed roughly two thirds of the sky, to the co-added depth of \textit{r}$\approx$26 mag.
-
-\magcom{A bit about the importance of the distance estimation in the MW, dust implications (for extragalactic science too).}
-
-There are a variety of astronomical methods to estimate distances to stars, ranging from direct geometric (trigonometric) methods for nearby stars to indirect methods based on astrophysics for more distant stars.
-
-\magcom{Mention \cite{bailer-jones_estimating_2021}, \cite{gordon_panchromatic_2016}, \cite{green_measuring_2014,green_three-dimensional_2015,green_3d_2019}, \cite{juric_milky_2008} and \cite{lallement_3d_2014}, \cite{queiroz_starhorse_2018}}.
-
-Layout of the paper is...
-
-\section{Method} \label{sec:method}
-
-The photometric distance estimation method (hereafter \pd) is conceptually quite simple and relies on the strong correlations between the stellar colors and spectral energy distributions (SED) for dominant stellar populations.
-The stellar spectral energy distributions, and consequently colors, are determined by the effective temperature (\textit{T\textsubscript{eff}}), the surface gravity (\textit{g}), and the metallicity ([M/H]), or alternatively, by the absolute magnitude in band \textit{b} (\textit{M\textsubscript{b}}), [M/H] and age.
-
-The distributions that describe these correlations are obtained either from models or from observations.
-For example, the distribution of stellar SEDs in the color-color diagram in Figure \ref{fig:2c_example} provides key insights in stellar evolution and classification of different stellar populations such as main-sequence stars, giant stars, white dwarf stars, a majority of unresolved binary stars and even extragalactic objects.
-Analogous distributions are responsible for the abundant structure seen in the Hertzsprung-Russell diagram (HRD).
-
-\begin{figure}[ht!]
-	\plotone{GOOZvV.png}
-	\caption{Color-color diagram.
-\magcom{Add a version with overlaid evolutionary tracks in order to emphasize the required precision of the photometry required to disentangle the dwarfs and giants and different metallicities.
-\textbf{Are ev.
-tracks with different metallicities overlapping}?
-Can there be a situation where there is confusion between giants and dwarfs with different metallicities?} \label{fig:2c_example}}
-\end{figure}
-
-Metallicity is an important factor in these correlations, as it has a strong effect on the luminosity of the stars.
-This is reflected in the width of the main stellar loci of the color-magnitude diagrams (CMD) of the stellar populations observed at the same distance and the two-color diagrams (\magcom{quantify}), as seen in Figure \ref{fig:metallicity}.
-The best photometric estimators of metallicity are colors whose shorter-wavelength component includes the metal absorbtion bands at near-UV wavelengths, short of Balmer break (300 $\lessapprox\lambda$ [nm] $\lessapprox$ 400).
-Therefore, the LSST has a comparative advantage over the surveys lacking \textit{u}-band measurements, and could provide accurate distances within the range of 5-10\%.
-\magcom{A plot of model spectra, fixed, \textit{log(g)} and \textit{T\textsubscript{eff}}, several different metallicities?}
-
-\begin{figure}[ht!]
-	\plotone{Fdtez8.png}
-	\caption{Yadayada. \label{fig:metallicity}}
-\end{figure}
-
-Extinction is a major source of systematic errors in the process of luminosity and distance determination.
-The fact that the extinction vector is nearly parallel to the main stellar locus in the two-color diagrams gives rise to degeneracies that complicate the determination of the stellar type.
-An example is displayed in Fig.
-\ref{fig:degeneracies}, where in the left panel any of the different star types designated as 1,2 and 3 can have the same observed colors as the star marked as "Obs".
-This degeneracy is a result of the combination of colors chosen for the two-color diagram and depends on the adopted extinction curve \textit{R\textsubscript{V}} and the position on the stellar locus.
-The degeneracy can be broken if several different colors are used, particularly those towards the infrared, where the stellar locus is not kinked and the extinction vector is not parallel to it (as shown in the right panel of the Figure \ref{fig:degeneracies}).
-
-\magcom{Explain the choice of \RV.}
-
-\begin{figure}[ht!]
-	\plotone{BYiR7o.png}
-	\caption{Extinction \& degeneracies. \label{fig:degeneracies}}
-\end{figure}
-
-$L=4\pi\sigma R^2 T^4$\\
-$g = \frac{GM}{R^2}$\\
-$L=4\pi \sigma \frac{GM}{g} T^4$\\
-$\mu = m_b - M_b {\ }(- A_b)$\\
-$log(d) = 1 + \frac{\mu}{5}$\\
-
-Another important degeneracy arises from the fact that even for a fixed \teff and \mh, the \logg thus luminosity are not uniquely determined.
-A way to mitigate this issue is to adopt a Bayesian framework
-
-overlap in the color-color diagrams (e.g. Figure \ref{fig:2c_example}). This effectively means that for a fixed \teff proxy ($g-r$),
-
-Given the observed colors, and assuming small errors, the estimation of stellar parameters is relatively straight forward and follows one of the two possible approaches: empirical or model-based.
-In the former case, the case the underlying distribution to which the observations are matched is obtained from (ideally) spectroscopic and photometric observations, while in the latter case the distributions are generated from population synthesis models.
-
-\magcom{Advantages \& disadvantages of the model-based and empirical approaches, how model based approaches can be improved by adding empirical information for specific cases}.
-
-\magcom{\cite{gordon_panchromatic_2016} na \href{https://beast.readthedocs.io/en/latest/beast\_graphical\_model.html}{BEAST webu} imaju zgodne diagrame; možda bi i mi mogli nešto tog tipa napraviti, bar za draft, primjer }
-
-\begin{figure}[ht!]
-	\plottwo{beast-graphic-text.png}{beast-graphic-math.png}
-	\caption{BEAST. \label{fig:BEAST}}
-\end{figure}
-
-% -- modeli za boje moraju biti savršeni
-% -- novi prior iz trilegala
-% -- robustni framework koji može obraditi 10B zvijezda, i da nije spor
-% -- probali smo s MCMC, exhaustive locus search, i neuronske mreže koje su bile najbrže (drugi članak)
-
-\section{Testing}
  
@@
 % -- kuglasti skupovi
-
-\section{Discussion}
-
-% -- popraviti modele
-% -- dovući model s mladim zvijezdama
-% -- dodati modele za WD, BHB, unresolved binaries, RRL i druge populacije
-
-% Here we provide a layout of our method that simultaneously provides distance estimates as well as the three-dimensional estimation of the dust distribution.
-The distance modulus of a star can be estimated given its absolute and apparent magnitudes and the extinction along the line of sight.
-Therefore, given the observed colors it is necessary to simultaneously estimate the effective temperature, metallicity and extinction.
-
-\appendix
-
-\bibliography{PhotoD}{}
-\bibliographystyle{aasjournal}
-
-%% Include this line if you are using the \added, \replaced, \deleted
-%% commands to see a summary list of all changes at the end of the article.
-%\listofchanges
-
-\end{document}

### 2023-11-06 paper/discussion.tex
individual files

--- 2023-11-02 paper/main.tex
@@
-\documentclass[linenumbers, twocolumn, trackchanges]{aastex631}
-
-%% The default is a single spaced, 10 point font, single spaced article.
-%% There are 5 other style options available via an optional argument. They
-%% can be invoked like this:
-%%
-%% \documentclass[arguments]{aastex631}
-%% 
-%% where the layout options are:
-%%
-%%  twocolumn   : two text columns, 10 point font, single spaced article.
-%%                This is the most compact and represent the final published
-%%                derived PDF copy of the accepted manuscript from the publisher
-%%  manuscript  : one text column, 12 point font, double spaced article.
-%%  preprint    : one text column, 12 point font, single spaced article.  
-%%  preprint2   : two text columns, 12 point font, single spaced article.
-%%  modern      : a stylish, single text column, 12 point font, article with
-%% 		  wider left and right margins. This uses the Daniel
-%% 		  Foreman-Mackey and David Hogg design.
-%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts 
-%%                but now that abstracts are required this is obsolete for
-%%                AAS Journals. Authors might need it for other reasons. DO NOT
-%%                use \begin{abstract} and \end{abstract} with this style.
-%%
-%% Note that you can submit to the AAS Journals in any of these 6 styles.
-%%
-%% There are other optional arguments one can invoke to allow other stylistic
-%% actions. The available options are:
-%%
-%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
-%%   tighten      : Makes baselineskip slightly smaller, only works with 
-%%                  the twocolumn substyle.
-%%   times        : uses times font instead of the default
-%%   linenumbers  : turn on lineno package.
-%%   trackchanges : required to see the revision mark up and print its output
-%%   longauthor   : Do not use the more compressed footnote style (default) for 
-%%                  the author/collaboration/affiliations. Instead print all
-%%                  affiliation information after each name. Creates a much 
-%%                  longer author list but may be desirable for short 
-%%                  author papers.
-%% twocolappendix : make 2 column appendix.
-%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
-%%                  for dual anonymous review.
-%%
-%% these can be used in any combination, e.g.
-%%
-%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
-%%
-%% AASTeX v6.* now includes \hyperref support. While we have built in specific
-%% defaults into the classfile you can manually override them with the
-%% \hypersetup command. For example,
-%%
-%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
-%%
-%% will change the color of the internal links to red, the links to the
-%% bibliography to green, the file links to cyan, and the external links to
-%% magenta. Additional information on \hyperref options can be found here:
-%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
-%%
-%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
-%% to improve the accessibility of the compiled pdf file.
-%%
-%% If you want to create your own macros, you can do so
-%% using \newcommand. Your macros should appear before
-%% the \begin{document} command.
-%%
-
-\usepackage{xspace}
-
-\newcommand{\vdag}{(v)^\dagger}
-\newcommand\aastex{AAS\TeX}
-\newcommand\latex{La\TeX}
-\newcommand{\magcom}[1]{\textcolor{magenta}{#1}} % {\textbf{#1}}}
-\newcommand{\redcom}[1]{\textcolor{red}{#1}} % {\textbf{#1}}}
-\newcommand{\yelcom}[1]{\colorbox{yellow}{#1}} % {\textbf{#1}}}
-\newcommand{\pd}{\texttt{photoD}\xspace}
-\newcommand{\mh}{[M/H]\xspace}
-\newcommand{\Mb}{\textit{M\textsubscript{b}}\xspace}
-\newcommand{\logg}{\textit{log(g)}\xspace}
-\newcommand{\teff}{\textit{T\textsubscript{eff}}\xspace}
-\newcommand{\RV}{\textit{R\textsubscript{V}}\xspace}
-
-%% Reintroduced the \received and \accepted commands from AASTeX v5.2
-%\received{March 1, 2021}
-%\revised{April 1, 2021}
-%\accepted{\today}
-
-%% Command to document which AAS Journal the manuscript was submitted to.
-%% Adds "Submitted to " the argument.
-%\submitjournal{PSJ}
-
-%% The following command can be used to set the latex table counters.  It
-%% is needed in this document because it uses a mix of latex tabular and
-%% AASTeX deluxetables.  In general it should not be needed.
-%\setcounter{table}{1}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%
-%% The following section outlines numerous optional output that
-%% can be displayed in the front matter or as running meta-data.
-%%
-%% If you wish, you may supply running head information, although
-%% this information may be modified by the editorial offices.
-\shorttitle{PhotoD}
-\shortauthors{Schwarz et al.}
-%%
-%% You can add a light gray and diagonal water-mark to the first page 
-%% with this command:
-%% \watermark{text}
-%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
-%% long you can control the water-mark size with:
-%% \setwatermarkfontsize{dimension}
-%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
-%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\graphicspath{{./}{figures/}}
-
-\begin{document}
-
-\title{PhotoD: LSST photometric distances out to 100 kpc.}
-
-%% LaTeX will automatically break titles if they run longer than
-%% one line. However, you may use \\ to force a line break if
-%% you desire. In v6.31 you can include a footnote in the title.
-
-%%
-%% The \author command is the same as before except it now takes an optional
-%% argument which is the 16 digit ORCID. The syntax is:
-%% \author[xxxx-xxxx-xxxx-xxxx]{Author Name}
-
-%% Use \affiliation for affiliation information. The old \affil is now aliased
-%% to \affiliation. AASTeX v6.31 will automatically index these in the header.
-%% When a duplicate is found its index will be the same as its previous entry.
-%%
-%% Note that \altaffilmark and \altaffiltext have been removed and thus 
-%% can not be used to document secondary affiliations. If they are used latex
-%% will issue a specific error message and quit. Please use multiple 
-%% \affiliation calls for to document more than one affiliation.
-%%
-%% The new \altaffiliation can be used to indicate some secondary information
-%% such as fellowships. This command produces a non-numeric footnote that is
-%% set away from the numeric \affiliation footnotes.  NOTE that if an
-%% \altaffiliation command is used it must come BEFORE the \affiliation call,
-%% right after the \author command, in order to place the footnotes in
-%% the proper location.
-%%
-%% Use \email to set provide email addresses. Each \email will appear on its
-%% own line so you can put multiple email address in one \email call. A new
-%% \correspondingauthor command is available in V6.31 to identify the
-%% corresponding author of the manuscript. It is the author's responsibility
-%% to make sure this name is also in the author list.
-%%
-%% While authors can be grouped inside the same \author and \affiliation
-%% commands it is better to have a single author for each. This allows for
-%% one to exploit all the new benefits and should make book-keeping easier.
-%%
-%% If done correctly the peer review system will be able to
-%% automatically put the author and affiliation information from the manuscript
-%% and save the corresponding author the trouble of entering it by hand.
-
-\correspondingauthor{August Muench}
-\email{greg.schwarz@aas.org, gus.muench@aas.org}
-
-\author[0000-0002-0786-7307]{Greg J. Schwarz}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\author{August Muench}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\collaboration{6}{(AAS Journals Data Editors)}
-
-\author{Butler Burton}
-\affiliation{Leiden University}
-\affiliation{AAS Journals Associate Editor-in-Chief}
-
-\author{Amy Hendrickson}
-\altaffiliation{AASTeX v6+ programmer}
-\affiliation{TeXnology Inc.}
-
-\author{Julie Steffen}
-\affiliation{AAS Director of Publishing}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\author{Magaret Donnelly}
-\affiliation{IOP Publishing, Washington, DC 20005}
-
-%% Mark off the abstract in the ``abstract'' environment. 
-\begin{abstract}
-
-This example manuscript is intended to serve as a tutorial and template for
-authors to use when writing their own AAS Journal articles. The manuscript
-includes a history of \aastex\ and documents the new features in the
-previous versions as well as the bug fixes in version 6.31. This
-manuscript includes many figure and table examples to illustrate these new
-features.  Information on features not explicitly mentioned in the article
-can be viewed in the manuscript comments or more extensive online
-documentation. Authors are welcome replace the text, tables, figures, and
-bibliography with their own and submit the resulting manuscript to the AAS
-Journals peer review system.  The first lesson in the tutorial is to remind
-authors that the AAS Journals, the Astrophysical Journal (ApJ), the
-Astrophysical Journal Letters (ApJL), the Astronomical Journal (AJ), and
-the Planetary Science Journal (PSJ) all have a 250 word limit for the 
-abstract\footnote{Abstracts for Research Notes of the American Astronomical 
-Society (RNAAS) are limited to 150 words}.  If you exceed this length the
-Editorial office will ask you to shorten it. This abstract has 182 words.
-
-\end{abstract}
-
-%% Keywords should appear after the \end{abstract} command. 
-%% The AAS Journals now uses Unified Astronomy Thesaurus concepts:
-%% https://astrothesaurus.org
-%% You will be asked to selected these concepts during the submission process
-%% but this old "keyword" functionality is maintained in case authors want
-%% to include these concepts in their preprints.
-\keywords{Distance measure (395) --- Interstellar extinction (841) --- Photometry (1234) --- Stellar distance (1595) --- Two-color diagrams (1724)}
-
-%% From the front matter, we move on to the body of the paper.
-%% Sections are demarcated by \section and \subsection, respectively.
-%% Observe the use of the LaTeX \label
-%% command after the \subsection to give a symbolic KEY to the
-%% subsection for cross-referencing in a \ref command.
-%% You can use LaTeX's \ref and \label commands to keep track of
-%% cross-references to sections, equations, tables, and figures.
-%% That way, if you change the order of any elements, LaTeX will
-%% automatically renumber them.
-%%
-%% We recommend that authors also use the natbib \citep
-%% and \citet commands to identify citations.  The citations are
-%% tied to the reference list via symbolic KEYs. The KEY corresponds
-%% to the KEY in the \bibitem in the reference list below.
-
-\section{Introduction} \label{sec:intro}
-
-Thanks to the \textit{Vera C.
-Rubin} observatory's \textit{Legacy Survey of Space and Time} (LSST), for the first time in history, an astronomical catalog will contain more Milky Way stars than there are living people on Earth -- of the order 10-20 billion, depending on model assumptions.
-In order to map the Milky Way in three dimensions, distances to these stars must be accurately  estimated.
-In this paper we describe a method that will deliver LSST-based stellar distance estimations complementary to \textit{Gaia's} state-of-the-art trigonometric parallaxes and reach about 10 times further, to approximately 100 kpc.
-These results will be transformative for the studies of the Milky Way in general, and the stellar and the dark matter halo in particular as never before was there a survey that simultaneously observed roughly two thirds of the sky, to the co-added depth of \textit{r}$\approx$26 mag.
-
-\magcom{A bit about the importance of the distance estimation in the MW, dust implications (for extragalactic science too).}
-
-There are a variety of astronomical methods to estimate distances to stars, ranging from direct geometric (trigonometric) methods for nearby stars to indirect methods based on astrophysics for more distant stars.
-
-\magcom{Mention \cite{bailer-jones_estimating_2021}, \cite{gordon_panchromatic_2016}, \cite{green_measuring_2014,green_three-dimensional_2015,green_3d_2019}, \cite{juric_milky_2008} and \cite{lallement_3d_2014}, \cite{queiroz_starhorse_2018}}.
-
-Layout of the paper is...
-
-\section{Method} \label{sec:method}
-
-The photometric distance estimation method (hereafter \pd) is conceptually quite simple and relies on the strong correlations between the stellar colors and spectral energy distributions (SED) for dominant stellar populations.
-The stellar spectral energy distributions, and consequently colors, are determined by the effective temperature (\textit{T\textsubscript{eff}}), the surface gravity (\textit{g}), and the metallicity ([M/H]), or alternatively, by the absolute magnitude in band \textit{b} (\textit{M\textsubscript{b}}), [M/H] and age.
-
-The distributions that describe these correlations are obtained either from models or from observations.
-For example, the distribution of stellar SEDs in the color-color diagram in Figure \ref{fig:2c_example} provides key insights in stellar evolution and classification of different stellar populations such as main-sequence stars, giant stars, white dwarf stars, a majority of unresolved binary stars and even extragalactic objects.
-Analogous distributions are responsible for the abundant structure seen in the Hertzsprung-Russell diagram (HRD).
-
-\begin{figure}[ht!]
-	\plotone{GOOZvV.png}
-	\caption{Color-color diagram.
-\magcom{Add a version with overlaid evolutionary tracks in order to emphasize the required precision of the photometry required to disentangle the dwarfs and giants and different metallicities.
-\textbf{Are ev.
-tracks with different metallicities overlapping}?
-Can there be a situation where there is confusion between giants and dwarfs with different metallicities?} \label{fig:2c_example}}
-\end{figure}
-
-Metallicity is an important factor in these correlations, as it has a strong effect on the luminosity of the stars.
-This is reflected in the width of the main stellar loci of the color-magnitude diagrams (CMD) of the stellar populations observed at the same distance and the two-color diagrams (\magcom{quantify}), as seen in Figure \ref{fig:metallicity}.
-The best photometric estimators of metallicity are colors whose shorter-wavelength component includes the metal absorbtion bands at near-UV wavelengths, short of Balmer break (300 $\lessapprox\lambda$ [nm] $\lessapprox$ 400).
-Therefore, the LSST has a comparative advantage over the surveys lacking \textit{u}-band measurements, and could provide accurate distances within the range of 5-10\%.
-\magcom{A plot of model spectra, fixed, \textit{log(g)} and \textit{T\textsubscript{eff}}, several different metallicities?}
-
-\begin{figure}[ht!]
-	\plotone{Fdtez8.png}
-	\caption{Yadayada. \label{fig:metallicity}}
-\end{figure}
-
-Extinction is a major source of systematic errors in the process of luminosity and distance determination.
-The fact that the extinction vector is nearly parallel to the main stellar locus in the two-color diagrams gives rise to degeneracies that complicate the determination of the stellar type.
-An example is displayed in Fig.
-\ref{fig:degeneracies}, where in the left panel any of the different star types designated as 1,2 and 3 can have the same observed colors as the star marked as "Obs".
-This degeneracy is a result of the combination of colors chosen for the two-color diagram and depends on the adopted extinction curve \textit{R\textsubscript{V}} and the position on the stellar locus.
-The degeneracy can be broken if several different colors are used, particularly those towards the infrared, where the stellar locus is not kinked and the extinction vector is not parallel to it (as shown in the right panel of the Figure \ref{fig:degeneracies}).
-
-\magcom{Explain the choice of \RV.}
-
-\begin{figure}[ht!]
-	\plotone{BYiR7o.png}
-	\caption{Extinction \& degeneracies. \label{fig:degeneracies}}
-\end{figure}
-
-$L=4\pi\sigma R^2 T^4$\\
-$g = \frac{GM}{R^2}$\\
-$L=4\pi \sigma \frac{GM}{g} T^4$\\
-$\mu = m_b - M_b {\ }(- A_b)$\\
-$log(d) = 1 + \frac{\mu}{5}$\\
-
-Another important degeneracy arises from the fact that even for a fixed \teff and \mh, the \logg thus luminosity are not uniquely determined.
-A way to mitigate this issue is to adopt a Bayesian framework
-
-overlap in the color-color diagrams (e.g. Figure \ref{fig:2c_example}). This effectively means that for a fixed \teff proxy ($g-r$),
-
-Given the observed colors, and assuming small errors, the estimation of stellar parameters is relatively straight forward and follows one of the two possible approaches: empirical or model-based.
-In the former case, the case the underlying distribution to which the observations are matched is obtained from (ideally) spectroscopic and photometric observations, while in the latter case the distributions are generated from population synthesis models.
-
-\magcom{Advantages \& disadvantages of the model-based and empirical approaches, how model based approaches can be improved by adding empirical information for specific cases}.
-
-\magcom{\cite{gordon_panchromatic_2016} na \href{https://beast.readthedocs.io/en/latest/beast\_graphical\_model.html}{BEAST webu} imaju zgodne diagrame; možda bi i mi mogli nešto tog tipa napraviti, bar za draft, primjer }
-
-\begin{figure}[ht!]
-	\plottwo{beast-graphic-text.png}{beast-graphic-math.png}
-	\caption{BEAST. \label{fig:BEAST}}
-\end{figure}
-
-% -- modeli za boje moraju biti savršeni
-% -- novi prior iz trilegala
-% -- robustni framework koji može obraditi 10B zvijezda, i da nije spor
-% -- probali smo s MCMC, exhaustive locus search, i neuronske mreže koje su bile najbrže (drugi članak)
-
-\section{Testing}
-
-% -- aginst Rich et al. BDSD survey ugriz podaci
-% -- Bailer-Jones i S82
-% -- Trilegal simulacije s SDSS SED modelima (validacija)
-% -- kuglasti skupovi
-
-\section{Discussion}
-
 % -- popraviti modele
@@
 Therefore, given the observed colors it is necessary to simultaneously estimate the effective temperature, metallicity and extinction.
-
-\appendix
-
-\bibliography{PhotoD}{}
-\bibliographystyle{aasjournal}
-
-%% Include this line if you are using the \added, \replaced, \deleted
-%% commands to see a summary list of all changes at the end of the article.
-%\listofchanges
-
-\end{document}

### 2023-11-06 paper/intro.tex
individual files

--- 2023-11-02 paper/main.tex
@@
-\documentclass[linenumbers, twocolumn, trackchanges]{aastex631}
-
-%% The default is a single spaced, 10 point font, single spaced article.
-%% There are 5 other style options available via an optional argument. They
-%% can be invoked like this:
-%%
-%% \documentclass[arguments]{aastex631}
-%% 
-%% where the layout options are:
-%%
-%%  twocolumn   : two text columns, 10 point font, single spaced article.
-%%                This is the most compact and represent the final published
-%%                derived PDF copy of the accepted manuscript from the publisher
-%%  manuscript  : one text column, 12 point font, double spaced article.
-%%  preprint    : one text column, 12 point font, single spaced article.  
-%%  preprint2   : two text columns, 12 point font, single spaced article.
-%%  modern      : a stylish, single text column, 12 point font, article with
-%% 		  wider left and right margins. This uses the Daniel
-%% 		  Foreman-Mackey and David Hogg design.
-%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts 
-%%                but now that abstracts are required this is obsolete for
-%%                AAS Journals. Authors might need it for other reasons. DO NOT
-%%                use \begin{abstract} and \end{abstract} with this style.
-%%
-%% Note that you can submit to the AAS Journals in any of these 6 styles.
-%%
-%% There are other optional arguments one can invoke to allow other stylistic
-%% actions. The available options are:
-%%
-%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
-%%   tighten      : Makes baselineskip slightly smaller, only works with 
-%%                  the twocolumn substyle.
-%%   times        : uses times font instead of the default
-%%   linenumbers  : turn on lineno package.
-%%   trackchanges : required to see the revision mark up and print its output
-%%   longauthor   : Do not use the more compressed footnote style (default) for 
-%%                  the author/collaboration/affiliations. Instead print all
-%%                  affiliation information after each name. Creates a much 
-%%                  longer author list but may be desirable for short 
-%%                  author papers.
-%% twocolappendix : make 2 column appendix.
-%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
-%%                  for dual anonymous review.
-%%
-%% these can be used in any combination, e.g.
-%%
-%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
-%%
-%% AASTeX v6.* now includes \hyperref support. While we have built in specific
-%% defaults into the classfile you can manually override them with the
-%% \hypersetup command. For example,
-%%
-%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
-%%
-%% will change the color of the internal links to red, the links to the
-%% bibliography to green, the file links to cyan, and the external links to
-%% magenta. Additional information on \hyperref options can be found here:
-%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
-%%
-%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
-%% to improve the accessibility of the compiled pdf file.
-%%
-%% If you want to create your own macros, you can do so
-%% using \newcommand. Your macros should appear before
-%% the \begin{document} command.
-%%
-
-\usepackage{xspace}
-
-\newcommand{\vdag}{(v)^\dagger}
-\newcommand\aastex{AAS\TeX}
-\newcommand\latex{La\TeX}
-\newcommand{\magcom}[1]{\textcolor{magenta}{#1}} % {\textbf{#1}}}
-\newcommand{\redcom}[1]{\textcolor{red}{#1}} % {\textbf{#1}}}
-\newcommand{\yelcom}[1]{\colorbox{yellow}{#1}} % {\textbf{#1}}}
-\newcommand{\pd}{\texttt{photoD}\xspace}
-\newcommand{\mh}{[M/H]\xspace}
-\newcommand{\Mb}{\textit{M\textsubscript{b}}\xspace}
-\newcommand{\logg}{\textit{log(g)}\xspace}
-\newcommand{\teff}{\textit{T\textsubscript{eff}}\xspace}
-\newcommand{\RV}{\textit{R\textsubscript{V}}\xspace}
-
-%% Reintroduced the \received and \accepted commands from AASTeX v5.2
-%\received{March 1, 2021}
-%\revised{April 1, 2021}
-%\accepted{\today}
-
-%% Command to document which AAS Journal the manuscript was submitted to.
-%% Adds "Submitted to " the argument.
-%\submitjournal{PSJ}
-
-%% The following command can be used to set the latex table counters.  It
-%% is needed in this document because it uses a mix of latex tabular and
-%% AASTeX deluxetables.  In general it should not be needed.
-%\setcounter{table}{1}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%
-%% The following section outlines numerous optional output that
-%% can be displayed in the front matter or as running meta-data.
-%%
-%% If you wish, you may supply running head information, although
-%% this information may be modified by the editorial offices.
-\shorttitle{PhotoD}
-\shortauthors{Schwarz et al.}
-%%
-%% You can add a light gray and diagonal water-mark to the first page 
-%% with this command:
-%% \watermark{text}
-%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
-%% long you can control the water-mark size with:
-%% \setwatermarkfontsize{dimension}
-%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
-%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\graphicspath{{./}{figures/}}
-
-\begin{document}
-
-\title{PhotoD: LSST photometric distances out to 100 kpc.}
-
-%% LaTeX will automatically break titles if they run longer than
-%% one line. However, you may use \\ to force a line break if
-%% you desire. In v6.31 you can include a footnote in the title.
-
-%%
-%% The \author command is the same as before except it now takes an optional
-%% argument which is the 16 digit ORCID. The syntax is:
-%% \author[xxxx-xxxx-xxxx-xxxx]{Author Name}
-
-%% Use \affiliation for affiliation information. The old \affil is now aliased
-%% to \affiliation. AASTeX v6.31 will automatically index these in the header.
-%% When a duplicate is found its index will be the same as its previous entry.
-%%
-%% Note that \altaffilmark and \altaffiltext have been removed and thus 
-%% can not be used to document secondary affiliations. If they are used latex
-%% will issue a specific error message and quit. Please use multiple 
-%% \affiliation calls for to document more than one affiliation.
-%%
-%% The new \altaffiliation can be used to indicate some secondary information
-%% such as fellowships. This command produces a non-numeric footnote that is
-%% set away from the numeric \affiliation footnotes.  NOTE that if an
-%% \altaffiliation command is used it must come BEFORE the \affiliation call,
-%% right after the \author command, in order to place the footnotes in
-%% the proper location.
-%%
-%% Use \email to set provide email addresses. Each \email will appear on its
-%% own line so you can put multiple email address in one \email call. A new
-%% \correspondingauthor command is available in V6.31 to identify the
-%% corresponding author of the manuscript. It is the author's responsibility
-%% to make sure this name is also in the author list.
-%%
-%% While authors can be grouped inside the same \author and \affiliation
-%% commands it is better to have a single author for each. This allows for
-%% one to exploit all the new benefits and should make book-keeping easier.
-%%
-%% If done correctly the peer review system will be able to
-%% automatically put the author and affiliation information from the manuscript
-%% and save the corresponding author the trouble of entering it by hand.
-
-\correspondingauthor{August Muench}
-\email{greg.schwarz@aas.org, gus.muench@aas.org}
-
-\author[0000-0002-0786-7307]{Greg J. Schwarz}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\author{August Muench}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\collaboration{6}{(AAS Journals Data Editors)}
-
-\author{Butler Burton}
-\affiliation{Leiden University}
-\affiliation{AAS Journals Associate Editor-in-Chief}
-
-\author{Amy Hendrickson}
-\altaffiliation{AASTeX v6+ programmer}
-\affiliation{TeXnology Inc.}
-
-\author{Julie Steffen}
-\affiliation{AAS Director of Publishing}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\author{Magaret Donnelly}
-\affiliation{IOP Publishing, Washington, DC 20005}
-
-%% Mark off the abstract in the ``abstract'' environment. 
-\begin{abstract}
-
-This example manuscript is intended to serve as a tutorial and template for
-authors to use when writing their own AAS Journal articles. The manuscript
-includes a history of \aastex\ and documents the new features in the
-previous versions as well as the bug fixes in version 6.31. This
-manuscript includes many figure and table examples to illustrate these new
-features.  Information on features not explicitly mentioned in the article
-can be viewed in the manuscript comments or more extensive online
-documentation. Authors are welcome replace the text, tables, figures, and
-bibliography with their own and submit the resulting manuscript to the AAS
-Journals peer review system.  The first lesson in the tutorial is to remind
-authors that the AAS Journals, the Astrophysical Journal (ApJ), the
-Astrophysical Journal Letters (ApJL), the Astronomical Journal (AJ), and
-the Planetary Science Journal (PSJ) all have a 250 word limit for the 
-abstract\footnote{Abstracts for Research Notes of the American Astronomical 
-Society (RNAAS) are limited to 150 words}.  If you exceed this length the
-Editorial office will ask you to shorten it. This abstract has 182 words.
-
-\end{abstract}
-
-%% Keywords should appear after the \end{abstract} command. 
-%% The AAS Journals now uses Unified Astronomy Thesaurus concepts:
-%% https://astrothesaurus.org
-%% You will be asked to selected these concepts during the submission process
-%% but this old "keyword" functionality is maintained in case authors want
-%% to include these concepts in their preprints.
-\keywords{Distance measure (395) --- Interstellar extinction (841) --- Photometry (1234) --- Stellar distance (1595) --- Two-color diagrams (1724)}
-
-%% From the front matter, we move on to the body of the paper.
-%% Sections are demarcated by \section and \subsection, respectively.
-%% Observe the use of the LaTeX \label
-%% command after the \subsection to give a symbolic KEY to the
-%% subsection for cross-referencing in a \ref command.
-%% You can use LaTeX's \ref and \label commands to keep track of
-%% cross-references to sections, equations, tables, and figures.
-%% That way, if you change the order of any elements, LaTeX will
-%% automatically renumber them.
-%%
-%% We recommend that authors also use the natbib \citep
-%% and \citet commands to identify citations.  The citations are
-%% tied to the reference list via symbolic KEYs. The KEY corresponds
-%% to the KEY in the \bibitem in the reference list below.
-
 \section{Introduction} \label{sec:intro}
@@
 Layout of the paper is...
-
-\section{Method} \label{sec:method}
-
-The photometric distance estimation method (hereafter \pd) is conceptually quite simple and relies on the strong correlations between the stellar colors and spectral energy distributions (SED) for dominant stellar populations.
-The stellar spectral energy distributions, and consequently colors, are determined by the effective temperature (\textit{T\textsubscript{eff}}), the surface gravity (\textit{g}), and the metallicity ([M/H]), or alternatively, by the absolute magnitude in band \textit{b} (\textit{M\textsubscript{b}}), [M/H] and age.
-
-The distributions that describe these correlations are obtained either from models or from observations.
-For example, the distribution of stellar SEDs in the color-color diagram in Figure \ref{fig:2c_example} provides key insights in stellar evolution and classification of different stellar populations such as main-sequence stars, giant stars, white dwarf stars, a majority of unresolved binary stars and even extragalactic objects.
-Analogous distributions are responsible for the abundant structure seen in the Hertzsprung-Russell diagram (HRD).
-
-\begin{figure}[ht!]
-	\plotone{GOOZvV.png}
-	\caption{Color-color diagram.
-\magcom{Add a version with overlaid evolutionary tracks in order to emphasize the required precision of the photometry required to disentangle the dwarfs and giants and different metallicities.
-\textbf{Are ev.
-tracks with different metallicities overlapping}?
-Can there be a situation where there is confusion between giants and dwarfs with different metallicities?} \label{fig:2c_example}}
-\end{figure}
-
-Metallicity is an important factor in these correlations, as it has a strong effect on the luminosity of the stars.
-This is reflected in the width of the main stellar loci of the color-magnitude diagrams (CMD) of the stellar populations observed at the same distance and the two-color diagrams (\magcom{quantify}), as seen in Figure \ref{fig:metallicity}.
-The best photometric estimators of metallicity are colors whose shorter-wavelength component includes the metal absorbtion bands at near-UV wavelengths, short of Balmer break (300 $\lessapprox\lambda$ [nm] $\lessapprox$ 400).
-Therefore, the LSST has a comparative advantage over the surveys lacking \textit{u}-band measurements, and could provide accurate distances within the range of 5-10\%.
-\magcom{A plot of model spectra, fixed, \textit{log(g)} and \textit{T\textsubscript{eff}}, several different metallicities?}
-
-\begin{figure}[ht!]
-	\plotone{Fdtez8.png}
-	\caption{Yadayada. \label{fig:metallicity}}
-\end{figure}
-
-Extinction is a major source of systematic errors in the process of luminosity and distance determination.
-The fact that the extinction vector is nearly parallel to the main stellar locus in the two-color diagrams gives rise to degeneracies that complicate the determination of the stellar type.
-An example is displayed in Fig.
-\ref{fig:degeneracies}, where in the left panel any of the different star types designated as 1,2 and 3 can have the same observed colors as the star marked as "Obs".
-This degeneracy is a result of the combination of colors chosen for the two-color diagram and depends on the adopted extinction curve \textit{R\textsubscript{V}} and the position on the stellar locus.
-The degeneracy can be broken if several different colors are used, particularly those towards the infrared, where the stellar locus is not kinked and the extinction vector is not parallel to it (as shown in the right panel of the Figure \ref{fig:degeneracies}).
-
-\magcom{Explain the choice of \RV.}
-
-\begin{figure}[ht!]
-	\plotone{BYiR7o.png}
-	\caption{Extinction \& degeneracies. \label{fig:degeneracies}}
-\end{figure}
-
-$L=4\pi\sigma R^2 T^4$\\
-$g = \frac{GM}{R^2}$\\
-$L=4\pi \sigma \frac{GM}{g} T^4$\\
-$\mu = m_b - M_b {\ }(- A_b)$\\
-$log(d) = 1 + \frac{\mu}{5}$\\
-
-Another important degeneracy arises from the fact that even for a fixed \teff and \mh, the \logg thus luminosity are not uniquely determined.
-A way to mitigate this issue is to adopt a Bayesian framework
-
-overlap in the color-color diagrams (e.g. Figure \ref{fig:2c_example}). This effectively means that for a fixed \teff proxy ($g-r$),
-
-Given the observed colors, and assuming small errors, the estimation of stellar parameters is relatively straight forward and follows one of the two possible approaches: empirical or model-based.
-In the former case, the case the underlying distribution to which the observations are matched is obtained from (ideally) spectroscopic and photometric observations, while in the latter case the distributions are generated from population synthesis models.
-
-\magcom{Advantages \& disadvantages of the model-based and empirical approaches, how model based approaches can be improved by adding empirical information for specific cases}.
-
-\magcom{\cite{gordon_panchromatic_2016} na \href{https://beast.readthedocs.io/en/latest/beast\_graphical\_model.html}{BEAST webu} imaju zgodne diagrame; možda bi i mi mogli nešto tog tipa napraviti, bar za draft, primjer }
-
-\begin{figure}[ht!]
-	\plottwo{beast-graphic-text.png}{beast-graphic-math.png}
-	\caption{BEAST. \label{fig:BEAST}}
-\end{figure}
-
-% -- modeli za boje moraju biti savršeni
-% -- novi prior iz trilegala
-% -- robustni framework koji može obraditi 10B zvijezda, i da nije spor
-% -- probali smo s MCMC, exhaustive locus search, i neuronske mreže koje su bile najbrže (drugi članak)
-
-\section{Testing}
-
-% -- aginst Rich et al. BDSD survey ugriz podaci
-% -- Bailer-Jones i S82
-% -- Trilegal simulacije s SDSS SED modelima (validacija)
-% -- kuglasti skupovi
-
-\section{Discussion}
-
-% -- popraviti modele
-% -- dovući model s mladim zvijezdama
-% -- dodati modele za WD, BHB, unresolved binaries, RRL i druge populacije
-
-% Here we provide a layout of our method that simultaneously provides distance estimates as well as the three-dimensional estimation of the dust distribution.
-The distance modulus of a star can be estimated given its absolute and apparent magnitudes and the extinction along the line of sight.
-Therefore, given the observed colors it is necessary to simultaneously estimate the effective temperature, metallicity and extinction.
-
-\appendix
-
-\bibliography{PhotoD}{}
-\bibliographystyle{aasjournal}
-
-%% Include this line if you are using the \added, \replaced, \deleted
-%% commands to see a summary list of all changes at the end of the article.
-%\listofchanges
-
-\end{document}

### 2023-11-06 paper/methodology.tex
individual files

--- 2023-11-06 paper/main.tex
@@
-\documentclass[linenumbers, twocolumn, trackchanges]{aastex631}
-
-%% The default is a single spaced, 10 point font, single spaced article.
-%% There are 5 other style options available via an optional argument. They
-%% can be invoked like this:
-%%
-%% \documentclass[arguments]{aastex631}
-%% 
-%% where the layout options are:
-%%
-%%  twocolumn   : two text columns, 10 point font, single spaced article.
-%%                This is the most compact and represent the final published
-%%                derived PDF copy of the accepted manuscript from the publisher
-%%  manuscript  : one text column, 12 point font, double spaced article.
-%%  preprint    : one text column, 12 point font, single spaced article.  
-%%  preprint2   : two text columns, 12 point font, single spaced article.
-%%  modern      : a stylish, single text column, 12 point font, article with
-%% 		  wider left and right margins. This uses the Daniel
-%% 		  Foreman-Mackey and David Hogg design.
-%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts 
-%%                but now that abstracts are required this is obsolete for
-%%                AAS Journals. Authors might need it for other reasons. DO NOT
-%%                use \begin{abstract} and \end{abstract} with this style.
-%%
-%% Note that you can submit to the AAS Journals in any of these 6 styles.
-%%
-%% There are other optional arguments one can invoke to allow other stylistic
-%% actions. The available options are:
-%%
-%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
-%%   tighten      : Makes baselineskip slightly smaller, only works with 
-%%                  the twocolumn substyle.
-%%   times        : uses times font instead of the default
-%%   linenumbers  : turn on lineno package.
-%%   trackchanges : required to see the revision mark up and print its output
-%%   longauthor   : Do not use the more compressed footnote style (default) for 
-%%                  the author/collaboration/affiliations. Instead print all
-%%                  affiliation information after each name. Creates a much 
-%%                  longer author list but may be desirable for short 
-%%                  author papers.
-%% twocolappendix : make 2 column appendix.
-%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
-%%                  for dual anonymous review.
-%%
-%% these can be used in any combination, e.g.
-%%
-%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
-%%
-%% AASTeX v6.* now includes \hyperref support. While we have built in specific
-%% defaults into the classfile you can manually override them with the
-%% \hypersetup command. For example,
-%%
-%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
-%%
-%% will change the color of the internal links to red, the links to the
-%% bibliography to green, the file links to cyan, and the external links to
-%% magenta. Additional information on \hyperref options can be found here:
-%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
-%%
-%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
-%% to improve the accessibility of the compiled pdf file.
-%%
-%% If you want to create your own macros, you can do so
-%% using \newcommand. Your macros should appear before
-%% the \begin{document} command.
-%%
-
-\usepackage{xspace}
-\usepackage{amsmath}
-
-\newcommand{\vdag}{(v)^\dagger}
-\newcommand\aastex{AAS\TeX}
-\newcommand\latex{La\TeX}
-\newcommand{\magcom}[1]{\textcolor{magenta}{#1}} % {\textbf{#1}}}
-\newcommand{\redcom}[1]{\textcolor{red}{#1}} % {\textbf{#1}}}
-\newcommand{\yelcom}[1]{\colorbox{yellow}{#1}} % {\textbf{#1}}}
-\newcommand{\pd}{\texttt{photoD}\xspace}
-\newcommand{\mh}{[M/H]\xspace}
-\newcommand{\mhm}{[\mathrm{M} / \mathrm{H}]\xspace}
-\newcommand{\Mb}{\ensuremath{M{\mathrm{_b}}}\xspace}
-\newcommand{\Ar}{\ensuremath{A{\mathrm{_r}}}\xspace}
-
-% \newcommand{\Mb}{\textit{M\textsubscript{b}}\xspace}
-\newcommand{\logg}{\textit{log(g)}\xspace}
-\newcommand{\teff}{\textit{T\textsubscript{eff}}\xspace}
-\newcommand{\RV}{\textit{R\textsubscript{V}}\xspace}
-%\newcommand{\AV}{\textit{A\textsubscript{V}}\xspace}
-
-\newcommand{\ul}{\ensuremath{u}\xspace}
-\newcommand{\gl}{\ensuremath{g}\xspace}
-\newcommand{\rl}{\ensuremath{r}\xspace}
-\newcommand{\il}{\ensuremath{i}\xspace}
-\newcommand{\zl}{\ensuremath{z}\xspace}
-\newcommand{\yl}{\ensuremath{y}\xspace}
-
-\renewcommand{\arg}[1]{\! \left( #1 \right)}
-\newcommand{\Pcond}[2]{P \arg{ #1 \, | \, #2 }}
-% \newcommand{\yps}{\ensuremath{y_{\mathrm{P1}}}\xspace}
-%\newcommand{\EBV}{\ensuremath{E \left( B \! - \! V \right)}\xspace}
-%\newcommand{\Egr}{\ensuremath{E \left( g \! - \! r \right)}\xspace}
-%\newcommand{\EBPRP}{\ensuremath{E \left( BP \! - \! RP \right)}\xspace}
-
-%% Reintroduced the \received and \accepted commands from AASTeX v5.2
-%\received{March 1, 2021}
-%\revised{April 1, 2021}
-%\accepted{\today}
-
-%% Command to document which AAS Journal the manuscript was submitted to.
-%% Adds "Submitted to " the argument.
-%\submitjournal{PSJ}
-
-%% The following command can be used to set the latex table counters.  It
-%% is needed in this document because it uses a mix of latex tabular and
-%% AASTeX deluxetables.  In general it should not be needed.
-%\setcounter{table}{1}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%
-%% The following section outlines numerous optional output that
-%% can be displayed in the front matter or as running meta-data.
-%%
-%% If you wish, you may supply running head information, although
-%% this information may be modified by the editorial offices.
-\shorttitle{PhotoD}
-\shortauthors{Schwarz et al.}
-%%
-%% You can add a light gray and diagonal water-mark to the first page 
-%% with this command:
-%% \watermark{text}
-%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
-%% long you can control the water-mark size with:
-%% \setwatermarkfontsize{dimension}
-%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
-%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-\graphicspath{{./}{figures/}}
-
-\begin{document}
-
-\title{PhotoD: LSST photometric distances out to 100 kpc.}
-
-%% LaTeX will automatically break titles if they run longer than
-%% one line. However, you may use \\ to force a line break if
-%% you desire. In v6.31 you can include a footnote in the title.
-
-%%
-%% The \author command is the same as before except it now takes an optional
-%% argument which is the 16 digit ORCID. The syntax is:
-%% \author[xxxx-xxxx-xxxx-xxxx]{Author Name}
-
-%% Use \affiliation for affiliation information. The old \affil is now aliased
-%% to \affiliation. AASTeX v6.31 will automatically index these in the header.
-%% When a duplicate is found its index will be the same as its previous entry.
-%%
-%% Note that \altaffilmark and \altaffiltext have been removed and thus 
-%% can not be used to document secondary affiliations. If they are used latex
-%% will issue a specific error message and quit. Please use multiple 
-%% \affiliation calls for to document more than one affiliation.
-%%
-%% The new \altaffiliation can be used to indicate some secondary information
-%% such as fellowships. This command produces a non-numeric footnote that is
-%% set away from the numeric \affiliation footnotes.  NOTE that if an
-%% \altaffiliation command is used it must come BEFORE the \affiliation call,
-%% right after the \author command, in order to place the footnotes in
-%% the proper location.
-%%
-%% Use \email to set provide email addresses. Each \email will appear on its
-%% own line so you can put multiple email address in one \email call. A new
-%% \correspondingauthor command is available in V6.31 to identify the
-%% corresponding author of the manuscript. It is the author's responsibility
-%% to make sure this name is also in the author list.
-%%
-%% While authors can be grouped inside the same \author and \affiliation
-%% commands it is better to have a single author for each. This allows for
-%% one to exploit all the new benefits and should make book-keeping easier.
-%%
-%% If done correctly the peer review system will be able to
-%% automatically put the author and affiliation information from the manuscript
-%% and save the corresponding author the trouble of entering it by hand.
-
-\correspondingauthor{August Muench}
-\email{greg.schwarz@aas.org, gus.muench@aas.org}
-
-\author[0000-0002-0786-7307]{Greg J. Schwarz}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\author{August Muench}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\collaboration{6}{(AAS Journals Data Editors)}
-
-\author{Butler Burton}
-\affiliation{Leiden University}
-\affiliation{AAS Journals Associate Editor-in-Chief}
-
-\author{Amy Hendrickson}
-\altaffiliation{AASTeX v6+ programmer}
-\affiliation{TeXnology Inc.}
-
-\author{Julie Steffen}
-\affiliation{AAS Director of Publishing}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\author{Magaret Donnelly}
-\affiliation{IOP Publishing, Washington, DC 20005}
-
-%% Mark off the abstract in the ``abstract'' environment. 
-\begin{abstract}
-
-This example manuscript is intended to serve as a tutorial and template for
-authors to use when writing their own AAS Journal articles. The manuscript
-includes a history of \aastex\ and documents the new features in the
-previous versions as well as the bug fixes in version 6.31. This
-manuscript includes many figure and table examples to illustrate these new
-features.  Information on features not explicitly mentioned in the article
-can be viewed in the manuscript comments or more extensive online
-documentation. Authors are welcome replace the text, tables, figures, and
-bibliography with their own and submit the resulting manuscript to the AAS
-Journals peer review system.  The first lesson in the tutorial is to remind
-authors that the AAS Journals, the Astrophysical Journal (ApJ), the
-Astrophysical Journal Letters (ApJL), the Astronomical Journal (AJ), and
-the Planetary Science Journal (PSJ) all have a 250 word limit for the 
-abstract\footnote{Abstracts for Research Notes of the American Astronomical 
-Society (RNAAS) are limited to 150 words}.  If you exceed this length the
-Editorial office will ask you to shorten it. This abstract has 182 words.
-
-\end{abstract}
-
-%% Keywords should appear after the \end{abstract} command. 
-%% The AAS Journals now uses Unified Astronomy Thesaurus concepts:
-%% https://astrothesaurus.org
-%% You will be asked to selected these concepts during the submission process
-%% but this old "keyword" functionality is maintained in case authors want
-%% to include these concepts in their preprints.
-\keywords{Distance measure (395) --- Interstellar extinction (841) --- Photometry (1234) --- Stellar distance (1595) --- Two-color diagrams (1724)}
-
-%% From the front matter, we move on to the body of the paper.
-%% Sections are demarcated by \section and \subsection, respectively.
-%% Observe the use of the LaTeX \label
-%% command after the \subsection to give a symbolic KEY to the
-%% subsection for cross-referencing in a \ref command.
-%% You can use LaTeX's \ref and \label commands to keep track of
-%% cross-references to sections, equations, tables, and figures.
-%% That way, if you change the order of any elements, LaTeX will
-%% automatically renumber them.
-%%
-%% We recommend that authors also use the natbib \citep
-%% and \citet commands to identify citations.  The citations are
-%% tied to the reference list via symbolic KEYs. The KEY corresponds
-%% to the KEY in the \bibitem in the reference list below.
-
-\section{Introduction} \label{sec:intro}
-
-Thanks to the \textit{Vera C.
-Rubin} observatory's \textit{Legacy Survey of Space and Time} (LSST), for the first time in history, an astronomical catalog will contain more Milky Way stars than there are living people on Earth -- of the order 10-20 billion, depending on model assumptions.
-In order to map the Milky Way in three dimensions, distances to these stars must be accurately  estimated.
-In this paper we describe a method that will deliver LSST-based stellar distance estimations complementary to \textit{Gaia's} state-of-the-art trigonometric parallaxes and reach about 10 times further, to approximately 100 kpc.
-These results will be transformative for the studies of the Milky Way in general, and the stellar and the dark matter halo in particular as never before was there a survey that simultaneously observed roughly two thirds of the sky, to the co-added depth of \textit{r}$\approx$26 mag.
-
-\magcom{A bit about the importance of the distance estimation in the MW, dust implications (for extragalactic science too).}
-
-There are a variety of astronomical methods to estimate distances to stars, ranging from direct geometric (trigonometric) methods for nearby stars to indirect methods based on astrophysics for more distant stars.
-
-\magcom{Mention \cite{bailer-jones_estimating_2021}, \cite{gordon_panchromatic_2016}, \cite{green_measuring_2014,green_three-dimensional_2015,green_3d_2019}, \cite{juric_milky_2008} and \cite{lallement_3d_2014}, \cite{queiroz_starhorse_2018}}.
-
-Layout of the paper is...
-
 \section{Method} \label{sec:method}
@@
 
-\section{Testing}
+%%% METHODOLOGY
 
-% -- aginst Rich et al. BDSD survey ugriz podaci
-% -- Bailer-Jones i S82
-% -- Trilegal simulacije s SDSS SED modelima (validacija)
-% -- kuglasti skupovi
+% 1) astrophysics
 
-\section{Discussion}
+\begin{figure*}[ht!]
+\plotone{plot3diagsData.png}
+\caption{Copy caption from email} \label{fig:3dataDiags}
+\end{figure*}
 
-% -- popraviti modele
-% -- dovući model s mladim zvijezdama
-% -- dodati modele za WD, BHB, unresolved binaries, RRL i druge populacije
+\begin{figure*}[ht!]
+\plotone{plot3HRdiags.png}
+\caption{Copy caption from email} \label{fig:3HRdiags}
+\end{figure*}
 
-% Here we provide a layout of our method that simultaneously provides distance estimates as well as the three-dimensional estimation of the dust distribution.
-The distance modulus of a star can be estimated given its absolute and apparent magnitudes and the extinction along the line of sight.
-Therefore, given the observed colors it is necessary to simultaneously estimate the effective temperature, metallicity and extinction.
+% 2) statistics:
 
-\appendix
+\begin{figure*}[ht!]
+\plotone{bayesPanels_ex1.png}
+\caption{Write caption: prior, likelihood, posterior maps}
+\label{fig:bayesPanels}
+\end{figure*}
 
-\bibliography{PhotoD}{}
-\bibliographystyle{aasjournal}
+% 3D Bayes results for star i= 1
+% r mag: 21.19 g-r: 0.481 chi2min: 2.7487013195215177
+% Mr: true= 4.58 estimate= 4.49193855789433  +-  0.418936340852099
+% FeH: true= -0.76 estimate= -0.6860134814469806  +-  0.14586005018974993
+% Ar: true= 0.367 estimate= 0.3349289555609109  +-  0.07573177303312867
+% Qr: true= 4.947 estimate= 4.831891055998594  +-  0.38018147434019617
 
-%% Include this line if you are using the \added, \replaced, \deleted
-%% commands to see a summary list of all changes at the end of the article.
-%\listofchanges
+\begin{figure*}[ht!]
+\plotone{cornerPlot3_ex1.png}
+\caption{Write caption: 2-param covariances and marginal distributions }
+\label{fig:cornerPlot3}
+\end{figure*}
 
-\end{document}
+\begin{figure*}[ht!]
+\plotone{margPosteriors3D_ex1.png}
+\caption{Write caption: prior, likelihood, posterior marginal distributions}
+\label{fig:margPosteriors3D}
+\end{figure*}
+
+% 3) code testing:
+
+\begin{figure*}[ht!]
+\plotone{qpBmeans_chiTest4_3D.png}
+\caption{Write caption: TRILEGAL mean values of input model params in umag vs. g-i }
+\label{fig:qpBmeans_chiTest4_3D.}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{qpBcmd_chiTest4_3D_Mr.png}
+\caption{Write caption: performance for Mr vs. true Mr and FeH}
+\label{fig:qpBcmd_chiTest4_3D_Mr}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{qpBcmd_chiTest4_3D_FeH.png}
+\caption{Write caption: performance for FeH vs. true Mr and FeH}
+\label{fig:qpBcmd_chiTest4_3D_FeH}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{qpBcmd_chiTest4_3D_Ar.png}
+\caption{Write caption: performance for Ar vs. true Mr and FeH}
+\label{fig:qpBcmd_chiTest4_3D_Ar}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{qpB_chiTest4_3D_Mr.png}
+\caption{Write caption: performance for Mr vs. estimated Mr and FeH}
+\label{fig:qpB_chiTest4_3D_Mr}
+\end{figure*}
+
+% *** data-based comparisons ***
+% Stripe 82 vs. Gaia
+% Bulge BDSD sample?

### 2023-11-06 paper/notebooks/paperPlots.py
first import

@@ -1,0 +1,132 @@
+
+import numpy as np
+import pylab as plt
+import scipy.stats as stats
+from scipy.stats import norm
+from astroML.stats import binned_statistic_2d
+
+
+def plot3diagsData(df1, df2, df3, df4, L0, L1, L2):
+
+    fig=plt.figure(2,figsize=(12,4))
+    fig.subplots_adjust(wspace=0.2,hspace=0.2,top=0.97,bottom=0.1,left=0.09,right=0.95)
+    plt.rcParams['font.size'] = 12
+
+    def plotPanel(ax, df1, df2, df3, df4, Xname, Yname, xLabel, yLabel, xMin, xMax, yMin, yMax):
+        ax.scatter(df1[Xname], df1[Yname],s=0.01, c='blue')
+        ax.scatter(df2[Xname], df2[Yname],s=0.01, c='red')
+        ax.scatter(df3[Xname], df3[Yname],s=0.01, c='cyan')
+        ax.scatter(df4[Xname], df4[Yname],s=0.01, c='green')
+        ax.set_xlabel(xLabel)
+        ax.set_ylabel(yLabel)
+        ax.set_xlim(xMin, xMax)
+        ax.set_ylim(yMin, yMax)
+        # plt.title(title, x=0.30, y=1.008)
+        
+    cLocus = 'yellow'
+  
+    ax=fig.add_subplot(131)
+    plotPanel(ax, df1, df2, df3, df4, 'gi', 'rmag', 'g-i', 'SDSS r mag', -1.0, 3.9, 22.0, 13.0)
+
+    ax=fig.add_subplot(132)
+    plotPanel(ax, df1, df2, df3, df4, 'ug', 'gr', 'u-g', 'g-r', -1.0, 4.0, -0.7, 1.9)
+    ax.plot(L0['ug'], L0['gr'], ls='--', c=cLocus)
+    ax.plot(L1['ug'], L1['gr'], ls='-', c=cLocus)
+    ax.plot(L2['ug'], L2['gr'], ls='-.', c=cLocus)
+
+
+    ax=fig.add_subplot(133)
+    plotPanel(ax, df1, df2, df3, df4, 'gr', 'ri', 'g-r', 'r-i', -0.7, 1.9, -0.5, 2.3)
+    ax.plot(L0['gr'], L0['ri'], ls='--', c=cLocus)
+    ax.plot(L1['gr'], L1['ri'], ls='-', c=cLocus)
[…]

### 2023-11-06 paper/main.tex
new structure

--- 2023-11-02 paper/main.tex
@@
 \shorttitle{PhotoD}
-\shortauthors{Schwarz et al.}
+\shortauthors{Palaversa et al.}
 %%
@@
 
-\correspondingauthor{August Muench}
-\email{greg.schwarz@aas.org, gus.muench@aas.org}
+\correspondingauthor{Lovro Palaversa}
+\email{lp@irb.hr}
 
-\author[0000-0002-0786-7307]{Greg J. Schwarz}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
+\author[0000-0012-3245-123X]{LokiZg}
+\affiliation{IRB}
 
-\author{August Muench}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
+\author[0000-0012-3245-123X]{\v{Z}eljko Ivezi\'{c}}
+\affiliation{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195, USA}
 
-\collaboration{6}{(AAS Journals Data Editors)}
-
-\author{Butler Burton}
-\affiliation{Leiden University}
-\affiliation{AAS Journals Associate Editor-in-Chief}
-
-\author{Amy Hendrickson}
-\altaffiliation{AASTeX v6+ programmer}
-\affiliation{TeXnology Inc.}
-
-\author{Julie Steffen}
-\affiliation{AAS Director of Publishing}
-\affiliation{American Astronomical Society \\
-1667 K Street NW, Suite 800 \\
-Washington, DC 20006, USA}
-
-\author{Magaret Donnelly}
-\affiliation{IOP Publishing, Washington, DC 20005}
+\author[0000-0012-3245-123X]{Karlo Mrakov\v{c}i\'{c}}
+\affiliation{Faculty of Physics, University of Rijeka, Radmile Matej\v{c}i\'{c} 2, 51000 Rijeka, Croatia}
 
@@
 
-\section{Introduction} \label{sec:intro}
+\include{intro}
 
-Thanks to the \textit{Vera C.
-Rubin} observatory's \textit{Legacy Survey of Space and Time} (LSST), for the first time in history, an astronomical catalog will contain more Milky Way stars than there are living people on Earth -- of the order 10-20 billion, depending on model assumptions.
-In order to map the Milky Way in three dimensions, distances to these stars must be accurately  estimated.
-In this paper we describe a method that will deliver LSST-based stellar distance estimations complementary to \textit{Gaia's} state-of-the-art trigonometric parallaxes and reach about 10 times further, to approximately 100 kpc.
-These results will be transformative for the studies of the Milky Way in general, and the stellar and the dark matter halo in particular as never before was there a survey that simultaneously observed roughly two thirds of the sky, to the co-added depth of \textit{r}$\approx$26 mag.
+\include{methodology}
 
-\magcom{A bit about the importance of the distance estimation in the MW, dust implications (for extragalactic science too).}
+%\include{analysis}
 
-There are a variety of astronomical methods to estimate distances to stars, ranging from direct geometric (trigonometric) methods for nearby stars to indirect methods based on astrophysics for more distant stars.
-
-\magcom{Mention \cite{bailer-jones_estimating_2021}, \cite{gordon_panchromatic_2016}, \cite{green_measuring_2014,green_three-dimensional_2015,green_3d_2019}, \cite{juric_milky_2008} and \cite{lallement_3d_2014}, \cite{queiroz_starhorse_2018}}.
-
-Layout of the paper is...
-
-\section{Method} \label{sec:method}
-
-The photometric distance estimation method (hereafter \pd) is conceptually quite simple and relies on the strong correlations between the stellar colors and spectral energy distributions (SED) for dominant stellar populations.
-The stellar spectral energy distributions, and consequently colors, are determined by the effective temperature (\textit{T\textsubscript{eff}}), the surface gravity (\textit{g}), and the metallicity ([M/H]), or alternatively, by the absolute magnitude in band \textit{b} (\textit{M\textsubscript{b}}), [M/H] and age.
-
-The distributions that describe these correlations are obtained either from models or from observations.
-For example, the distribution of stellar SEDs in the color-color diagram in Figure \ref{fig:2c_example} provides key insights in stellar evolution and classification of different stellar populations such as main-sequence stars, giant stars, white dwarf stars, a majority of unresolved binary stars and even extragalactic objects.
-Analogous distributions are responsible for the abundant structure seen in the Hertzsprung-Russell diagram (HRD).
-
-\begin{figure}[ht!]
-	\plotone{GOOZvV.png}
-	\caption{Color-color diagram.
-\magcom{Add a version with overlaid evolutionary tracks in order to emphasize the required precision of the photometry required to disentangle the dwarfs and giants and different metallicities.
-\textbf{Are ev.
-tracks with different metallicities overlapping}?
-Can there be a situation where there is confusion between giants and dwarfs with different metallicities?} \label{fig:2c_example}}
-\end{figure}
-
-Metallicity is an important factor in these correlations, as it has a strong effect on the luminosity of the stars.
-This is reflected in the width of the main stellar loci of the color-magnitude diagrams (CMD) of the stellar populations observed at the same distance and the two-color diagrams (\magcom{quantify}), as seen in Figure \ref{fig:metallicity}.
-The best photometric estimators of metallicity are colors whose shorter-wavelength component includes the metal absorbtion bands at near-UV wavelengths, short of Balmer break (300 $\lessapprox\lambda$ [nm] $\lessapprox$ 400).
-Therefore, the LSST has a comparative advantage over the surveys lacking \textit{u}-band measurements, and could provide accurate distances within the range of 5-10\%.
-\magcom{A plot of model spectra, fixed, \textit{log(g)} and \textit{T\textsubscript{eff}}, several different metallicities?}
-
-\begin{figure}[ht!]
-	\plotone{Fdtez8.png}
-	\caption{Yadayada. \label{fig:metallicity}}
-\end{figure}
-
-Extinction is a major source of systematic errors in the process of luminosity and distance determination.
-The fact that the extinction vector is nearly parallel to the main stellar locus in the two-color diagrams gives rise to degeneracies that complicate the determination of the stellar type.
-An example is displayed in Fig.
-\ref{fig:degeneracies}, where in the left panel any of the different star types designated as 1,2 and 3 can have the same observed colors as the star marked as "Obs".
-This degeneracy is a result of the combination of colors chosen for the two-color diagram and depends on the adopted extinction curve \textit{R\textsubscript{V}} and the position on the stellar locus.
-The degeneracy can be broken if several different colors are used, particularly those towards the infrared, where the stellar locus is not kinked and the extinction vector is not parallel to it (as shown in the right panel of the Figure \ref{fig:degeneracies}).
-
-\magcom{Explain the choice of \RV.}
-
-\begin{figure}[ht!]
-	\plotone{BYiR7o.png}
-	\caption{Extinction \& degeneracies. \label{fig:degeneracies}}
-\end{figure}
-
-$L=4\pi\sigma R^2 T^4$\\
-$g = \frac{GM}{R^2}$\\
-$L=4\pi \sigma \frac{GM}{g} T^4$\\
-$\mu = m_b - M_b {\ }(- A_b)$\\
-$log(d) = 1 + \frac{\mu}{5}$\\
-
-Another important degeneracy arises from the fact that even for a fixed \teff and \mh, the \logg thus luminosity are not uniquely determined.
-A way to mitigate this issue is to adopt a Bayesian framework
-
-overlap in the color-color diagrams (e.g. Figure \ref{fig:2c_example}). This effectively means that for a fixed \teff proxy ($g-r$),
-
-Given the observed colors, and assuming small errors, the estimation of stellar parameters is relatively straight forward and follows one of the two possible approaches: empirical or model-based.
-In the former case, the case the underlying distribution to which the observations are matched is obtained from (ideally) spectroscopic and photometric observations, while in the latter case the distributions are generated from population synthesis models.
-
-\magcom{Advantages \& disadvantages of the model-based and empirical approaches, how model based approaches can be improved by adding empirical information for specific cases}.
-
-\magcom{\cite{gordon_panchromatic_2016} na \href{https://beast.readthedocs.io/en/latest/beast\_graphical\_model.html}{BEAST webu} imaju zgodne diagrame; možda bi i mi mogli nešto tog tipa napraviti, bar za draft, primjer }
-
-\begin{figure}[ht!]
-	\plottwo{beast-graphic-text.png}{beast-graphic-math.png}
-	\caption{BEAST. \label{fig:BEAST}}
-\end{figure}
-
-% -- modeli za boje moraju biti savršeni
-% -- novi prior iz trilegala
-% -- robustni framework koji može obraditi 10B zvijezda, i da nije spor
-% -- probali smo s MCMC, exhaustive locus search, i neuronske mreže koje su bile najbrže (drugi članak)
-
-\section{Testing}
-
-% -- aginst Rich et al. BDSD survey ugriz podaci
-% -- Bailer-Jones i S82
-% -- Trilegal simulacije s SDSS SED modelima (validacija)
-% -- kuglasti skupovi
-
-\section{Discussion}
-
-% -- popraviti modele
-% -- dovući model s mladim zvijezdama
-% -- dodati modele za WD, BHB, unresolved binaries, RRL i druge populacije
-
-% Here we provide a layout of our method that simultaneously provides distance estimates as well as the three-dimensional estimation of the dust distribution.
-The distance modulus of a star can be estimated given its absolute and apparent magnitudes and the extinction along the line of sight.
-Therefore, given the observed colors it is necessary to simultaneously estimate the effective temperature, metallicity and extinction.
+%\include{discussion}
 

### 2023-11-06 src/BayesTools.py
minor edits

@@ -28,6 +28,10 @@ def getBayesConstants():
     # where 1.3 <= ArCoef0, 0.1 <= ArCoef1 and 0.01 mag <= ArCoef2 are set here (or user supplied)
     BayesConst['ArCoeff0'] = 1.3
     BayesConst['ArCoeff1'] = 0.1
+    BayesConst['ArCoeff2'] = 0.01
+    
+    BayesConst['ArCoeff0'] = 1.4
+    BayesConst['ArCoeff1'] = 0.3
     BayesConst['ArCoeff2'] = 0.01 
     
     return BayesConst   
@@ -241,7 +245,7 @@ def getQmap(cube, FeH1d, Mr1d, Ar1d):
             for k in range(0,np.size(Ar1d)):
                 Mr = Qr1d[j] - Ar1d[k]
                 # now need to get the value of index for this Mr
-                jk = np.int((Mr-Mr1d[0])/(Mr1d[1]-Mr1d[0])) 
+                jk = int((Mr-Mr1d[0])/(Mr1d[1]-Mr1d[0])) 
                 if ((jk>=0)&(jk<np.size(Mr1d))):
                     Ssum += cube[i,jk,k]
             Qmap[i,j] = Ssum
@@ -702,7 +706,7 @@ def makeBayesEstimates3D(catalog, fitColors, locusData, locus3DList, ArGridList,
     catalog['ArEstUnc'] = 0*catalog['Mr'] -1 
     catalog['QrEst'] = 0*catalog['Mr'] - 99 
     catalog['QrEstUnc'] = 0*catalog['Mr'] -1 
-    catalog['chi2min'] = 0*catalog['Mr'] + 999
+    catalog['chi2min'] = 0*catalog['Mr'] - 99
     catalog['MrdS'] = 0*catalog['Mr'] - 1 
     catalog['FeHdS'] = 0*catalog['Mr'] - 1 
     catalog['ArdS'] = 0*catalog['Mr'] - 1 

### 2023-11-06 src/LocusTools.py

@@ -431,11 +431,12 @@ def readKarloMLestimates3D(inKfile):
         return simsML
   
         
-def LSSTsimsLocus(fixForStripe82=True): 
+def LSSTsimsLocus(fixForStripe82=True, datafile=""): 
         ## Mr, as function of [Fe/H], along the SDSS/LSST stellar 
         ## for more details see the file header
         colnames = ['Mr', 'FeH', 'ug', 'gr', 'ri', 'iz', 'zy']
-        LSSTlocus = Table.read('../data/LocusData/MSandRGBcolors_v1.3.txt', format='ascii', names=colnames)
+        if (datafile==""): datafile = '../data/LocusData/MSandRGBcolors_v1.3.txt'
+        LSSTlocus = Table.read(datafile, format='ascii', names=colnames)
         LSSTlocus['gi'] = LSSTlocus['gr'] + LSSTlocus['ri']
         if (fixForStripe82):
             print('Fixing input Mr-FeH-colors grid to agree with the SDSS v4.2 catalog')
@@ -701,7 +702,6 @@ def getPhotoDchi2map3D(i, colors, colorReddCoeffs, data2fit, locus, ArCoeff, mas
 
 
 def get3DmodelList(locusData, fitColors, agressive=False):
-
 
     if agressive:
         ## AGRESSIVE 

### 2023-11-06 src/PlotTools.py

@@ -37,7 +37,7 @@ def makeDMhistogram(x, xMin=1, xMax=22):
 
 
 def plot2Dmap(x, y, xMin, xMax, nXbin, yMin, yMax, nYbin, xLabel, yLabel, addMedians=True, logScale=False):
-
+   
     data = np.vstack([x, y])
     kde = gaussian_kde(data)
 
@@ -67,7 +67,13 @@ def plot2Dmap(x, y, xMin, xMax, nXbin, yMin, yMax, nYbin, xLabel, yLabel, addMed
         
     plt.xlabel(xLabel)
     plt.ylabel(yLabel)
+    plt.xlim(xMin, xMax)
+    plt.ylim(yMin, yMax)
 
+    if (0):
+         # plt.plot([3, 11.5], [3, 11.5], ls='--', lw=1, c='red')
+         plt.plot([2.5, 10.5], [0, 0], ls='--', lw=1, c='red')
+   
     if (addMedians):     
         xBin, nPts, medianBin, sigGbin = lt.fitMedians(x, y, xMin, xMax, 60)
         plt.scatter(xBin, medianBin, s=40, c='red')
@@ -76,6 +82,10 @@ def plot2Dmap(x, y, xMin, xMax, nXbin, yMin, yMax, nYbin, xLabel, yLabel, addMed
         plt.plot([xMin, xMax], [0,0], c='black')   
         return (xBin, medianBin)
     else:
+        # save
+        plt.savefig('../plots/2Dmap.png')
+        print('made ../plots/2Dmap.png')
+        plt.show() 
         return 
 
 def replot2Dmap(Xgrid, Ygrid, Z, xMin, xMax, nXbin, yMin, yMax, nYbin, xLabel, yLabel, logScale=False):

### 2023-11-08 paper/methodology.tex
Minor edits to the text.

@@
 
-\begin{figure}[ht!]
-	\plotone{GOOZvV.png}
-	\caption{Color-color diagram.
-\magcom{Add a version with overlaid evolutionary tracks in order to emphasize the required precision of the photometry required to disentangle the dwarfs and giants and different metallicities.
-\textbf{Are ev.
-tracks with different metallicities overlapping}?
-Can there be a situation where there is confusion between giants and dwarfs with different metallicities?} \label{fig:2c_example}}
-\end{figure}
+\begin{figure*}[ht!]
+	\plotone{plot3diagsData.png}
+	\caption{Copy caption from email} \label{fig:3dataDiags}
+\end{figure*}
 
@@
 \magcom{A plot of model spectra, fixed, \textit{log(g)} and \textit{T\textsubscript{eff}}, several different metallicities?}
-
-\begin{figure}[ht!]
-	\plotone{Fdtez8.png}
-	\caption{Yadayada. \label{fig:metallicity}}
-\end{figure}
 
@@
 
-\begin{figure}[ht!]
-	\plotone{BYiR7o.png}
-	\caption{Extinction \& degeneracies. \label{fig:degeneracies}}
-\end{figure}
-
-Another important degeneracy arises from the fact that even for a fixed \teff and \mh, the \logg and thus the luminosity are not uniquely determined by the colors: a degeneracy may exist between the giant branch and the main sequence as the colors constructed from \textit{ugrizy} bands are not sensitive to \logg (\magcom{As evident from ŽI's gr-ri plot.}).
+Another important degeneracy arises from the fact that even for a fixed \teff and \feh, the \logg and thus the luminosity are not uniquely determined by the colors: a degeneracy may exist between the giant branch and the main sequence as the colors constructed from \textit{ugrizy} bands are not sensitive to \logg (\magcom{As evident from ŽI's gr-ri plot.}).
 We treat this issue by adopting a prior based on bins in apparent magnitude.
 
-We adopt a Bayesian framework in which we simultaneously fit for \Mb, \mh and \Ar, assuming a fixed \RV value of 3.1\footnote{In principle \RV could be also fitted for.} The posterior for each individual star with LSST photometry is then given as:
+We adopt a Bayesian framework in which we simultaneously fit for \Mb, \feh and \Ar, assuming a fixed \RV value of 3.1\footnote{In principle \RV could be also fitted for.} The posterior for each individual star with LSST photometry is then given as:
 
@@
 	\begin{split}
-		    &\Pcond{\Mb, \, \mh, \, \Ar}{\vec{c}} = \\
-		& \frac{\Pcond{\vec{c}}{\Mb, \, \mh, \, \Ar} P\arg{\Mb, \, \mh, \, \Ar}}{P\arg{\vec{c}}} \,
+		    &\Pcond{\Mb, \, \feh, \, \Ar}{\vec{c}} = \\
+		& \frac{\Pcond{\vec{c}}{\Mb, \, \feh, \, \Ar} P\arg{\Mb, \, \feh, \, \Ar}}{P\arg{\vec{c}}} \,
 	\end{split}
@@
 \begin{equation}
-	ln \mathcal{L} = {-1 \over \sqrt{2\pi}} \,\sum_{i=1}^N \left({ c^{obs}_i - c^{mod}_i  \over \sigma_i } \right)^2 \,
+	ln (\mathcal{L}) = C - {1 \over 2} \,\sum_{i=1}^N \left({ c^{obs}_i - c^{mod}_i  \over \sigma_i } \right)^2 \,
 \end{equation}
@@
 
-% -- modeli za boje moraju biti savršeni
-% -- novi prior iz trilegala
-% -- robustni framework koji može obraditi 10B zvijezda, i da nije spor
-% -- probali smo s MCMC, exhaustive locus search, i neuronske mreže koje su bile najbrže (drugi članak)
-
-%$L=4\pi\sigma R^2 T^4=4\pi \sigma \frac{GM}{g} T^4$\\
-%$g = \frac{GM}{R^2}$\\
-%$\mu = m_b - M_b {\ }(+ A_b)$\\
-%$log(d) = 1 + \frac{\mu}{5}$\\
-
 %%% METHODOLOGY
@@
 % 1) astrophysics
-
-\begin{figure*}[ht!]
-\plotone{plot3diagsData.png}
-\caption{Copy caption from email} \label{fig:3dataDiags}
-\end{figure*}
 

### 2023-11-08 paper/main.tex
Minor edits to the text.
Fixed typos.

@@
 \usepackage{amsmath}
+\usepackage[T1]{fontenc}
 
@@
 \newcommand{\pd}{\texttt{photoD}\xspace}
-\newcommand{\mh}{[M/H]\xspace}
-\newcommand{\mhm}{[\mathrm{M} / \mathrm{H}]\xspace}
+\newcommand{\mh}{\ensuremath{[M/H]}\xspace}
+\newcommand{\feh}{\ensuremath{[Fe/H]}\xspace}
 \newcommand{\Mb}{\ensuremath{M{\mathrm{_b}}}\xspace}
+\newcommand{\Mr}{\ensuremath{M{\mathrm{_r}}}\xspace}
 \newcommand{\Ar}{\ensuremath{A{\mathrm{_r}}}\xspace}
-
-% \newcommand{\Mb}{\textit{M\textsubscript{b}}\xspace}
-\newcommand{\logg}{\textit{log(g)}\xspace}
-\newcommand{\teff}{\textit{T\textsubscript{eff}}\xspace}
-\newcommand{\RV}{\textit{R\textsubscript{V}}\xspace}
-%\newcommand{\AV}{\textit{A\textsubscript{V}}\xspace}
+\newcommand{\EBV}{\ensuremath{E \left( B \! - \! V \right)}\xspace}
+\newcommand{\logg}{\ensuremath{log(g)}\xspace}
+\newcommand{\teff}{\ensuremath{T\mathrm{_{eff}}}\xspace}
+\newcommand{\RV}{\ensuremath{R\mathrm{V}}\xspace}
+\newcommand{\AV}{\ensuremath{A{\mathrm{_V}}}\xspace}
 
@@
 % \newcommand{\yps}{\ensuremath{y_{\mathrm{P1}}}\xspace}
-%\newcommand{\EBV}{\ensuremath{E \left( B \! - \! V \right)}\xspace}
-%\newcommand{\Egr}{\ensuremath{E \left( g \! - \! r \right)}\xspace}
-%\newcommand{\EBPRP}{\ensuremath{E \left( BP \! - \! RP \right)}\xspace}
 
@@
 
-\author[0000-0012-3245-123X]{LokiZg}
-\affiliation{IRB}
+\author[0000-0003-3710-0331]{Lovro Palaversa}
+\affiliation{Ru{\dj}er Bo\v{s}kovi\'{c} Institute, Bijeni\v{c}ka cesta 54, 10000 Zagreb, Croatia }
 
@@
 
-This example manuscript is intended to serve as a tutorial and template for
-authors to use when writing their own AAS Journal articles. The manuscript
-includes a history of \aastex\ and documents the new features in the
-previous versions as well as the bug fixes in version 6.31. This
-manuscript includes many figure and table examples to illustrate these new
-features.  Information on features not explicitly mentioned in the article
-can be viewed in the manuscript comments or more extensive online
-documentation. Authors are welcome replace the text, tables, figures, and
-bibliography with their own and submit the resulting manuscript to the AAS
-Journals peer review system.  The first lesson in the tutorial is to remind
-authors that the AAS Journals, the Astrophysical Journal (ApJ), the
-Astrophysical Journal Letters (ApJL), the Astronomical Journal (AJ), and
-the Planetary Science Journal (PSJ) all have a 250 word limit for the 
-abstract\footnote{Abstracts for Research Notes of the American Astronomical 
-Society (RNAAS) are limited to 150 words}.  If you exceed this length the
-Editorial office will ask you to shorten it. This abstract has 182 words.
+Abstract here, mind the 250 word limit.
 

### 2023-11-08 paper/methodology.tex
Fixed typos.

@@
 The photometric distance estimation method (hereafter \pd) is conceptually quite simple and relies on the strong correlations between the stellar colors and spectral energy distributions (SED) for dominant stellar populations.
-The stellar spectral energy distributions, and consequently colors, are determined by the effective temperature (\textit{T\textsubscript{eff}}), the surface gravity (\textit{g}), and the metallicity ([M/H]), or alternatively, by the absolute magnitude in band \textit{b} (\textit{M\textsubscript{b}}), [M/H] and age.
+The SEDs, and consequently colors, are determined by the effective temperature (\teff), the surface gravity (usually denoted as \logg), and the metallicity (\mh), or alternatively, by the absolute magnitude in band \textit{b} (\Mb), \mh and age.
 
 The distributions that describe these correlations are obtained either from models or from observations.
-For example, the distribution of stellar SEDs in the color-color diagram in Figure \ref{fig:2c_example} provides key insights in stellar evolution and classification of different stellar populations such as main-sequence stars, giant stars, white dwarf stars, a majority of unresolved binary stars and even extragalactic objects.
+For example, the distribution of stellar SEDs in the color-color diagram in the middle and left panels of Figure~\ref{fig:3dataDiags} provide key insights in stellar evolution and classification of different stellar populations such as main-sequence stars, giant stars, white dwarf stars, a majority of unresolved binary stars and even extragalactic objects.
 Analogous distributions are responsible for the abundant structure seen in the Hertzsprung-Russell diagram (HRD).
@@
 	\plotone{plot3diagsData.png}
-	\caption{Copy caption from email} \label{fig:3dataDiags}
+	\caption{The blue dots in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
+A subset	of 415,000 stars with $r < 22$ and $u<22$ are shown as red dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ are shown as cyan dots.
+Finally, 63,000 stars that have signal-to-noise ratio for Gaia’s parallax measurements of at least 20 are shown as green dots.
+The same color scheme is used in other two panels.
+The three yellow lines in the middle panel show stellar locus parametrization used by Green et al.
+(2014) for three values of metallicity (left to right): $[Fe/H] = -2.0, -1.0, 0.0$.
+In the right panel, the impact of metallicity on color-color tracks is negligible and all three are indistinguishable from each other.
+} \label{fig:3dataDiags}
 \end{figure*}
@@
 Metallicity is an important factor in these correlations, as it has a strong effect on the luminosity of the stars.
-This is reflected in the width of the main stellar loci of the color-magnitude diagrams (CMD) of the stellar populations observed at the same distance and the two-color diagrams (\magcom{quantify}), as seen in Figure \ref{fig:metallicity}.
+This is reflected in the width of the main stellar loci of the two-color diagrams (middle and right hand panels in Figure~\ref{fig:3dataDiags} and the color-absolute magnitude diagrams (CAMD) in Figure~\ref{fig:3HRdiags}.
 The best photometric estimators of metallicity are colors whose shorter-wavelength component includes the metal absorbtion bands at near-UV wavelengths, short of Balmer break (300 $\lessapprox\lambda$ [nm] $\lessapprox$ 400).
@@
 
+\begin{figure*}[ht!]
+	\plotone{plot3HRdiags.png}
+	\caption{The left panel shows the absolute magnitude vs.
+color parametrization for main sequence and red giant stars.
+The symbols are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0.0 (blue to red).
+The three lines correspond to three values of metallicity: $[Fe/H] = -2.0, -1.0, 0.0$ (dot-dashed, solid and dashed, respectively).
+The middle panel shows a sample of 63,000 stars that have signal-to-noise ratio for Gaia’s parallax measurements of at least 20 (white dwarfs can be seen in the lower left corner).
+The dot-dashed, solid and dashed black lines are the same as in the left panel.
+For comparison, the dotted lines were computed using eqs.
+A2 and A7 from Ivezic et al.
+(2008).
+The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots, and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots.
+Their absolute magnitudes were computed using the so-called “photo-geometric” distances from Bailer-Jones et al.
+(2021).
+The dot-dashed, solid and dashed black lines are the same as in the left and middle panels.
+About 10,000 outliers seen at $g-i = 0.4$ and $Mr > 7$ are predominantly found at the faint end ($r>20$).}
+	\label{fig:3HRdiags}
+\end{figure*}
+
 Extinction is another major source of systematic errors in the process of luminosity and distance determination.
 The fact that the extinction vector is nearly parallel to the main stellar locus in the two-color diagrams gives rise to degeneracies that complicate the determination of the stellar type.
-An example is displayed in Fig.
-\ref{fig:degeneracies}, where in the left panel any of the different star types designated as 1,2 and 3 can have the same observed colors as the star marked as "Obs".
+An example is displayed in Figure~\ref{fig:extinction}, where in the left panel any of the different star types designated as 1,2 and 3 can have the same observed colors as the star marked as "Obs".
 This degeneracy is a result of the combination of colors chosen for the two-color diagram and depends on the position on the stellar locus and the adopted extinction curve parametrized by a single parameter \textit{R\textsubscript{V}}
+
+\begin{figure}[ht!]
+	\plotone{BYiR7o.png}
+	\caption{Add caption.} \label{fig:extinction}
+\end{figure}
 
@@
 with $A_r$ designating extinction in \textit{r}-band and $C_{\lambda}(R_V)$ describing the shape of the extinction curve.
-The degeneracy from the left panel in Fig.
-\ref{fig:degeneracies} can be broken if several different colors are used, particularly those towards the infrared, where the stellar locus is not kinked and the extinction vector is not parallel to it (as shown in the right panel of the Figure \ref{fig:degeneracies}), where $r-i$ and $i-z$ colors are used, and assuming a fixed extinction law a unique solution for the extinction is possible.
+The degeneracy from the left panel in Figure~\ref{fig:extinction} can be broken if several different colors are used, particularly those towards the infrared.
+There the stellar locus is not as kinked and the extinction vector is not parallel to it.
+A possible example is shown in the right-hand panel of the Fig.~\ref{fig:extinction}, where $r-i$ and $i-z$ colors are used, and assuming a fixed extinction law a unique solution for the extinction is possible.
 
@@
 
-Another important degeneracy arises from the fact that even for a fixed \teff and \feh, the \logg and thus the luminosity are not uniquely determined by the colors: a degeneracy may exist between the giant branch and the main sequence as the colors constructed from \textit{ugrizy} bands are not sensitive to \logg (\magcom{As evident from ŽI's gr-ri plot.}).
+Another important degeneracy arises from the fact that even for a fixed \teff and \feh, the \logg and thus the luminosity are not uniquely determined by the colors: a degeneracy may exist between the giant branch and the main sequence as the colors constructed from \textit{ugrizy} bands are not sensitive to \logg.
 We treat this issue by adopting a prior based on bins in apparent magnitude.
@@
 
-\begin{figure}[ht!]
-	\plottwo{beast-graphic-text.png}{beast-graphic-math.png}
-	\caption{BEAST. \label{fig:BEAST}}
-\end{figure}
-
 %%% METHODOLOGY
-
-% 1) astrophysics
-
-\begin{figure*}[ht!]
-\plotone{plot3HRdiags.png}
-\caption{Copy caption from email} \label{fig:3HRdiags}
-\end{figure*}
 

### 2023-11-08 paper/methodology.tex
Bayes & TRILEGAL.

@@
 Another important degeneracy arises from the fact that even for a fixed \teff and \feh, the \logg and thus the luminosity are not uniquely determined by the colors: a degeneracy may exist between the giant branch and the main sequence as the colors constructed from \textit{ugrizy} bands are not sensitive to \logg.
-We treat this issue by adopting a prior based on bins in apparent magnitude.
 
@@
 
-where \ensuremath{c^{obs}_i} are the observed colors and \ensuremath{c^{mod}_i} model colors.
-The values of model colors and the priors are extracted from from TRILEGAL (\citet{dal_tio_simulating_2022}).
-In order to extract the priors (i.e.
+where \ensuremath{c^{obs}_i} are the observed colors and \ensuremath{c^{mod}_i} the model colors.
+The priors are established by partitioning the TRILEGAL galaxy model \citep{dal_tio_simulating_2022} in healpixels, and each of the pixel in one-magnitude wide bins in apparent magnitude.
+The model colors are then obtained from models based on SDSS results (\magcom{reference}), where the input for the models are the \Mr and the \feh obtained from TRILEGAL.
+Given the assumed extinction curve, these colors are then reddened up to the maximum reddening which is estimated from \cite{schlegel_maps_1998} (SFD98) maps.
+The latter upper limit is used in order to provide a realistic stop condition on the amount of extinction and reduce the processing time.
+This is usually a valid assumption because the SFD98 maps provide \textit{total} extinction along a line of sight.
+In order to account for the eventual underestimation in the SFD98 maps we increase the SFD98 extinctions by 20\%.
+
+% We extract the priors based on position and apparent magnitude from TRILEGAL (\citet{dal_tio_simulating_2022}).
+Ideally, we would In order to extract the priors (i.e.
 prior maps), we divide TRILEGAL data in healpix bins, and further subdivide them in one-magnitude wide bins in apparent magnitude.
@@
 
-\magcom{Add plots describing the method, and go through one example like in Željko's slides. }
+%Our method is basically brute-force fitting with some intelligent tricks leveraged to obtain faster execution that will be required for 10B LSST stars.
 
-Our method is basically brute-force fitting with some intelligent tricks leveraged to obtain faster execution that will be required for 10B LSST stars.
-We use \cite{schlegel_maps_1998} (SFD98) maps in order to limit the range of the extinction.
-This is usually a valid assumption because the SFD98 maps provide \textit{total} extinction along a line of sight.
 Our fitting procedure is also executed on an adaptive grid, a coarse search over the parameter space is performed first in order to establish the layout of the manifold.
@@
 %Under the assumptions that the locus of the color-color diagram is representative of the stellar SEDs, that the shape of the extinction curve can be parametrized with a single parameter \RV
-
-\magcom{Advantages \& disadvantages of the model-based and empirical approaches, how model based approaches can be improved by adding empirical information for specific cases}.
-
-\magcom{\cite{gordon_panchromatic_2016} na \href{https://beast.readthedocs.io/en/latest/beast\_graphical\_model.html}{BEAST webu} imaju zgodne diagrame; možda bi i mi mogli nešto tog tipa napraviti, bar za draft, primjer }
 

### 2023-11-09 paper/methodology.tex
switched red and blue color to correspond to stellar colors in plot3diagsData.png

@@
 	\plotone{plot3diagsData.png}
-	\caption{The blue dots in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
-A subset	of 415,000 stars with $r < 22$ and $u<22$ are shown as red dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ are shown as cyan dots.
+	\caption{The red dots in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
+A subset of 415,000 stars with $r < 22$ and $u<22$ are overplotted as blue dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ are overplotted as cyan dots.
 Finally, 63,000 stars that have signal-to-noise ratio for Gaia’s parallax measurements of at least 20 are shown as green dots.

### 2023-11-09 paper/notebooks/paperPlots.py

@@ -13,10 +13,10 @@ def plot3diagsData(df1, df2, df3, df4, L0, L1, L2):
     plt.rcParams['font.size'] = 12
 
     def plotPanel(ax, df1, df2, df3, df4, Xname, Yname, xLabel, yLabel, xMin, xMax, yMin, yMax):
-        ax.scatter(df1[Xname], df1[Yname],s=0.01, c='blue')
-        ax.scatter(df2[Xname], df2[Yname],s=0.01, c='red')
-        ax.scatter(df3[Xname], df3[Yname],s=0.01, c='cyan')
-        ax.scatter(df4[Xname], df4[Yname],s=0.01, c='green')
+        ax.scatter(df1[Xname], df1[Yname],s=0.01, c='red')
+        ax.scatter(df2[Xname], df2[Yname],s=0.012, c='blue')
+        ax.scatter(df3[Xname], df3[Yname],s=0.014, c='cyan')
+        ax.scatter(df4[Xname], df4[Yname],s=0.016, c='green')
         ax.set_xlabel(xLabel)
         ax.set_ylabel(yLabel)
         ax.set_xlim(xMin, xMax)

### 2023-11-09 src/BayesTools.py

@@ -30,10 +30,6 @@ def getBayesConstants():
     BayesConst['ArCoeff1'] = 0.1
     BayesConst['ArCoeff2'] = 0.01
     
-    BayesConst['ArCoeff0'] = 1.4
-    BayesConst['ArCoeff1'] = 0.3
-    BayesConst['ArCoeff2'] = 0.01 
-    
     return BayesConst   
 
 

### 2024-01-08 paper/paper.tex
changed file names

--- 2023-11-08 paper/main.tex
@@
-\documentclass[linenumbers, twocolumn, trackchanges]{aastex631}
+\documentclass{aastex63} 
+% \documentclass[12pt,preprint]{aastex}
+% \documentclass[linenumbers, twocolumn, trackchanges]{aastex631}
 
-%% The default is a single spaced, 10 point font, single spaced article.
-%% There are 5 other style options available via an optional argument. They
-%% can be invoked like this:
-%%
-%% \documentclass[arguments]{aastex631}
-%% 
-%% where the layout options are:
-%%
-%%  twocolumn   : two text columns, 10 point font, single spaced article.
-%%                This is the most compact and represent the final published
-%%                derived PDF copy of the accepted manuscript from the publisher
-%%  manuscript  : one text column, 12 point font, double spaced article.
-%%  preprint    : one text column, 12 point font, single spaced article.  
-%%  preprint2   : two text columns, 12 point font, single spaced article.
-%%  modern      : a stylish, single text column, 12 point font, article with
-%% 		  wider left and right margins. This uses the Daniel
-%% 		  Foreman-Mackey and David Hogg design.
-%%  RNAAS       : Supresses an abstract. Originally for RNAAS manuscripts 
-%%                but now that abstracts are required this is obsolete for
-%%                AAS Journals. Authors might need it for other reasons. DO NOT
-%%                use \begin{abstract} and \end{abstract} with this style.
-%%
-%% Note that you can submit to the AAS Journals in any of these 6 styles.
-%%
-%% There are other optional arguments one can invoke to allow other stylistic
-%% actions. The available options are:
-%%
-%%   astrosymb    : Loads Astrosymb font and define \astrocommands. 
-%%   tighten      : Makes baselineskip slightly smaller, only works with 
-%%                  the twocolumn substyle.
-%%   times        : uses times font instead of the default
-%%   linenumbers  : turn on lineno package.
-%%   trackchanges : required to see the revision mark up and print its output
-%%   longauthor   : Do not use the more compressed footnote style (default) for 
-%%                  the author/collaboration/affiliations. Instead print all
-%%                  affiliation information after each name. Creates a much 
-%%                  longer author list but may be desirable for short 
-%%                  author papers.
-%% twocolappendix : make 2 column appendix.
-%%   anonymous    : Do not show the authors, affiliations and acknowledgments 
-%%                  for dual anonymous review.
-%%
-%% these can be used in any combination, e.g.
-%%
-%% \documentclass[twocolumn,linenumbers,trackchanges]{aastex631}
-%%
-%% AASTeX v6.* now includes \hyperref support. While we have built in specific
-%% defaults into the classfile you can manually override them with the
-%% \hypersetup command. For example,
-%%
-%% \hypersetup{linkcolor=red,citecolor=green,filecolor=cyan,urlcolor=magenta}
-%%
-%% will change the color of the internal links to red, the links to the
-%% bibliography to green, the file links to cyan, and the external links to
-%% magenta. Additional information on \hyperref options can be found here:
-%% https://www.tug.org/applications/hyperref/manual.html#x1-40003
-%%
-%% Note that in v6.3 "bookmarks" has been changed to "true" in hyperref
-%% to improve the accessibility of the compiled pdf file.
-%%
-%% If you want to create your own macros, you can do so
-%% using \newcommand. Your macros should appear before
-%% the \begin{document} command.
-%%
+\usepackage{amssymb,amsmath}
+% \usepackage{lsst}
+\usepackage{xspace}
+\usepackage[english]{babel}
+\usepackage[utf8x]{inputenc}
+\usepackage{hyperref}
+\usepackage{amsmath}
+\usepackage{graphicx}
+\usepackage{longtable}
+\usepackage{comment}
+\usepackage{booktabs}
+\addto\extrasenglish{%
+  \renewcommand{\equationautorefname}{eq.}%
+}
+
+%%%%%%%%%%%%%%%%%%%%%%
+\received{\today}
+\revised{XXX}
+\accepted{XXX}
+\submitjournal{AJ}
+
+% Croatian letter "dj"
+\def\d   {{d $\mkern-14.3mu \mathchar'26 $}}
 
@@
 %% this information may be modified by the editorial offices.
-\shorttitle{PhotoD}
+\shorttitle{PhotoD with LSST}
 \shortauthors{Palaversa et al.}
-%%
-%% You can add a light gray and diagonal water-mark to the first page 
-%% with this command:
-%% \watermark{text}
-%% where "text", e.g. DRAFT, is the text to appear.  If the text is 
-%% long you can control the water-mark size with:
-%% \setwatermarkfontsize{dimension}
-%% where dimension is any recognized LaTeX dimension, e.g. pt, in, etc.
-%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 \graphicspath{{./}{figures/}}
@@
 
-\title{PhotoD: LSST photometric distances out to 100 kpc.}
-
-%% LaTeX will automatically break titles if they run longer than
-%% one line. However, you may use \\ to force a line break if
-%% you desire. In v6.31 you can include a footnote in the title.
-
-%%
-%% The \author command is the same as before except it now takes an optional
-%% argument which is the 16 digit ORCID. The syntax is:
-%% \author[xxxx-xxxx-xxxx-xxxx]{Author Name}
-
-%% Use \affiliation for affiliation information. The old \affil is now aliased
-%% to \affiliation. AASTeX v6.31 will automatically index these in the header.
-%% When a duplicate is found its index will be the same as its previous entry.
-%%
-%% Note that \altaffilmark and \altaffiltext have been removed and thus 
-%% can not be used to document secondary affiliations. If they are used latex
-%% will issue a specific error message and quit. Please use multiple 
-%% \affiliation calls for to document more than one affiliation.
-%%
-%% The new \altaffiliation can be used to indicate some secondary information
-%% such as fellowships. This command produces a non-numeric footnote that is
-%% set away from the numeric \affiliation footnotes.  NOTE that if an
-%% \altaffiliation command is used it must come BEFORE the \affiliation call,
-%% right after the \author command, in order to place the footnotes in
-%% the proper location.
-%%
-%% Use \email to set provide email addresses. Each \email will appear on its
-%% own line so you can put multiple email address in one \email call. A new
-%% \correspondingauthor command is available in V6.31 to identify the
-%% corresponding author of the manuscript. It is the author's responsibility
-%% to make sure this name is also in the author list.
-%%
-%% While authors can be grouped inside the same \author and \affiliation
-%% commands it is better to have a single author for each. This allows for
-%% one to exploit all the new benefits and should make book-keeping easier.
-%%
-%% If done correctly the peer review system will be able to
-%% automatically put the author and affiliation information from the manuscript
-%% and save the corresponding author the trouble of entering it by hand.
-
-\correspondingauthor{Lovro Palaversa}
-\email{lp@irb.hr}
+\title{PhotoD with LSST: stellar photometric distances out to the edge of the Galaxy}
 
@@
 \author[0000-0012-3245-123X]{\v{Z}eljko Ivezi\'{c}}
-\affiliation{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195, USA}
+\affiliation{Department of Astronomy and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}
 
@@
 
+\author[0000-0000-0000-0000]{Bob Abel}
+\affiliation{The Paradise Island, WA, USA }
+
+\author[0000-0000-0000-0000]{Dani Chao}
+\affiliation{Ru{\dj}er Bo\v{s}kovi\'{c} Institute, Bijeni\v{c}ka cesta 54, 10000 Zagreb, Croatia }
+
+\author[0000-0002-3045-0446]{Oleksandra Razim}
+\affiliation{Ru{\dj}er Bo\v{s}kovi\'{c} Institute, Bijeni\v{c}ka cesta 54, 10000 Zagreb, Croatia }
+
+\author[0000-0000-0000-0000]{Filip Matkovi\'{c}}
+\affiliation{Hvar Observatory, 10000 Zagreb, Croatia}
+
+\author[0000-0000-0000-0000]{Connor Yablonski}
+\affiliation{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195, USA}
+
+\author[0000-0000-0000-0000]{Toni \v{S}ari\'{c}}
+\affiliation{University of Split, Croatia}
+
+\author[0000-0000-0000-0000]{Tomislav Jurki\'{c}}
+\affiliation{University of Rijeka, Croatia}
+
+\author[0000-0000-0000-0000]{Mario Juri\'{c}}
+\affiliation{Department of Astronomy and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}
+
+\author[0000-0000-0000-0000]{Forgotten Collaborator}
+\affiliation{Somewhere}
+
+%\correspondingauthor{Lovro Palaversa}
+%\email{lp@irb.hr}
+
 %% Mark off the abstract in the ``abstract'' environment. 
 \begin{abstract}
-
-Abstract here, mind the 250 word limit.
-
+As demonstrated with SDSS, Pan-STARRS and most recently with Gaia data, broad-band near-UV to near-IR photometry can be used 
+to estimate distances, metallicity and the interstellar dust extinction along the line of sight for stars in the Galaxy. Anticipating
+photometric catalogs with tens of billions of stars from Rubin's Legacy Survey of Space and Time (LSST), we present a fully Bayesian procedure 
+and pipeline that build on previous work and can handle LSST-sized datasets. Likelihood computations utilize MIST/Dartmouth isochrones
+and priors are derived from TRILEGAL-based simulated LSST catalogs from Dal Tio et al. (2022). The computation speed is about 10 mas per
+star for both optimized grid search and MCMC methods; we show in a companion paper by Mrakov\v{c}i\'{c} et al. how to utilize neural
+networks to accelerate this performance by an order of magnitude).  We validate our pipeline, named
+PhotoD (in analogy with photo-z, photometric redshifts of galaxies) using both simulated catalogs, and SDSS, DECam, and Gaia data. 
+We intend to make LSST-based value-added PhotoD catalogs publicly
+available via Rubin Science Platform with every LSST Data Release. 
 \end{abstract}
@@
 
-%% From the front matter, we move on to the body of the paper.
-%% Sections are demarcated by \section and \subsection, respectively.
-%% Observe the use of the LaTeX \label
-%% command after the \subsection to give a symbolic KEY to the
-%% subsection for cross-referencing in a \ref command.
-%% You can use LaTeX's \ref and \label commands to keep track of
-%% cross-references to sections, equations, tables, and figures.
-%% That way, if you change the order of any elements, LaTeX will
-%% automatically renumber them.
-%%
-%% We recommend that authors also use the natbib \citep
-%% and \citet commands to identify citations.  The citations are
-%% tied to the reference list via symbolic KEYs. The KEY corresponds
-%% to the KEY in the \bibitem in the reference list below.
-
 \include{intro}
@@
 \include{methodology}
+
+%\include{methodologyZI}
 
@@
 
+\acknowledgements
+\v{Z}.I. acknowledges funding by the Fulbright Foundation and thanks the Ru\d er Bo\v{s}kovi\'{c} Institute for hospitality.
+
+Funding for the SDSS and SDSS-II has been provided by the Alfred
+P. Sloan Foundation, the Participating Institutions, the National
+Science Foundation, the U.S. Department of Energy, the National
+Aeronautics and Space Administration, the Japanese Monbukagakusho, the
+Max Planck Society, and the Higher Education Funding Council for
+England. The SDSS Web Site is \url{https://www.sdss.org/}.
+
+This work has made use of data from the European Space Agency (ESA)
+mission Gaia \url{https://www.cosmos.esa.int/gaia}, processed by the Gaia
+Data Processing and Analysis Consortium (DPAC,
+\url{https://www.cosmos.esa.int/web/gaia/ dpac/consortium}). Funding for the
+DPAC has been provided by national institutions, in particular, the
+institutions participating in the Gaia Multilateral Agreement.
+
+\vspace{5mm}
+\facilities{Sloan Digital Sky Survey, Gaia}
+\software{numpy \citep{numpy}, matplotlib \citep{matplotlib}, seaborn \citep{seaborn}, scipy \citep{scipy}, pandas \cite{pandas}, 
+                astropy \citep{astropy-1, astropy-2}, astroML \citep{2012cidu.conf...47V}, Jupyter \citep{jupyter}, Python \citep{python}.}
+
 \appendix
+%\include{appendix}
 
-\bibliography{PhotoD}{}
+\bibliography{paper}{}
 \bibliographystyle{aasjournal}
 
-%% Include this line if you are using the \added, \replaced, \deleted
-%% commands to see a summary list of all changes at the end of the article.
-%\listofchanges
+\end{document}
 
-\end{document}
+	paper/Makefile
+	paper/aastex.cls
+	paper/aastex63.cls
+	paper/apj.bst
+	paper/appendix.tex
+	paper/deluxetable.sty
+	paper/lsst.sty
+	paper/methodologyZI.tex
+	paper/paper.bib
+	paper/paper.pdf
+	paper/paper.tex
+	paper/results.tex
+
+        plots/2Dmap.png
+	plots/2Dmap_Stripe82_MrSDSS_MrGaia.png
+	plots/2Dmap_Stripe82_MrSDSS_MrGaia_piSNR20.png
+	plots/2Dmap_Stripe82_MrdMr_piSNR20.png
+	plots/QrCornerPlot.png
+	plots/bayesPanels.png
+	plots/cornerPlot3.png
+	plots/margPosteriors.png
+	plots/margPosteriors3D.png
+	plots/qpB_Ar.png
+	plots/qpB_FeH.png
+	plots/qpB_Mr.png
+	plots/qpB_chiTest4_3D_Ar.png
+	plots/qpB_chiTest4_3D_FeH.png
+	plots/qpB_chiTest4_3D_Mr.png
+	plots/qpB_chiTest4_FeH.png
+	plots/qpB_chiTest4_Mr.png
+	plots/qpBcmd.png
+	plots/qpBcmd_chiTest4_3D_Ar.png
+	plots/qpBcmd_chiTest4_3D_FeH.png
+	plots/qpBcmd_chiTest4_3D_Mr.png
+	plots/qpBcmd_chiTest4_FeH.png
+	plots/qpBcmd_chiTest4_Mr.png
+	plots/qpBmeans_chiTest4_.png
+	plots/qpBmeans_chiTest4_3D.png

### 2024-01-08 paper/appendix.tex
revised style

\leftline{\bf Validation of proper motion systematic and random uncertainties using quasars}
\vskip 0.1in

We tested Gaia's proper motions and their uncertainties using
spectroscopically confirmed quasars from SDSS Data Release 7. There
are $\sim$367,000 SDSS quasars with Gaia's non-negative proper motion errors.
% (note that columns GAIA\_PM\_RA\_ERR and GAIA\_PM\_DEC\_ERR actually list inverse variance).
Their median  proper motion per
coordinate is about 0.01 mas/yr (indicating no substantial systematic
measurement errors) and the median proper motion magnitude
is about 1.1 mas/yr (indicating typical measurement uncertainty; the
median magnitude of this sample is $G\sim$20; for the FGKM sample
analyzed here, with $G<18$, the proper motion uncertainties are
$<0.15$ mas/yr).

We have verified that the width of  proper motion per
coordinate normalized by reported uncertainties (i.e. the width of corresponding
$\chi$ distributions) is 1.07 and 1.09, demonstrating Gaia's reliable
estimates of measurement uncertainties.

We didn't find any significant variation of the median quasar proper motion per
coordinate with position on the sky. The only ``interesting feature"
in the data is increased scatter of proper motion per
coordinate measurements in the so-called SDSS Stripe 82 region by
about 50\% compared to the rest of SDSS sky. This effect is easily understood as due to deeper quasar
sample in that region (due to details in SDSS spectroscopic target
selection) and the increase of Gaia's measurement uncertainties with
magnitude (and verified through no substantial increase in the
corresponding $\chi$ distributions).

### 2024-01-08 paper/paper.tex
abstract edits

@@
 \begin{abstract}
-As demonstrated with SDSS, Pan-STARRS and most recently with Gaia data, broad-band near-UV to near-IR photometry can be used 
-to estimate distances, metallicity and the interstellar dust extinction along the line of sight for stars in the Galaxy. Anticipating
-photometric catalogs with tens of billions of stars from Rubin's Legacy Survey of Space and Time (LSST), we present a fully Bayesian procedure 
+As demonstrated with SDSS, Pan-STARRS and most recently with Gaia data, broad-band near-UV to near-IR stellar photometry can be used 
+to estimate distance, metallicity and interstellar dust extinction along the line of sight for stars in the Galaxy. Anticipating
+photometric catalogs with tens of billions of stars from Rubin's Legacy Survey of Space and Time (LSST), we present a Bayesian procedure 
 and pipeline that build on previous work and can handle LSST-sized datasets. Likelihood computations utilize MIST/Dartmouth isochrones
-and priors are derived from TRILEGAL-based simulated LSST catalogs from Dal Tio et al. (2022). The computation speed is about 10 mas per
-star for both optimized grid search and MCMC methods; we show in a companion paper by Mrakov\v{c}i\'{c} et al. how to utilize neural
-networks to accelerate this performance by an order of magnitude).  We validate our pipeline, named
-PhotoD (in analogy with photo-z, photometric redshifts of galaxies) using both simulated catalogs, and SDSS, DECam, and Gaia data. 
+and priors are derived from TRILEGAL-based simulated LSST catalogs from Dal Tio et al. (2022). The computation speed is about 10 milliseconds per
+star on a single core for both optimized grid search and MCMC methods; we show in a companion paper by Mrakov\v{c}i\'{c} et al. how to utilize neural
+networks to accelerate this performance by an order of magnitude.  We validate our pipeline, named
+PhotoD (in analogy with photo-z, photometric redshifts of galaxies) using both simulated catalogs, and SDSS, DECam and Gaia data. 
 We intend to make LSST-based value-added PhotoD catalogs publicly
@@
 \end{document}
-
-	paper/Makefile
-	paper/aastex.cls
-	paper/aastex63.cls
-	paper/apj.bst
-	paper/appendix.tex
-	paper/deluxetable.sty
-	paper/lsst.sty
-	paper/methodologyZI.tex
-	paper/paper.bib
-	paper/paper.pdf
-	paper/paper.tex
-	paper/results.tex
-
-        plots/2Dmap.png
-	plots/2Dmap_Stripe82_MrSDSS_MrGaia.png
-	plots/2Dmap_Stripe82_MrSDSS_MrGaia_piSNR20.png
-	plots/2Dmap_Stripe82_MrdMr_piSNR20.png
-	plots/QrCornerPlot.png
-	plots/bayesPanels.png
-	plots/cornerPlot3.png
-	plots/margPosteriors.png
-	plots/margPosteriors3D.png
-	plots/qpB_Ar.png
-	plots/qpB_FeH.png
-	plots/qpB_Mr.png
-	plots/qpB_chiTest4_3D_Ar.png
-	plots/qpB_chiTest4_3D_FeH.png
-	plots/qpB_chiTest4_3D_Mr.png
-	plots/qpB_chiTest4_FeH.png
-	plots/qpB_chiTest4_Mr.png
-	plots/qpBcmd.png
-	plots/qpBcmd_chiTest4_3D_Ar.png
-	plots/qpBcmd_chiTest4_3D_FeH.png
-	plots/qpBcmd_chiTest4_3D_Mr.png
-	plots/qpBcmd_chiTest4_FeH.png
-	plots/qpBcmd_chiTest4_Mr.png
-	plots/qpBmeans_chiTest4_.png
-	plots/qpBmeans_chiTest4_3D.png

### 2024-01-19 paper/analysis.tex
more work on Methods

@@
 % -- kuglasti skupovi
+
+XXX moved from Methods, use when discussing 2-D posteriors...
+
+Another important degeneracy arises from the fact that even for a fixed \teff and \feh, the \logg and thus the luminosity are not uniquely determined by the colors: a degeneracy may exist between the giant branch and the main sequence as the colors constructed from \textit{ugrizy} bands are not sensitive to \logg.

### 2024-01-19 paper/discussion.tex
more work on Methods

@@
 Therefore, given the observed colors it is necessary to simultaneously estimate the effective temperature, metallicity and extinction.
+
+XXX  old text from methods 
+The best photometric estimators of metallicity are colors whose shorter-wavelength component includes the metal absorbtion bands at near-UV wavelengths, short of Balmer break (300 $\lessapprox\lambda$ [nm] $\lessapprox$ 400).
+Therefore, the LSST has a comparative advantage over the surveys lacking \textit{u}-band measurements, and could provide accurate distances within the range of 5-10\%.
+\magcom{A plot of model spectra, fixed, \textit{log(g)} and \textit{T\textsubscript{eff}}, several different metallicities?}
+
+Likelihood computations utilize MIST/Dartmouth isochrones
+and priors are derived from TRILEGAL-based simulated LSST catalogs from Dal Tio et al. (2022). The computation speed is about 10 milliseconds per
+star on a single core for both optimized grid search and MCMC methods; we show in a companion paper by Mrakov\v{c}i\'{c} et al. how to utilize neural
+networks to accelerate this performance by an order of magnitude.  We validate our pipeline, named
+PhotoD (in analogy with photo-z, photometric redshifts of galaxies) using both simulated catalogs, and SDSS, DECam and Gaia data. 
+We intend to make LSST-based value-added PhotoD catalogs publicly
+available via Rubin Science Platform with every LSST Data Release.
+
+In this paper we describe a method that will deliver LSST-based stellar distance estimations complementary to \textit{Gaia's} state-of-the-art trigonometric parallaxes and reach about 10 times further, to approximately 100 kpc.
+These results will be transformative for the studies of the Milky Way in general, and the stellar and the dark matter halo in particular as never before was there a survey that simultaneously observed roughly two thirds of the sky, to the co-added depth of \textit{r}$\approx$26 mag.
+
+\magcom{A bit about the importance of the distance estimation in the MW, dust implications (for extragalactic science too).}
+
+There are a variety of astronomical methods to estimate distances to stars, ranging from direct geometric (trigonometric) methods for nearby stars to indirect methods based on astrophysics for more distant stars.
+
+\magcom{Mention \cite{bailer-jones_estimating_2021}, \cite{gordon_panchromatic_2016}, \cite{green_measuring_2014,green_three-dimensional_2015,green_3d_2019}, \cite{juric_milky_2008} and \cite{lallement_3d_2014}, \cite{queiroz_starhorse_2018}}.
+
+Layout of the paper is...
+
+ assuming approximately 10 billion stars in the LSST stellar catalog as predicted by TRILEGAL, and a 100-core machine, a distance catalog could be produced in about 10 days (10 millisec/star).
+Input LSST photometry (with errors) and minimalistic auxiliary metadata  for 10 billion stars would amount to about 1 TB of data; the code outputs, including a covariance matrix of the three fitted parameters, would have approximately the same volume.
+
+ Thanks to the \textit{Vera C.
+Rubin} observatory's \textit{Legacy Survey of Space and Time} (LSST), for the first time in history, an astronomical catalog will contain more Milky Way stars than there are living people on Earth -- of the order 10-20 billion, depending on model assumptions.
+In order to map the Milky Way in three dimensions, distances to these stars must be accurately  estimated.

### 2024-01-19 paper/intro.tex
Minor edits to the text.
more work on Methods

@@
 
-Thanks to the \textit{Vera C.
-Rubin} observatory's \textit{Legacy Survey of Space and Time} (LSST), for the first time in history, an astronomical catalog will contain more Milky Way stars than there are living people on Earth -- of the order 10-20 billion, depending on model assumptions.
-In order to map the Milky Way in three dimensions, distances to these stars must be accurately  estimated.
-In this paper we describe a method that will deliver LSST-based stellar distance estimations complementary to \textit{Gaia's} state-of-the-art trigonometric parallaxes and reach about 10 times further, to approximately 100 kpc.
-These results will be transformative for the studies of the Milky Way in general, and the stellar and the dark matter halo in particular as never before was there a survey that simultaneously observed roughly two thirds of the sky, to the co-added depth of \textit{r}$\approx$26 mag.
+In order to map the Milky Way in three dimensions, distances to its stars must be accurately  estimated.
+Kinematic studies based on proper motion data also require estimates of stellar distances.  There are a variety
+of astronomical methods to estimate distances to stars, ranging from direct geometric (trigonometric) methods
+for nearby stars to indirect methods based on astrophysics for more distant stars.
 
-\magcom{A bit about the importance of the distance estimation in the MW, dust implications (for extragalactic science too).}
+As demonstrated with SDSS \citep{2008ApJ...673..864J, 2008ApJ...684..287I}, Pan-STARRS \citep{2014ApJ...783..114G, green_3d_2019} and most recently with
+Gaia data \citep{bailer-jones_estimating_2021}, broad-band near-UV to near-IR stellar photometry can be used 
+to estimate distance, metallicity and interstellar dust extinction along the line of sight for stars in the Galaxy.
+In analogy with photo-z, photometric redshifts of galaxies, hereafter we refer to these methods as photo-D. 
+The photo-D method is conceptually quite simple: multi-dimensional color tracks (either empirical or model-based),
+parametrized by luminosity, metallicity and extinction, are fit to observed colors and the best-fit produces estimates
+of the three model parameters. The method relies on strong correlations between stellar colors and
+stellar luminosity for dominant stellar populations such as main-sequence stars, red giants, white dwarfs, and even
+for the majority of unresolved binary stars. These are the same correlations that are responsible for the abundant
+structure seen in the Hertzsprung-Russell diagram. In addition to colors and luminosity, these correlations
+also involve metallicity in case of main-sequence and red giant stars, and surface gravity in case of white dwarfs.
+For the youngest main-sequence stars, stellar age may play a role, too. Measured stellar colors are also affected
+by interstellar dust extinction along the line of sight towards the star. Consequently, measurements of apparent
+brightness and sufficient number of UV to IR colors, such as those that Rubin Observatory's Legacy Survey of Space
+and Time (LSST; \citealt{2019ApJ...873..111I}) will provide, can be used to accurately estimate these parameters,
+and ultimately stellar distances.
 
-There are a variety of astronomical methods to estimate distances to stars, ranging from direct geometric (trigonometric) methods for nearby stars to indirect methods based on astrophysics for more distant stars.
+LSST-based stellar distance estimates will significantly improve available distance catalogs, such as those recently produced by \cite{green_3d_2019}
+and \cite{bailer-jones_estimating_2021}. First, the sample size will be increased by more than an order of magnitude, and exceed 10 billion stars.
+Distance accuracy for stars with sufficiently small photometric errors will be within the 5-10\% range, or about twice
+as accurate as for surveys lacking the UV $u$ band (which provides metallicity constraints). LSST-based stellar distances
+will reach about 10 times further than Gaia’s color-based distances and will be transformative for studies of the
+Milky Way in general, and of its stellar and dark matter halo in particular.
 
-\magcom{Mention \cite{bailer-jones_estimating_2021}, \cite{gordon_panchromatic_2016}, \cite{green_measuring_2014,green_three-dimensional_2015,green_3d_2019}, \cite{juric_milky_2008} and \cite{lallement_3d_2014}, \cite{queiroz_starhorse_2018}}.
-
-Layout of the paper is...
+In this paper we present a Bayesian procedure and pipeline that build on previous work and can handle LSST-sized datasets.
+In \S2, we describe methodology and in \S3 we test the pipeline using both simulated catalogs, and SDSS, DECam and Gaia
+photometry. We discuss possibilities for further improvements and catalog public release plans in \S4.

### 2024-01-19 paper/paper.tex
more work on Methods

@@
 
-\title{PhotoD with LSST: stellar photometric distances out to the edge of the Galaxy}
+\title{Photo-D with LSST: stellar photometric distances out to the edge of the Galaxy}
 
@@
 
-\author[0000-0000-0000-0000]{Bob Abel}
-\affiliation{The Paradise Island, WA, USA }
+\author[0000-0001-8418-3083]{Bob Abel}
+\affiliation{Olympic College, 1600 Chester Avenue, Bremerton, WA
+98337, USA, and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}
 
@@
 
-\author[0000-0000-0000-0000]{Filip Matkovi\'{c}}
-\affiliation{Hvar Observatory, 10000 Zagreb, Croatia}
+\author[0009-0002-5858-585X]{Filip Matkovi\'{c}}
+\affiliation{Hvar Observatory, Faculty of Geodesy, University of Zagreb, Ka\v{c}i\'{c}eva 26, 10000 Zagreb, Croatia}
 
@@
 
-\author[0000-0000-0000-0000]{Toni \v{S}ari\'{c}}
-\affiliation{University of Split, Croatia}
+\author[0000-0001-8731-8369]{Toni \v{S}ari\'{c}}
+\affiliation{University of Split - FESB, R. Bo\v{s}kovi\'{c}a 32, 21000 Split, Croatia}
 
 \author[0000-0000-0000-0000]{Tomislav Jurki\'{c}}
-\affiliation{University of Rijeka, Croatia}
+\affiliation{Faculty of Physics, University of Rijeka, Radmile Matej\v{c}i\'{c} 2, 51000 Rijeka, Croatia}
 
@@
 
-\author[0000-0000-0000-0000]{Forgotten Collaborator}
-\affiliation{Somewhere}
+%\author[0000-0000-0000-0000]{Forgotten Collaborator}
+%\affiliation{Somewhere}
 
@@
 networks to accelerate this performance by an order of magnitude.  We validate our pipeline, named
-PhotoD (in analogy with photo-z, photometric redshifts of galaxies) using both simulated catalogs, and SDSS, DECam and Gaia data. 
-We intend to make LSST-based value-added PhotoD catalogs publicly
+PhotoD (in analogy with photo-z, photometric redshifts of galaxies)
+using both simulated catalogs, and SDSS, DECam and Gaia photometry. We intend to make LSST-based value-added PhotoD catalogs publicly
 available via Rubin Science Platform with every LSST Data Release. 
@@
 \include{methodology}
-
-%\include{methodologyZI}
 

### 2024-01-19 paper/methodology.tex
more work on Methods

--- 2023-11-08 paper/methodology.tex
@@
-\section{Method} \label{sec:method}
+\section{Methodology} \label{sec:method}
 
-The photometric distance estimation method (hereafter \pd) is conceptually quite simple and relies on the strong correlations between the stellar colors and spectral energy distributions (SED) for dominant stellar populations.
-The SEDs, and consequently colors, are determined by the effective temperature (\teff), the surface gravity (usually denoted as \logg), and the metallicity (\mh), or alternatively, by the absolute magnitude in band \textit{b} (\Mb), \mh and age.
+In this Section we discuss a Bayesian method for stellar photometric distance estimation and its implementation.
+We start with a brief overview of Bayesian methodology and then discuss in detail our choices of likelihoods
+and priors, and how they differ from previous work. We conclude this section with a discussion of numerical
+pipeline implementation and discuss its performance in the next Section.
 
-The distributions that describe these correlations are obtained either from models or from observations.
-For example, the distribution of stellar SEDs in the color-color diagram in the middle and left panels of Figure~\ref{fig:3dataDiags} provide key insights in stellar evolution and classification of different stellar populations such as main-sequence stars, giant stars, white dwarf stars, a majority of unresolved binary stars and even extragalactic objects.
-Analogous distributions are responsible for the abundant structure seen in the Hertzsprung-Russell diagram (HRD).
+\subsection{Bayesian Approach to Stellar Photometric Distance Estimation}
 
-\begin{figure*}[ht!]
+In most general terms, our aim is for each star to estimate an array (a vector) of model parameters $\vec{\theta}$,
+for some model $M$ (e.g., main-sequence stars), using data vector $\vec{D}$ and priors for $M$ and $\vec{\theta}$.
+Data, or observations, include multi-band photometry that is used
+to construct colors, $\vec{c}$. In case of SDSS, colors include $u-g$, $g-r$, $r-i$, $i-z$, and in case of LSST also
+$z-y$. In addition to colors, $\vec{D}$ also includes an apparent magnitude and hereafter we choose the $r$-band
+magnitude. Therefore, $\vec{D}$ = ($r$, $\vec{c}$). Observations also provide stellar sky coordinates and we
+address their role further below when discussing priors.
+
+Models $M$ (either empirical or computational, see below) need to provide stellar colors as functions of model parameters
+$\vec{\theta}$. 
+The three principal parameters that control stellar colors at a fixed stellar age and at the accuracy level relevant here
+($\sim$1 \%) include absolute magnitude (here chosen in the $r$ band, $M_r$), metallicity ($[Fe/H]$) and surface
+gravity. In case of main-sequence stars and red giants, the color tracks can be expressed as functions of $M_r$
+and $[Fe/H]$ along an isochrone, without having to explicitly specify surface gravity. With other populations,
+such as white dwarfs, the roles of metallicity and surface gravity are different; we will
+assume here for notational simplicity that intrinsic stellar colors depend on $M_r$ and $[Fe/H]$. We do not
+consider stellar age as a model parameter and use it as a model label for reasons discussed in detail in the next section.
+
+\begin{figure*}[t!]
 \plotone{plot3diagsData.png}
-	\caption{The blue dots in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
-A subset	of 415,000 stars with $r < 22$ and $u<22$ are shown as red dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ are shown as cyan dots.
-Finally, 63,000 stars that have signal-to-noise ratio for Gaia’s parallax measurements of at least 20 are shown as green dots.
-The same color scheme is used in other two panels.
+\caption{An illustration of multiple stellar populations. The red dots in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
+(variable stars are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
+A subset of 415,000 stars with $r < 22$ and $u<22$ are overplotted as blue dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ (dominated by main-sequence stars and red giants) are overplotted as cyan dots.
+Finally, 63,000 stars that have signal-to-noise ratio for Gaia’s parallax measurements of at least 20 are shown as green dots (these stars can be used for the calibration of luminosity-color relations).
+The same symbol color scheme is used in other two panels.
 The three yellow lines in the middle panel show stellar locus parametrization used by Green et al.
-(2014) for three values of metallicity (left to right): $[Fe/H] = -2.0, -1.0, 0.0$.
+(2014) for three values of metallicity (left to right): $[Fe/H] = -2, -1, 0$.
 In the right panel, the impact of metallicity on color-color tracks is negligible and all three are indistinguishable from each other.
-} \label{fig:3dataDiags}
+In the bottom of the middle panel, at $-0.5 < g-r < 0$, the three dark blue sequences correspond to (from left to right) He white dwarfs, H white dwarfs, and blue horizontal branch stars.
+The clouds of pale blue dots visible above the main stellar locus in the middle and right panels correspond to unresolved binary stars \citep{2004ApJ...615L.141S}.} \label{fig:3dataDiags}
 \end{figure*}
 
-Metallicity is an important factor in these correlations, as it has a strong effect on the luminosity of the stars.
-This is reflected in the width of the main stellar loci of the two-color diagrams (middle and right hand panels in Figure~\ref{fig:3dataDiags} and the color-absolute magnitude diagrams (CAMD) in Figure~\ref{fig:3HRdiags}.
-The best photometric estimators of metallicity are colors whose shorter-wavelength component includes the metal absorbtion bands at near-UV wavelengths, short of Balmer break (300 $\lessapprox\lambda$ [nm] $\lessapprox$ 400).
-Therefore, the LSST has a comparative advantage over the surveys lacking \textit{u}-band measurements, and could provide accurate distances within the range of 5-10\%.
-\magcom{A plot of model spectra, fixed, \textit{log(g)} and \textit{T\textsubscript{eff}}, several different metallicities?}
+The observed stellar colors also depend on the interstellar dust extinction along the line of sight. Hereafter
+we will assume (and justify further below) that dust extinction is fully specified by a single model parameter, $A_r$.
+$A_r$ is extinction in the $r$ band and extinction in other bands is proportional to $A_r$, with known constants
+of proportionality (this assumption can be relaxed, see below). Once $M_r$ and $A_r$ are constrained, distance
+modulus $\Delta$ can be computed from
+\begin{equation}
+  \label{eq:distmod}
+                   r = M_r + A_r + \Delta,
+\end{equation}
+or in probabilistic form
+\begin{equation}
+  \label{eq:distmodpdf}
+                       p(\Delta) = r - p(M_r + A_r). 
+\end{equation}
+
+Therefore, $\vec{D}$ = ($r$, $\vec{c}$) and $\vec{\theta}$ = ($M_r$, $[Fe/H]$, $A_r$). Data $\vec{D}$ and
+model parameters $\vec{\theta}$ are related via the Bayes theorem (see, e.g., Chapter 5 in \citealt{2020sdmm.book.....I}),
+\begin{equation}
+  \label{eq:BayesFull}
+         p(M,\vec{\theta}|D,I) = {  p(D|M,\vec{\theta},I) \, p(M,\vec{\theta}|I) \over p(D|I)}.
+\end{equation}
+where $I$ is prior information. Strictly speaking, the vector $\vec{\theta}$
+should be labeled by $M$ since different models may be described by different parameters (e.g., main-sequence
+stars vs. white dwarfs).
+
+The result $p(M,\vec{\theta}|D,I)$ is called the {\it posterior} pdf for model $M$ {\it and} parameters
+$\vec{\theta}$, given data $D$ and other prior information $I$. This term is a $(k+1)$-dimensional
+pdf in the space spanned by $k=3$ model parameters and the model index $M$. The term $p(D|M,\vec{\theta},I)$
+is the {\it likelihood} of data {\it given} some model $M$ and given some fixed values of
+parameters $\vec{\theta}$ describing it, and all other prior information $I$. The term $p(M,\vec{\theta}|I)$
+is the a priori joint probability, or simply prior, for model $M$ and its parameters $\vec{\theta}$ in the absence of any
+of the data used to compute likelihood. The prior can be expanded as
+\begin{equation}
+ \label{eq:BayesPriorExpand}
+        p(M,\vec{\theta}|I) = p(\vec{\theta}|M,I) \,p(M|I).
+\end{equation}
+
+The term $p(D|I)$ is the {\it probability of data}, or the prior predictive probability for $D$. 
+It provides proper normalization for the posterior  pdf and usually it is not explicitly computed
+when estimating model parameters: rather, $p(M,\vec{\theta}|D,I)$ for a given $M$
+is simply renormalized so that
+its integral over all model parameters $\vec{\theta}$ is unity. The integral of the prior $p(\vec{\theta}|M,I)$
+over all parameters should also be unity, but for the same reason, calculations of the posterior pdf are
+often done with an arbitrary normalization. An important exception is model selection discussed further
+below, where the correct normalization of the product $p(D|M,\vec{\theta},I) \, p(\vec{\theta}|M,I)$ is crucial.
+
+This approach is essentially the same as used in recent papers\footnote{This approach greatly simplifies when studying
+faint and distant blue halo stars, as in e.g., \cite{2008ApJ...673..864J}. Such stars are beyond the dust layer which is confined close
+to the disk and thus $A_r$ can be obtained from IR maps, have halo metallicities ($[Fe/H] \sim -1.5$), and can be assumed dominated
+by main-sequence stars. As a result, a simple functional relationship, $M_r = f(g-i)$, or its generalized version that
+accounts for the shift of $M_r$ as a function of metallicity \citep{2008ApJ...684..287I} can be used to estimate distance
+in a straightforward manner.}
+by, e.g., \cite{2011MNRAS.411..435B},  \cite{2014ApJ...783..114G},  \cite{green_3d_2019} and \cite{bailer-jones_estimating_2021}.
+The main differences compared to these works include
+\begin{enumerate}
+\item The use of multiple stellar populations, in addition to main-sequence stars and red giants (white dwarfs, unresolved binary stars,
+                potentially quasars and RR Lyrae stars, which can also be recognized and rejected using variability).
+\item Improved color tracks for main-sequence stars and (especially) red giants, including the use of very young ($<$1 Gyr) populations 
+               and an extended $[Fe/H]$ range. 
+\item Priors based on sophisticated TRILEGAL simulations by \cite{2022ApJS..262...22D} that include multiple stellar populations and
+               also account for the Galaxy's bulge component. 
+\end{enumerate}
+
+We discuss these improvements in detail in the next few sections.
+
+\subsection{Likelihood Computation}
+
+Given chosen model (population) $M$, the likelihood $p(D|M,\vec{\theta},I)$ can be explicitly written as
+\begin{equation}
+          \mathcal{L} \equiv p(\vec{c}|M_r, [Fe/H], A_r).
+\end{equation}
+Assuming Gaussian photometric errors that are parametrized by a vector of color uncertainties $\vec{\sigma}$,
+the log-likelihood is given by
+\begin{equation}
+       ln (\mathcal{L}) = constant - {1 \over 2} \,\sum_{i=1}^N \left({ c^{obs}_i - c^{mod}_i  \over \sigma_i } \right)^2 \,
+\end{equation}
+where summation is over all colors (for example, $N=4$ for SDSS, and $N=5$ for LSST), \ensuremath{c^{obs}_i} are the
+observed colors and \ensuremath{c^{mod}_i} are the model colors (they are functions of $M_r$, $[Fe/H]$, and $A_r$
+but for notational simplicity we don't explicitly list model parameters). The model colors can be computed as
+\begin{equation}
+                       \vec{c}^{\,\,mod}  = \vec{c}_0(M_r, [Fe/H]) + \vec{\delta c}(A_r),
+\end{equation}
+where $\vec{c_0}(M_r, [Fe/H])$ are intrinsic stellar colors for a given stellar population and $\vec{\delta c}(A_r)$ are
+color corrections due to dust reddening.
+
+Multiple stellar populations are illustrated in Figure~\ref{fig:3dataDiags}. With accurate multi-band photometry that
+includes a UV (here SDSS $u$) band main-sequence stars, red giants, white dwarfs, blue horizontal branch stars, and
+unresolved binary stars can be identified (UV photometry is also crucial for constraining metallicity). We first discuss
+our choice of $\vec{c_0}(M_r, [Fe/H])$ for main-sequence stars and red giants and then for white dwarfs and unresolved binary stars.
+
+\subsubsection{Empirical Luminosity-Color Tracks for Main-sequence Stars and Red Giants}
+
+Both \cite{2012ApJ...757..166B} and \cite{2014ApJ...783..114G} used empirical color tracks for main-sequence stars
+and red giants (see the left panel in Figure~\ref{fig:3HRdiags}, modeled after Figure 1 in \citealt{2014ApJ...783..114G})
+derived from SDSS data for globular clusters (for technical details, see
+Appendix A in \citealt{2008ApJ...684..287I}). These color tracks suffer from three problems. First, as can be seen in Figure 11 from
+\cite{2014ApJ...783..114G}, their predicted colors for stars between main-sequence turn-off and red giant branch are too blue by
+about 0.1--0.2 mag. Second, their metallicity grid does not extend to the $[Fe/H]>0$ range relevant for some disk stars. Finally, they
+correspond to very old populations (older than a few Gyr) and cannot be used for stars younger than about 1 Gyr.
 
@@
 \plotone{plot3HRdiags.png}
-	\caption{The left panel shows the absolute magnitude vs.
-color parametrization for main sequence and red giant stars.
-The symbols are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0.0 (blue to red).
-The three lines correspond to three values of metallicity: $[Fe/H] = -2.0, -1.0, 0.0$ (dot-dashed, solid and dashed, respectively).
-The middle panel shows a sample of 63,000 stars that have signal-to-noise ratio for Gaia’s parallax measurements of at least 20 (white dwarfs can be seen in the lower left corner).
-The dot-dashed, solid and dashed black lines are the same as in the left panel.
-For comparison, the dotted lines were computed using eqs.
-A2 and A7 from Ivezic et al.
-(2008).
+\caption{The left panel shows SDSS-based empirical absolute magnitude vs. color parametrization for main-sequence stars and red giants.
+The symbols are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
+metallicity: $[Fe/H] = -2, -1, 0$ (dot-dashed, solid and dashed, respectively). The middle panel shows a sample of 63,000 stars that have
+signal-to-noise ratio for Gaia’s parallax measurements of at least 20 (white dwarfs can be seen in the lower left corner). The dot-dashed,
+solid and dashed black lines are the same as in the left panel. For comparison, the essentially identical yellow dotted lines were computed
+using eqs. A2 and A7 from Ivezi\'{c} et al. (2008). Note the discrepancy between these parametrizations and data for sub-giant stars ($M_r \sim 3$
+and $g-i \sim 0.8$).
 The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots, and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots.
@@
 The dot-dashed, solid and dashed black lines are the same as in the left and middle panels.
-About 10,000 outliers seen at $g-i = 0.4$ and $Mr > 7$ are predominantly found at the faint end ($r>20$).}
+About 10,000 outliers (about 2.5\% of the full
+sample) seen at $g-i = 0.4$ and $Mr > 7$ are predominantly found at the faint end ($r>20$).}
 \label{fig:3HRdiags}
 \end{figure*}
+
+We use a combination of SDSS and Gaia data to demonstrate the first problem. The middle panel in Figure~\ref{fig:3HRdiags} 
+shows a clear discrepancy between SDSS-based empirical tracks and data for sub-giant stars. Nevertheless, it is noteworthy that
+for main-sequence stars the agreement is excellent. Furthermore, the right panel in Figure~\ref{fig:3HRdiags} demonstrates
+that SDSS-based photometric distances and Gaia-based ``photo-geometric'' distances from \cite{bailer-jones_estimating_2021}
+for main-sequence stars are on the ``same scale''.  XXX provide numerical results for offsets!
+
+We address all three problems by switching from the SDSS-based empirical isochrones that correspond to old
+populations to model-based isochrones that span a range of ages and a wider range of metallicities.
+
+\subsubsection{MIST/Dartmouth Isochrones for Main-sequence Stars and Red Giants}
+
+Look at the code first!
+
+We considered two sets of isochrones:
+
+\begin{figure*}[ht!]
+\plotone{compare2isochrones_augmentedSDSS_1Gyr_10Gyr.png}
+\caption{Augmented locus}
+\label{fig:augmLocus}
+\end{figure*}
+
+These color tracks account for an overwhelming majority of stars expected in LSST catalogs. Nevertheless,
+
+\subsubsection{Luminosity-Color Tracks for White Dwarfs}
+
+Bob Abel will provide text and plots.
+
+\begin{figure*}[ht!]
+\plotone{plot3diagsBA.png}
+\caption{Augmented locus}
+\label{fig:augmLocus}
+\end{figure*}
+
+\subsubsection{Luminosity-Color Tracks for Unresolved Binaries }
+
+Bob Abel will provide text and plots.
+
+\begin{figure*}[ht!]
+\plotone{plot3diagsBAzoom.png}
+\caption{Augmented locus}
+\label{fig:augmLocus}
+\end{figure*}
+
+\subsubsection{Luminosity-Color Tracks for Unaccounted Populations}
+
+Reiterate that variable objects, such as quasars and RR Lyrae are not considered here
+but can be addressed in a straighforward manner. Cite Sesar et al. papers
+
+Examples: BHB stars, AGB stars...
+
+\subsubsection{Accounting for Interstellar Dust Extinction}
+
+Following \cite{2012ApJ...757..166B}
+
+prior from IR dust maps
 
@@
 
-Another important degeneracy arises from the fact that even for a fixed \teff and \feh, the \logg and thus the luminosity are not uniquely determined by the colors: a degeneracy may exist between the giant branch and the main sequence as the colors constructed from \textit{ugrizy} bands are not sensitive to \logg.
+\subsection{TRILEGAL-based Priors}
 
-We adopt a Bayesian framework in which we simultaneously fit for \Mb, \feh and \Ar, assuming a fixed \RV value of 3.1\footnote{In principle \RV could be also fitted for.} The posterior for each individual star with LSST photometry is then given as:
+Priors bring dependence on sky position and apparent magnitude
 
-\begin{equation} \label{eq:likelihood}
-	\begin{split}
-		    &\Pcond{\Mb, \, \feh, \, \Ar}{\vec{c}} = \\
-		& \frac{\Pcond{\vec{c}}{\Mb, \, \feh, \, \Ar} P\arg{\Mb, \, \feh, \, \Ar}}{P\arg{\vec{c}}} \,
-	\end{split}
-\end{equation}
-
-with \ensuremath{\vec{c}} standing for the vector of input colors (\ul-\gl, \gl-\rl and so on). The log-likelihood is given by:
-
-\begin{equation}
-	ln (\mathcal{L}) = C - {1 \over 2} \,\sum_{i=1}^N \left({ c^{obs}_i - c^{mod}_i  \over \sigma_i } \right)^2 \,
-\end{equation}
-
-where \ensuremath{c^{obs}_i} are the observed colors and \ensuremath{c^{mod}_i} the model colors.
 The priors are established by partitioning the TRILEGAL galaxy model \citep{dal_tio_simulating_2022} in healpixels, and each of the pixel in one-magnitude wide bins in apparent magnitude.
@@
 %Our method is basically brute-force fitting with some intelligent tricks leveraged to obtain faster execution that will be required for 10B LSST stars.
+
+\subsection{Bayesian Model Selection}
+
+\subsection{Implementation: Photo-D pipeline}
+
+optimized grid search, MCMC, refer to NN paper, also in Discussion
 

### 2024-01-19 paper/results.tex
first import

\section{Results\label{sec:results}}

Here we test the validity of the Bond et al. kinematic models for disk and halo stars, as enabled by the high-metallicity FGKM sample
and the red giant sample introduced in the preceding Section.
We first discuss the behavior of the first two statistical moments (mean and dispersion) for 3-dimensional velocity distribution
and then compare the predicted and observed shapes of the rotational velocity distribution in more detail. We test
model predictions for the behavior of quantities directly measured by Gaia, stellar radial velocities and proper motions, across
the whole sky. Finally, we verify that the behavior of Gaia data for halo stars is consistent with a velocity ellipsoid that is
spatially invariant when expressed in spherical coordinates.

\subsection{Comparison of predicted and observed 3-dimensional velocity distribution for disk stars}

We first emulate the top left panels in Figures 5, 7 and 11 from the Bond et al. study. We select stars towards the North and South
Galactic poles ($|b|>80^\circ$) and show the variation of their velocity components with $|Z|$ in \autoref{fig:3v}. Due to the
galactic latitude constraint, the $v_Z$ component is dominated by the radial velocity measurements while the other two components
are dominated by proper motion measurements. Compared to the Bond et al. high-luminosity sample, the distance range probed
by Gaia's FGKM sample is closer: from about 50 pc to about 1 kpc vs. 0.8--5 kpc (at the far end, the SDSS sample is limited by
sample contamination due to halo stars).

The Bond et al. model predictions for the velocity mean and dispersion are in agreement with Gaia's observations. In particular, the
gradient of the rotational velocity with $|Z|$ is evident, and the extrapolation of SDSS rotational velocity measurements to $Z=0$
is quantitatively supported by Gaia. The same conclusions remain valid when the full sample is considered (that is, without the $|b|>80^\circ$ restriction).

The $<$1 kpc distance range probed by FGKM stars is significantly extended with the red giant sample, as illustrated in
\autoref{fig:3vRGs}. In conclusion, the Bond et al. models for the velocity mean and dispersion of disk stars appear validated
in the $|Z|$ range 0--5 kpc, without any appreciable north-south asymmetry.

\begin{figure}[!ht]
\includegraphics[width=0.999\textwidth,angle=0]{figures/plot3VvsZ_NGP.png}
\includegraphics[width=0.999\textwidth,angle=0]{figures/plot3VvsZ_SGP.png}
\caption{The variation of cylindrical velocity components with distance from the plane, $|Z|$, for FGKM stars with measured radial velocities, shown as dots. 
The top row shows 6,548 stars towards the North Galactic Pole ($b >80^\circ$) and the bottom row shows 7,578 stars towards the South Galactic Pole
($b < -80^\circ$). Triangles show the mean values in bins of $|Z|$ and the thick dashed lines show the $\pm2\sigma$ envelope around means, where
$\sigma$ is the standard deviation for each bin (i.e., velocity dispersion). The thick dot-dashed lines are models for the mean velocity (0 for $v_R$ and
$v_Z$, and given by eq.~\ref{eq:Dlag} for $v_\phi$). The thin dot-dashed lines show the $\pm2\sigma$ envelope, with $\sigma$ (velocity dispersion)
given by eqs.~\ref{eq:Ddisp}, \ref{eq:DdispR}, and \ref{DdispZ}, for $\phi$, $R$ and $Z$ panels, respectively.  The thin dashed lines in the two left panels
also show $\pm2\sigma$ envelopes, but with the velocity dispersion computed using eq.~\ref{eq:pDvPhi} (see \S~\ref{sec:vshape}).} 
\label{fig:3v} 
\end{figure}

\begin{figure}[!ht]
\includegraphics[width=0.999\textwidth,angle=0]{figures/plot3VvsZ_giantsD.png} 
\caption{Similar to \autoref{fig:3v}, except that here about five times larger distances are probed with 4.3 million red giant stars
  that have [Fe/H]$>-0.5$ and cylindrical galacto-centric radius between 7 kpc and 10 kpc. The meaning of symbols and lines
  is the same as in \autoref{fig:3v}, and exactly the same models are shown.} 
\label{fig:3vRGs} 
\end{figure}

\subsection{Comparison of predicted and observed shapes of the rotational velocity distribution for disk stars\label{sec:vshape}}

As already suggested in \cite{2008ApJ...684..287I} and confirmed by Bond et al. (see their Figure 6), the shape of the rotational velocity
distribution for disk stars is strongly non-Gaussian (skewed) and it evolves with the distance from the Galactic plane. Histograms based
on Gaia's data shown in \autoref{fig:vPhiShape} confirm this expectation. Furthermore, new data still admit detailed quantitative
modeling of observed distributions as a sum of two Gaussians, with a fixed normalization ratio and a fixed offset of their mean values (see
\autoref{eq:pDvPhi}). Since Gaia's velocity measurement errors are negligible in this context (compared to the much larger width of the observed
velocity distributions; for validation using quasar data, see Appendix), we have reoptimized fits by allowing six parameters to vary. Their best-fit values remain close to the SDSS
values: $f_1=0.40$, $f_2=0.60$, $v_0 = -186 {\rm \, km~s^{-1}}$, $\Delta v_n = 38 {\rm \, km~s^{-1}}$, $a_1 = 22 {\rm \, km~s^{-1}}$ and $a_2= 17
{\rm \, km~s^{-1}}$.

When the resulting re-optimized pdf, $p(x=v_\phi|Z)$, is used to evaluate the mean $\langle v_\phi \rangle$ and its dispersion
$\sigma_{\phi}$  as functions of $Z$, there is no discernible change for the former compared to \autoref{eq:Dlag}, while the latter is
about 20\% smaller than values given by \autoref{eq:Ddisp} (for illustration of the difference, see the two left panels in \autoref{fig:3v}).
Since there is no implied physics in this two-component statistical model of the skewed velocity distribution, minor numerical adjustments
of the few free model parameters are probably inconsequential. \\

These conclusions based on the FGKM sample are supported by the behavior of the red giant sample, as illustrated in \autoref{fig:vPhiShapeRG}. 
In particular, the vertical gradient of rotational velocity is evident.

\begin{figure}[!ht]
\includegraphics[width=0.999\textwidth,angle=0]{figures/plotVphiDistr4.png}
\caption{A comparison of the observed rotational velocity distribution ($v_\phi$) for FGKM stars, shown as histograms, and a two-component model (dashed
lines: individual components; solid line: their sum) given by \autoref{eq:pDvPhi} (using updated parameters discussed in text; the same
$Z$-dependent model is shown in all four panels). The four panels show results for two $|Z|$ bins (top row vs. bottom row) and towards North and South
($|b|>80^\circ$, left vs. right). The sample sizes are about 4,000 stars for the nearer $|Z|$ bin and about 900 stars for the more distant bin.} 
\label{fig:vPhiShape} 
\end{figure}

\begin{figure}[!ht]
\includegraphics[width=0.999\textwidth,angle=0]{figures/plotVphiDistr4rg.png}
\caption{Analogous to \autoref{fig:vPhiShape}, except for a subsample of the red giant sample shown in \autoref{fig:3vRGs} restricted to $|b|>60^\circ$.
  The top two panels show the same $|Z|$ bins as in \autoref{fig:vPhiShape}, while the two bottom panels extend to larger $|Z|$. 
  The sample sizes vary from 10,000 stars to 2,000 stars for the most distant bin.} 
\label{fig:vPhiShapeRG} 
\end{figure}

\subsection{Comparison of predicted and observed radial velocity and proper motion sky distributions for disk stars\label{sec:allsky}}

\begin{figure}[!ht]
\includegraphics[width=0.5\textwidth,angle=0]{figures/projview_pmLongData.png}
\includegraphics[width=0.5\textwidth,angle=0]{figures/projview_pmLongDMresiduals_Dslice.png}
\includegraphics[width=0.5\textwidth,angle=0]{figures/projview_pmLatData.png}
\includegraphics[width=0.5\textwidth,angle=0]{figures/projview_pmLatDMresiduals_Dslice.png}
\includegraphics[width=0.5\textwidth,angle=0]{figures/projview_radVelData.png}
\includegraphics[width=0.5\textwidth,angle=0]{figures/projview_radVelDMresiduals_DsliceRV.png}
\caption{The top two left panels show observed distributions of mean proper motion per pixel (top: longitude; middle: latitude)
for 868,859 stars in the 0.4--0.6 kpc distance bin. The bottom left panel shows the observed distribution of mean radial velocity
for a subsample of 400,782 stars with good measurements. The healpix nside=32 maps are shown in the Hammer
projection of galactic coordinates. The maps on the right show residuals (using the same color-coding scale) after
subtracting model values computed using eq.~\ref{eq:Dlag} for $v_\phi$ and with $\langle v_R \rangle = \langle v_Z \rangle = 0$.} 
\label{fig:skyComparison} 
\end{figure}

As the final test, we extrapolate the Bond et al. model from the quarter of the sky observed by SDSS and compare its predictions for
the variation of mean proper motions and radial velocities to Gaia's measurements across the whole sky. The three left panels in
\autoref{fig:skyComparison} show mean proper motion components per pixel and mean radial velocity per pixel for stars with
distances in the range 400 pc to 600 pc. Their strong variation across the sky is evident and is mostly due to projection effects of the
solar motion (and a little bit due to spatial variation of the rotational velocity component with $|Z|$). The panels on the right
show residuals after subtracting proper motion and radial velocity values predicted by the Bond et al. model described in \S~\ref{sec:BondModels}.
As evident, the model fully captures the observed behavior. The only potentially significant feature is observed for radial velocity residuals towards the Galactic
center, at about $0 < b < 30^\circ$. This region was not observed by SDSS and we are not aware of any kinematic or other stellar features
reported for that region (e.g., the famous Sgr dwarf galaxy is located at negative galactic latitudes). A possibility that these residuals
reflect faulty measurements could be, at least in principle, easily checked because most of these stars are relatively bright ($r\sim14$).

We note that the data maps for stars with distances in the range of 800 pc to 1.2 kpc look qualitatively the same, except that the magnitude
of proper motion is smaller by about a factor of two, as expected. The maps of residuals appear the same, except for the radial velocity
residual ``feature'', which is observed at about 15 degrees higher latitudes than for the first distance bin.

The red giant sample can be used to extend this comparison to distances where the rotational velocity component is much
smaller than locally. We first established that the behavior for red giants in the 0.4--0.6 kpc distance bin is essentially identical
as shown for the FGKM sample in \autoref{fig:skyComparison}. The corresponding plot for red giants in the 2.8--3.2 kpc distance bin 
is shown in \autoref{fig:skyComparisonRG}. Due to about six times larger distances,  the morphology of data panels is rather different
from that in \autoref{fig:skyComparison} (especially for radial velocity, where projection effects of rotational velocity component
dominate). Nevertheless, the model for disk kinematics described in
\S~\ref{sec:BondModels} fully captures the data behavior, as shown by vanishing maps of (data-model) residuals.

\begin{figure}[!ht]
\includegraphics[width=0.5\textwidth,angle=0]{figures/projview_pmLongDataRG.png}
\includegraphics[width=0.5\textwidth,angle=0]{figures/projview_pmLongDMresiduals_DsliceRV_RGdisk.png}
\includegraphics[width=0.5\textwidth,angle=0]{figures/projview_pmLatDataRG.png}
\includegraphics[width=0.5\textwidth,angle=0]{figures/projview_pmLatDMresiduals_DsliceRV_RGdisk.png}
\includegraphics[width=0.5\textwidth,angle=0]{figures/projview_radVelDataRG.png}
\includegraphics[width=0.5\textwidth,angle=0]{figures/projview_radVelDMresiduals_DsliceRV_RGdisk.png}
\caption{Analogous to \autoref{fig:skyComparison}, except for a sample of 1.05 million red giants with [Fe/H]$>-0.5$
  in the 2.8--3.2 kpc distance bin. Note that observed proper motions are much smaller due to about six times larger distances,
  while radial velocity variation with galactic longitude is much stronger due to the projection effects of the rotational velocity component.} 
\label{fig:skyComparisonRG} 
\end{figure}

\subsection{Comparison of predicted and observed 3-dimensional velocity distribution for halo stars}

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.999\textwidth,angle=0]{figures/plot3VvsZ_giantsH.png}
    \caption{The variation of spherical velocity components with distance from the plane, $|Z|$, for $\sim$34,000 candidate halo red giants from the solar cylinder with measured radial velocities and [Fe/H]$<-1.2$. Triangles show the mean values in bins of $|Z|$ and the thick dashed lines show the $\pm2\sigma$ envelope around means, where $\sigma$ is standard deviation for each bin (i.e., velocity dispersion). The dot-dashed lines represent the halo velocity ellipsoid model described in Section~\ref{sec:BondModelsHalo}. \label{fig:3vRGhalo}}
\end{figure}

\begin{figure}[!ht]
\includegraphics[width=0.999\textwidth,angle=0]{figures/plotVvsV2n.png}
\vskip -2in   
\includegraphics[width=0.999\textwidth,angle=0]{figures/plotVvsV2s.png}
\vskip -2in   
\caption{Illustration of the change of orientation of velocity ellipsoid in cylindrical coordinates for halo stars,
  selected to have [Fe/H]$<-1.2$ and selected from a narrow range of cylindrical coordinates, as shown above
  each panel. The number of stars is about 1,600 for bins with smaller $R$ (left) and 800 for bins on the right.
  The top row shows bins above the plane and the bottom row their symmetric counterparts below the plane
  (the Sun's position is in the middle of the figure). 
  The dashed lines mark the direction towards the Galactic center. The change of the velocity ellipsoid tilt
  in cylindrical coordinates is evident; however, the velocity ellipsoid in spherical coordinates is invariant
  throughout the probed volume, as illustrated by these four spatial bins.} 
\label{fig:haloTilt} 
\end{figure}

Bond et al. found that the kinematics of halo stars can be modeled with a triaxial velocity ellipsoid that is invariant in
spherical coordinates. Motivated by this finding, we show in
% \autoref{fig:3vRGhalo}   % I couldnt make this ref work
Figure 11 
three spherical velocity components for halo red giants as functions of distance from the Galactic plane. Unlike the strong dependence on the velocity dispersion
on $|Z|$ for disk stars seen in Figures~\ref{fig:3v} and \ref{fig:3vRGs}, velocity dispersion for halo stars as measured by
Gaia is spatially invariant, confirming earlier results by Bond et al. We note that small deviation of mean rotational
velocity from zero at $|Z|<1$ kpc seen in the left panel is probably due to sample contamination by much more numerous
disk stars (at $|Z|=0$, halo stars contribute only about 0.5\% of the total count, see Table 10 in \citealt{2008ApJ...673..864J}).

Gaia data analyzed here provide strong support for a halo velocity ellipsoid that is invariant in spherical coordinates.
When the velocity ellipsoid is expressed in cylindrical coordinates instead, a strong covariance is seen between the $v_Z$ and $v_R$
components. We illustrate this covariance in
% \autoref{fig:haloTilt}.  %I couldnt make this ref work
Figure 12. 
The observed tilt varies with position such that the
velocity ellipsoid points towards the Galactic center (see also \citealt{2019MNRAS.489..910E}).

### 2024-01-25 paper/analysis.tex
first draft of Section 2

@@
 
+\section{Implementation: Photo-D pipeline and Tests}
+
+optimized grid search, MCMC, refer to NN paper, also in Discussion
+
+Our fitting procedure is also executed on an adaptive grid, a coarse search over the parameter space is performed first in order to establish the layout of the manifold.
+However, care is taken that any possible local minima are not missed by appropriately adjusting the step size \magcom{how?}.
+The located maxima are then explored with a smaller step size (\magcom{adjusted how?}).
+
+In addition to the approach described here, we also tested Markov Chain Monte Carlo and neural network approaches that will be/are described in forthcoming/published papers.
+
+%Under the assumptions that the locus of the color-color diagram is representative of the stellar SEDs, that the shape of the extinction curve can be parametrized with a single parameter \RV
+
+%%% METHODOLOGY
+
+%Our method is basically brute-force fitting with some intelligent tricks leveraged to obtain faster execution that will be required for 10B LSST stars.
+
+ XXX Green et al. (2014) used MCMC
+
+% 2) statistics:
+
+The priors are established by partitioning the TRILEGAL galaxy model
+\citep{dal_tio_simulating_2022} in healpixels,
+and each of the pixel in one-magnitude wide bins in apparent
+magnitude.
+
+The model colors are then obtained from models based on SDSS results (\magcom{reference}), where the input for the models are the \Mr and the \feh obtained from TRILEGAL.
+Given the assumed extinction curve, these colors are then reddened up to the maximum reddening which is estimated from \cite{schlegel_maps_1998} (SFD98) maps.
+The latter upper limit is used in order to provide a realistic stop condition on the amount of extinction and reduce the processing time.
+This is usually a valid assumption because the SFD98 maps provide \textit{total} extinction along a line of sight.
+In order to account for the eventual underestimation in the SFD98 maps we increase the SFD98 extinctions by 20\%.
+
+% 3) code testing:
+
+\begin{figure*}[ht!]
+\plotone{qpBmeans_chiTest4_3D.png}
+\caption{Write caption: TRILEGAL mean values of input model params in umag vs. g-i }
+\label{fig:qpBmeans_chiTest4_3D.}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{qpBcmd_chiTest4_3D_Mr.png}
+\caption{Write caption: performance for Mr vs. true Mr and FeH}
+\label{fig:qpBcmd_chiTest4_3D_Mr}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{qpBcmd_chiTest4_3D_FeH.png}
+\caption{Write caption: performance for FeH vs. true Mr and FeH}
+\label{fig:qpBcmd_chiTest4_3D_FeH}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{qpBcmd_chiTest4_3D_Ar.png}
+\caption{Write caption: performance for Ar vs. true Mr and FeH}
+\label{fig:qpBcmd_chiTest4_3D_Ar}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{qpB_chiTest4_3D_Mr.png}
+\caption{Write caption: performance for Mr vs. estimated Mr and FeH}
+\label{fig:qpB_chiTest4_3D_Mr}
+\end{figure*}
+
+% *** data-based comparisons ***
+% Stripe 82 vs. Gaia
+% Bulge BDSD sample?
+
 XXX moved from Methods, use when discussing 2-D posteriors...

### 2024-01-25 paper/intro.tex
first draft of Section 2

@@
 As demonstrated with SDSS \citep{2008ApJ...673..864J, 2008ApJ...684..287I}, Pan-STARRS \citep{2014ApJ...783..114G, green_3d_2019} and most recently with
-Gaia data \citep{bailer-jones_estimating_2021}, broad-band near-UV to near-IR stellar photometry can be used 
+Gaia data \citep{bailer-jones_estimating_2021}, broad-band near-UV to near-IR stellar photometry is sufficient
 to estimate distance, metallicity and interstellar dust extinction along the line of sight for stars in the Galaxy.
@@
 For the youngest main-sequence stars, stellar age may play a role, too. Measured stellar colors are also affected
-by interstellar dust extinction along the line of sight towards the star. Consequently, measurements of apparent
+by interstellar dust extinction along the line of sight towards the star. Consequently, sufficiently accurate measurements of apparent
 brightness and sufficient number of UV to IR colors, such as those that Rubin Observatory's Legacy Survey of Space

### 2024-01-25 paper/methodology.tex
first draft of Section 2

@@
-\section{Methodology} \label{sec:method}
+\section{Methodology}
+\label{sec:method}
 
@@
 We start with a brief overview of Bayesian methodology and then discuss in detail our choices of likelihoods
-and priors, and how they differ from previous work. We conclude this section with a discussion of numerical
-pipeline implementation and discuss its performance in the next Section.
+and priors, and how they differ from previous work. A pipeline implementation of this method and a discussions
+of its performance are presented in the next Section.
 
@@
 
-In most general terms, our aim is for each star to estimate an array (a vector) of model parameters $\vec{\theta}$,
+In most general terms, our aim is to estimate for each star an array (a vector) of model parameters $\vec{\theta}$,
 for some model $M$ (e.g., main-sequence stars), using data vector $\vec{D}$ and priors for $M$ and $\vec{\theta}$.
@@
 assume here for notational simplicity that intrinsic stellar colors depend on $M_r$ and $[Fe/H]$. We do not
-consider stellar age as a model parameter and use it as a model label for reasons discussed in detail in the next section.
-
-\begin{figure*}[t!]
-\plotone{plot3diagsData.png}
-\caption{An illustration of multiple stellar populations. The red dots in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
-(variable stars are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
-A subset of 415,000 stars with $r < 22$ and $u<22$ are overplotted as blue dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ (dominated by main-sequence stars and red giants) are overplotted as cyan dots.
-Finally, 63,000 stars that have signal-to-noise ratio for Gaia’s parallax measurements of at least 20 are shown as green dots (these stars can be used for the calibration of luminosity-color relations).
-The same symbol color scheme is used in other two panels.
-The three yellow lines in the middle panel show stellar locus parametrization used by Green et al.
-(2014) for three values of metallicity (left to right): $[Fe/H] = -2, -1, 0$.
-In the right panel, the impact of metallicity on color-color tracks is negligible and all three are indistinguishable from each other.
-In the bottom of the middle panel, at $-0.5 < g-r < 0$, the three dark blue sequences correspond to (from left to right) He white dwarfs, H white dwarfs, and blue horizontal branch stars.
-The clouds of pale blue dots visible above the main stellar locus in the middle and right panels correspond to unresolved binary stars \citep{2004ApJ...615L.141S}.} \label{fig:3dataDiags}
-\end{figure*}
+consider stellar age as a model parameter and use it as a model label for reasons discussed below.
 
@@
 \end{equation}
-or in probabilistic form
+or in a probabilistic form
 \begin{equation}
   \label{eq:distmodpdf}
-                       p(\Delta) = r - p(M_r + A_r). 
+                       p(\Delta) = p(r) - p(M_r + A_r). 
 \end{equation}
+Note that in practice the uncertainty of the observed magnitude, $r^{obs}$, is always significantly smaller than
+the width of $p(M_r + A_r)$ (at the bright end by an order of magnitude, $\sim$0.01 mag vs. $\sim$0.1 mag).
 
@@
 
-The result $p(M,\vec{\theta}|D,I)$ is called the {\it posterior} pdf for model $M$ {\it and} parameters
-$\vec{\theta}$, given data $D$ and other prior information $I$. This term is a $(k+1)$-dimensional
+The result $p(M,\vec{\theta}|D,I)$ is called the {\it posterior} pdf (probability density function) for model $M$ {\it and}
+parameters $\vec{\theta}$, given data $D$ and other prior information $I$. This term is a $(k+1)$-dimensional
 pdf in the space spanned by $k=3$ model parameters and the model index $M$. The term $p(D|M,\vec{\theta},I)$
@@
 
-This approach is essentially the same as used in recent papers\footnote{This approach greatly simplifies when studying
+\begin{figure*}[t!]
+\plotone{plot3diagsData.png}
+\caption{An illustration of multiple stellar populations. The red dots in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
+(variable stars are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
+A subset of 415,000 stars with $r < 22$ and $u<22$ are overplotted as blue dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ (dominated by main-sequence stars and red giants) are overplotted as cyan dots.
+Finally, 63,000 stars that have signal-to-noise ratio for Gaia’s parallax measurements of at least 20 are shown as green dots (these stars can be used for the calibration of luminosity-color relations).
+The same symbol color scheme is used in other two panels.
+The three yellow lines in the middle panel show stellar locus parametrization used by Green et al.
+(2014) for three values of metallicity (left to right): $[Fe/H] = -2, -1, 0$.
+In the right panel, the impact of metallicity on color-color tracks is negligible and all three are indistinguishable from each other.
+In the bottom of the middle panel, at $-0.5 < g-r < 0$, the three dark blue sequences correspond to (from left to right) He white dwarfs, H white dwarfs, and blue horizontal branch stars.
+The clouds of pale blue dots visible above the main stellar locus in the middle and right panels correspond to unresolved binary stars \citep{2004ApJ...615L.141S}.} \label{fig:3dataDiags}
+\end{figure*}
+
+Our approach adopted here is essentially the same as used in recent papers\footnote{This approach greatly simplifies when studying
 faint and distant blue halo stars, as in e.g., \cite{2008ApJ...673..864J}. Such stars are beyond the dust layer which is confined close
-to the disk and thus $A_r$ can be obtained from IR maps, have halo metallicities ($[Fe/H] \sim -1.5$), and can be assumed dominated
+to the disk and thus $A_r$ can be obtained from IR maps, they have halo metallicities ($[Fe/H] \sim -1.5$), and can be assumed dominated
 by main-sequence stars. As a result, a simple functional relationship, $M_r = f(g-i)$, or its generalized version that
-accounts for the shift of $M_r$ as a function of metallicity \citep{2008ApJ...684..287I} can be used to estimate distance
+accounts for the shift of $M_r$ as a function of metallicity, \citep{2008ApJ...684..287I} can be used to estimate distance
 in a straightforward manner.}
 by, e.g., \cite{2011MNRAS.411..435B},  \cite{2014ApJ...783..114G},  \cite{green_3d_2019} and \cite{bailer-jones_estimating_2021}.
-The main differences compared to these works include
+The main differences compared to these works include: 
 \begin{enumerate}
 \item The use of multiple stellar populations, in addition to main-sequence stars and red giants (white dwarfs, unresolved binary stars,
-                potentially quasars and RR Lyrae stars, which can also be recognized and rejected using variability).
+                blue horizontal branch stars; potentially Miras, quasars and RR Lyrae stars too, which can also be recognized and rejected using variability).
 \item Improved color tracks for main-sequence stars and (especially) red giants, including the use of very young ($<$1 Gyr) populations 
@@
 
-Given chosen model (population) $M$, the likelihood $p(D|M,\vec{\theta},I)$ can be explicitly written as
+Given chosen model (i.e., a stellar population) $M$, the likelihood $p(D|M,\vec{\theta},I)$ can be explicitly written as
 \begin{equation}
-          \mathcal{L} \equiv p(\vec{c}|M_r, [Fe/H], A_r).
+        \mathcal{L} \equiv p(D|M,\vec{\theta},I) = p(\vec{c}|M_r, [Fe/H], A_r).
 \end{equation}
@@
 \begin{equation}
-       ln (\mathcal{L}) = constant - {1 \over 2} \,\sum_{i=1}^N \left({ c^{obs}_i - c^{mod}_i  \over \sigma_i } \right)^2 \,
+   \label{eq:lnLcolors}
+   \ln (\mathcal{L}) =  - {N \over 2} \, \ln(2\pi) - \sum_{i=1}^N \ln(\sigma_i) - {1 \over 2} \,\sum_{i=1}^N \left({ c^{obs}_i - c^{mod}_i  \over \sigma_i } \right)^2 \,
 \end{equation}
@@
 observed colors and \ensuremath{c^{mod}_i} are the model colors (they are functions of $M_r$, $[Fe/H]$, and $A_r$
-but for notational simplicity we don't explicitly list model parameters). The model colors can be computed as
+but for notational simplicity we don't explicitly list model parameters). Note that only the last sum involves model predictions
+for colors ($c^{mod}_i$).
+
+The model colors can be computed as
 \begin{equation}
+  \label{eq:reddCorr} 
                        \vec{c}^{\,\,mod}  = \vec{c}_0(M_r, [Fe/H]) + \vec{\delta c}(A_r),
@@
 where $\vec{c_0}(M_r, [Fe/H])$ are intrinsic stellar colors for a given stellar population and $\vec{\delta c}(A_r)$ are
-color corrections due to dust reddening.
+color corrections due to interstellar dust reddening. Eq.~\ref{eq:reddCorr} can be thought of as a set of 
+3-dimensional data cubes, one for each color, that map the triplet ($M_r, [Fe/H], A_r$) to that color. 
+The likelihood function can be thought of as a  3-dimensional data cube that, for a given set of observed colors $c^{obs}_i$,
+maps the triplet ($M_r, [Fe/H], A_r$) to 1-dimensional scalar, that is, $ln (\mathcal{L})$ is a scalar function of
+$M_r$, $[Fe/H]$, and $A_r$.
 
-Multiple stellar populations are illustrated in Figure~\ref{fig:3dataDiags}. With accurate multi-band photometry that
-includes a UV (here SDSS $u$) band main-sequence stars, red giants, white dwarfs, blue horizontal branch stars, and
-unresolved binary stars can be identified (UV photometry is also crucial for constraining metallicity). We first discuss
-our choice of $\vec{c_0}(M_r, [Fe/H])$ for main-sequence stars and red giants and then for white dwarfs and unresolved binary stars.
+The existance of multiple stellar populations in an SDSS photometric catalog is illustrated in Figure~\ref{fig:3dataDiags}.
+With accurate multi-band photometry that includes a UV band (here SDSS $u$), main-sequence stars, red giants,
+white dwarfs, blue horizontal branch (BHB) stars, and unresolved binary stars can be reliably identified (UV photometry
+is also crucial for constraining metallicity). We discuss our choice of $\vec{c_0}(M_r, [Fe/H])$ for main-sequence stars
+and red giants next, and then for white dwarfs, unresolved binary stars and BHB stars.
 
 \subsubsection{Empirical Luminosity-Color Tracks for Main-sequence Stars and Red Giants}
+
+\begin{figure*}[b!]
+\plotone{plot3HRdiags.png}
+\caption{The left panel shows SDSS-based empirical absolute magnitude vs. color parametrization for main-sequence stars and red giants.
+The symbols are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
+metallicity: $[Fe/H] = -2, -1, 0$ (dot-dashed, solid and dashed, respectively). The middle panel shows a sample of 63,000 stars that have
+signal-to-noise ratio for Gaia’s parallax measurements of at least 20 (white dwarfs can be seen in the lower left corner; same sample is
+shown by green dots in Figure~\ref{fig:3HRdiags}). Their absolute magnitude are derived from parallax measurements. The dot-dashed,
+solid and dashed black lines are the same as in the left panel. For comparison, the essentially identical yellow dotted lines were computed
+using eqs. A2 and A7 from Ivezi\'{c} et al. (2008). Note the discrepancy between these parametrizations and data for sub-giant stars ($M_r \sim 3-4$
+and $g-i \sim 0.8-1.1$).
+The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots (shown by blue dots in Figure~\ref{fig:3HRdiags}), and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots.
+Their absolute magnitudes were computed
+using the so-called “photo-geometric” distances from Bailer-Jones et al. (2021). The dot-dashed, solid and dashed black lines are the
+same as in the left and middle panels. About 10,000 outliers (about 2.5\% of the full
+sample) seen at $g-i = 0.4$ and $Mr > 7$ are predominantly found at the faint end ($r>20$).}
+\label{fig:3HRdiags}
+\end{figure*}
 
@@
 Appendix A in \citealt{2008ApJ...684..287I}). These color tracks suffer from three problems. First, as can be seen in Figure 11 from
-\cite{2014ApJ...783..114G}, their predicted colors for stars between main-sequence turn-off and red giant branch are too blue by
+\cite{2014ApJ...783..114G}, their predicted colors for sub-giant stars between main-sequence turn-off and red giant branch are too blue by
 about 0.1--0.2 mag. Second, their metallicity grid does not extend to the $[Fe/H]>0$ range relevant for some disk stars. Finally, they
-correspond to very old populations (older than a few Gyr) and cannot be used for stars younger than about 1 Gyr.
-
-\begin{figure*}[ht!]
-\plotone{plot3HRdiags.png}
-\caption{The left panel shows SDSS-based empirical absolute magnitude vs. color parametrization for main-sequence stars and red giants.
-The symbols are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
-metallicity: $[Fe/H] = -2, -1, 0$ (dot-dashed, solid and dashed, respectively). The middle panel shows a sample of 63,000 stars that have
-signal-to-noise ratio for Gaia’s parallax measurements of at least 20 (white dwarfs can be seen in the lower left corner). The dot-dashed,
-solid and dashed black lines are the same as in the left panel. For comparison, the essentially identical yellow dotted lines were computed
-using eqs. A2 and A7 from Ivezi\'{c} et al. (2008). Note the discrepancy between these parametrizations and data for sub-giant stars ($M_r \sim 3$
-and $g-i \sim 0.8$).
-The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots, and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots.
-Their absolute magnitudes were computed using the so-called “photo-geometric” distances from Bailer-Jones et al.
-(2021).
-The dot-dashed, solid and dashed black lines are the same as in the left and middle panels.
-About 10,000 outliers (about 2.5\% of the full
-sample) seen at $g-i = 0.4$ and $Mr > 7$ are predominantly found at the faint end ($r>20$).}
-\label{fig:3HRdiags}
-\end{figure*}
+correspond to very old populations (older than a few Gyr) and cannot be used for stars younger than about 1-2 Gyr.
 
 We use a combination of SDSS and Gaia data to demonstrate the first problem. The middle panel in Figure~\ref{fig:3HRdiags} 
-shows a clear discrepancy between SDSS-based empirical tracks and data for sub-giant stars. Nevertheless, it is noteworthy that
+shows a clear discrepancy between SDSS-based empirical tracks and data for sub-giant stars, using parallax-based absolute
+magnitudes. Nevertheless, it is noteworthy that
 for main-sequence stars the agreement is excellent. Furthermore, the right panel in Figure~\ref{fig:3HRdiags} demonstrates
 that SDSS-based photometric distances and Gaia-based ``photo-geometric'' distances from \cite{bailer-jones_estimating_2021}
-for main-sequence stars are on the ``same scale''.  XXX provide numerical results for offsets!
+for main-sequence stars are on the ``same scale''.  XXX ZI: provide numerical results for offsets!
 
@@
 
-\subsubsection{MIST/Dartmouth Isochrones for Main-sequence Stars and Red Giants}
+\subsubsection{Model-based Isochrones for Main-sequence Stars and Red Giants}
 
-Look at the code first!
+We considered two sets of isochrones: the Dartmouth Stellar Evolution Database\footnote{See http://stellar.dartmouth.edu/models},
+hereafter DSED \citep{2008ApJS..178...89D}, and PARSEC\footnote{http://stev.oapd.inaf.it/cgi-bin/cmd\_3.7} \citep{2012MNRAS.427..127B} isochrones. 
+Both isochrone sets span adequate ranges of age and metallicity. Unfortunately, the computed color sequences show discrepancies
+with the observed SDSS stellar locus at the level of 0.1-0.2 mag and cannot be used without further adjustments.
 
-We considered two sets of isochrones:
+We ``augment'' the empirical SDSS-based isochrones in two steps:
+\begin{enumerate}
+\item We extend its metallicity range from $[Fe/H]=0$ to $[Fe/H]=+0.5$ by linear extrapolation of the color vs. metallicity
+  dependence around $[Fe/H]=0$. We adjust the gradient by a multiplicative factor (0.7) to ensure that the bright end of the
+  stellar locus at $[Fe/H]=+0.5$ agrees with SDSS-Gaia data (shown in the middle panel in Figure~\ref{fig:3HRdiags}).
+\item 
+  For $M_r < M_{TO}$, where turn-off absolute magnitude $M_{TO}$ depends on age and ranges from $M_{TO}$=4 for age = 1 Gyr
+  to $M_{TO}$ = 5 for age = 10 Gyr, we ``attach'' model-based isochrones (we use DSED isochrones hereafter) to empirical
+  SDSS-based isochrones.
+\end{enumerate}
 
-\begin{figure*}[ht!]
+Examples of resulting color tracks for two representative ages are shown in Figure~\ref{fig:augmLocus}. Sub-giant stars that could not
+be fit with empirical SDSS-based tracks can now be explained with tracks for old stars and intermediate-range metallicity. 
+We have computed such tracks for a grid of ages but believe that the two choices shown in Figure~\ref{fig:augmLocus}
+(1 Gyr and 10 Gyr) should suffice for ``non-specialized'' bulk processing. The reason is that the locii for ages above 1-2 Gyr look
+very similar to each other, while the fraction of stars younger than 1 Gyr is very small at the faint apparent magnitude
+levels probed by SDSS and LSST. Of course, in sky regions with intensive star formation, a ``specialized'' approach with
+a fine age grid can be easily executed.
+
+These color tracks for main-sequence stars and red giants account for an overwhelming majority of stars 
+expected in SDSS and LSST catalogs (approximately $>$95\% but the fraction varies with apparent magnitude and sky position). 
+Nevertheless, we also explicitly account for a few additional populations: white dwarfs, unresolved binary stars and
+blue horizontal branch stars.
+
+\begin{figure*}[b!]
 \plotone{compare2isochrones_augmentedSDSS_1Gyr_10Gyr.png}
-\caption{Augmented locus}
+\caption{Augmented SDSS color tracks for two isochrone ages (left: 1 Gyr; right: 10 Gyr) and the full metallicity range
+  ($-2.5 < [Fe/H] < +0.5$). The black dots show the same sample of 63,000 stars from the middle panel in
+  Figure~\ref{fig:3HRdiags}. As can be seen in the top right panel, sub-giant stars that could not be fit with emprical
+  SDSS-based tracks (see the middle panel in Figure~\ref{fig:3HRdiags}) can be explained with tracks for old stars and
+  intermediate-range metallicity. The sharp feature protruding from the main locus in the middle two panels corresponds
+  to most luminous and evolved high-metallicity stars. 
+}
 \label{fig:augmLocus}
@@
 
-These color tracks account for an overwhelming majority of stars expected in LSST catalogs. Nevertheless,
-
 \subsubsection{Luminosity-Color Tracks for White Dwarfs}
 
+There are two model families of tracks, for H and He white dwarfs, by \cite{1995PASP..107.1047B}.
 Bob Abel will provide text and plots.
+
+Figure~\ref{fig:locusWDs} shows preliminary results.
+
+\begin{figure*}[ht!]
+\plotone{plot3diagsBAzoom.png}
+\caption{Preliminary results for white dwarf luminosity-color
+  tracks. When updating this figure with Bob's improved tracks, add also tracks for BHB stars. }
+\label{fig:locusWDs}
+\end{figure*}
+
+\subsubsection{Luminosity-Color Tracks for Unresolved Binaries}
+
+Following \cite{2004ApJ...615L.141S}, we generate luminosity-color tracks for unresolved binaries
+consisting of a main-sequence star and white dwarf.
+
+Bob Abel will provide text and plots. Figure~\ref{fig:locusBinaries}
+shows preliminary results.
 
@@
 \plotone{plot3diagsBA.png}
-\caption{Augmented locus}
-\label{fig:augmLocus}
+\caption{Preliminary results for unresolved binaries composed of a main-sequence star and white dwarf.}
+\label{fig:locusBinaries}
 \end{figure*}
 
-\subsubsection{Luminosity-Color Tracks for Unresolved Binaries }
+\subsubsection{Luminosity-Color Tracks for Blue Horizontal Branch Stars}
 
-Bob Abel will provide text and plots.
-
-\begin{figure*}[ht!]
-\plotone{plot3diagsBAzoom.png}
-\caption{Augmented locus}
-\label{fig:augmLocus}
-\end{figure*}
+\v{Z}I has data files from the \cite{2004AJ....127..899S} paper and needs to write a few words
+(and add tracks to Figure~\ref{fig:locusWDs}).
 
@@
 
-Reiterate that variable objects, such as quasars and RR Lyrae are not considered here
-but can be addressed in a straighforward manner. Cite Sesar et al. papers
+There will always be sources that cannot be fully explained with any of the available families of
+luminosity-color tracks. For example, quasars and RR Lyrae are not considered here
+but can be easily recognized and removed from the sample in a straighforward manner since
+they are variable \citep[e.g., see][]{2007AJ....134.2236S}. The remaining non-variable sources
+that are not well fit with models can be recognized as ``bad fits'' (see the last subsection on
+model selection below).
 
-Examples: BHB stars, AGB stars...
+There will be impostors, too, with measured colors consistent with available model(s), but
+with a very different nature (e.g., ultra-cold white dwarfs ``hiding'' in the main stellar locus).
+As the time-domain information is built up with the progress of LSST, 
+it might be possible to recognize them, for example, as high proper motion sources or perhaps
+using a broader wavelength range through cross-correlation with other surveys (e.g., WISE or
+Roman surveys). While they will require specialized studies, based on TRILEGAL simulations
+their fraction in the sample should be very small (most likely $<1$\%).
 
@@
 
-Following \cite{2012ApJ...757..166B}
+Given a stellar population and resulting color tracks for intrinsic un-extincted colors, colors need to
+be corrected for interstellar dust extinction using an adequate dust extinction model. Additive color
+corrections $\vec{\delta c}(A_r)$ (see eq.~\ref{eq:reddCorr}) can be computed as 
+\begin{equation}
+ \label{eq:extCorr} 
+                      \delta c = (C_{m2}- C_{m1}) \, A_r,
+\end{equation}
+where $m1$ and $m2$ stand for two bandpasses that define the color (e.g., $u$ and $g$).  
+Dust extinction models, such as \cite{1989ApJ...345..245C} and
+\cite{1999PASP..111...63F}, parametrize extinction coefficients $C_{m}$ as functions of $R_V$, where
+$R_V = A_V/E(B-V)$ and $E(B-V)$ is the stellar ``color excess'' \citep{1989ApJ...345..245C}.
 
-prior from IR dust maps
+Recent studies, such as \cite{2012ApJ...757..166B} based on SDSS data (including the Galactic plane), 
+find that the scatter of $R_V$ around its mean value $R_V=3.1$ is very small. For this reason, we adopt
+empirical results from their Table 1: $C_m$ = (1.810, 1.400, 0.759, 0.561) in $(u, g, i, z)$, respectively ($C_r=1$
+by definition). This choice does not preclude the use of our framework to fit for $R_V$ as the fourth free
+model parameter in Galactic plane regions with very large $A_r$; we  can simply add multiple models
+with extincted colors generated using different values of $C_m$.
 
-Extinction is another major source of systematic errors in the process of luminosity and distance determination.
-The fact that the extinction vector is nearly parallel to the main stellar locus in the two-color diagrams gives rise to degeneracies that complicate the determination of the stellar type.
-An example is displayed in Figure~\ref{fig:extinction}, where in the left panel any of the different star types designated as 1,2 and 3 can have the same observed colors as the star marked as "Obs".
-This degeneracy is a result of the combination of colors chosen for the two-color diagram and depends on the position on the stellar locus and the adopted extinction curve parametrized by a single parameter \textit{R\textsubscript{V}}
+In our implementation, constraints on $A_r$ for two nearby stars, even if they have similar distances,
+are independent. We note that one could use the so-called hierarchical Bayesian modeling  and specify prior
+for the line-of-sight extinction profile: nearby stars would then jointly constrain it. For more
+details, see Green et al. (2014).
 
-\begin{figure}[ht!]
-	\plotone{BYiR7o.png}
-	\caption{Add caption.} \label{fig:extinction}
-\end{figure}
-
-\begin{equation}
-	R_V = \frac{A_V}{E(B-V)},
-\end{equation}
-
-where $A_V$ stands for extinction in \textit{V}-band and \textit{E(B-V)} is the color excess. This relationship can be extended to an arbitrary photometric bandpass $\lambda$:
-
-\begin{equation}
-	A_{\lambda} = C_{\lambda}(R_V)A_r,
-\end{equation}
-
-with $A_r$ designating extinction in \textit{r}-band and $C_{\lambda}(R_V)$ describing the shape of the extinction curve.
-The degeneracy from the left panel in Figure~\ref{fig:extinction} can be broken if several different colors are used, particularly those towards the infrared.
-There the stellar locus is not as kinked and the extinction vector is not parallel to it.
-A possible example is shown in the right-hand panel of the Fig.~\ref{fig:extinction}, where $r-i$ and $i-z$ colors are used, and assuming a fixed extinction law a unique solution for the extinction is possible.
-
-\magcom{Explain the choice of \RV.}
+Finally, it is noteworthy to point out that the dust extinction vector is nearly parallel to the main stellar locus,
+and this fact may cause model parameter degeneracies for certain choices of colors (e.g., see Figure~9 in
+\citealt{2012ApJ...757..166B}). Such degeneracies are broken when multiple colors that span a wide
+wavelength range and extend into near-IR (such as the $i-z$ color) are available. For more details, see
+Section 2.8 in \cite{2012ApJ...757..166B}.
 
@@
 
-Priors bring dependence on sky position and apparent magnitude
+\begin{figure*}[t!]
+\hskip -0.5in
+\includegraphics[width=1.07\textwidth,angle=0]{figures/Stripe82priorMosaic.png}
+\vskip -0.4in  
+\caption{Examples of priors for model parameters, $p(M_r, [Fe/H])$, for main-sequence stars and red giants,
+  for four sky positions (rows, for locations see the first column) and for several brightness levels in the LSST
+  magnitude range (left to right: bright to faint). The integral of each map over $M_r$ and $[Fe/H]$ is unity. 
+  The dashed lines are the same in each panel and are added to guide the eye. 
+}
+\label{fig:priorsAstroLab}
+\end{figure*}
 
-The priors are established by partitioning the TRILEGAL galaxy model \citep{dal_tio_simulating_2022} in healpixels, and each of the pixel in one-magnitude wide bins in apparent magnitude.
-The model colors are then obtained from models based on SDSS results (\magcom{reference}), where the input for the models are the \Mr and the \feh obtained from TRILEGAL.
-Given the assumed extinction curve, these colors are then reddened up to the maximum reddening which is estimated from \cite{schlegel_maps_1998} (SFD98) maps.
-The latter upper limit is used in order to provide a realistic stop condition on the amount of extinction and reduce the processing time.
-This is usually a valid assumption because the SFD98 maps provide \textit{total} extinction along a line of sight.
-In order to account for the eventual underestimation in the SFD98 maps we increase the SFD98 extinctions by 20\%.
+In addition to specifying likelihood $p(\vec{c}|M_r, [Fe/H], A_r)$ for a given model $M$, we need to specify
+prior probability distribution for model parameters, $p(M_r, [Fe/H],A_r)$ (see eq.~\ref{eq:BayesPriorExpand};
+we address the model probability $p(M|I)$ further below). First, we assume that the stellar model parameters
+are unrelated to the distribution of interstellar dust along the line of sight,
+\begin{equation}
+ \label{eq:BayesPrior2}
+               p(M_r, [Fe/H],A_r) = p(M_r, [Fe/H]) \, p(A_r).
+\end{equation}
+We adopt a uniform prior for $p(A_r)$ using the values $A_r^{SFD}$ taken from the \cite{schlegel_maps_1998}
+dust extinction maps. To account for potential map errors, we set the maximum allowed value of $A_r$ as
+\begin{equation}
+          A_r^{max} =  a \, A_r^{SFD} \, + \, b,
+\end{equation}
+and $a=1.3$, $b=0.1$. Therefore, $p(A_r) = 1/A_r^{max}$ for $0 \le A_r \le A_r^{max}$ and $p(A_r) = 0$ for $A_r > A_r^{max}$.
 
-% We extract the priors based on position and apparent magnitude from TRILEGAL (\citet{dal_tio_simulating_2022}).
-Ideally, we would In order to extract the priors (i.e.
-prior maps), we divide TRILEGAL data in healpix bins, and further subdivide them in one-magnitude wide bins in apparent magnitude.
-The latter subdivision is helpful in breaking the degeneracies between the giant and dwarf stars, as intrinsically luminous stars become strongly disfavored at faint magnitudes\footnote{In other words, an apparently faint giant star would imply a very large distance.
-For example a moderately bright giant star with \ensuremath{M_r}=0 mag and \ensuremath{r}=22 mag would imply a distance modulus of 22 mag, or distance of approximately 251 kpc.}.
+The prior distribution of stellar model parameters, $p(M_r, [Fe/H])$, depends on sky position and apparent
+magnitude (here $r$) due to the complex structure of the Milky Way. For example,  \cite{2014ApJ...783..114G}
+generated priors using SDSS-based analytic descriptions of the 3-dimensional stellar distribution in the Milky Way
+\citep{2008ApJ...673..864J} and metallicity distribution for disk and halo components \citep{2008ApJ...684..287I}. 
+Since LSST aims to provide good coverage of the Galactic plane and the Bulge, where SDSS data did not provide strong
+constraints, we opted to utilize recent simulations that implement constraints from a variety of modern surveys (some
+include the Galactic plane and Bulge regions).
 
-%Our method is basically brute-force fitting with some intelligent tricks leveraged to obtain faster execution that will be required for 10B LSST stars.
+\cite{dal_tio_simulating_2022} have generated a mock catalog of the Milky Way stars to LSST depth ($r$=27.5) and over
+the entire LSST survey area. The simulation is based on the TRILEGAL code, incorporates all principal stellar populations,
+includes about 10 billion stars, and the catalog is publicly accessible through the NOIRLab Astro Data
+Lab\footnote{Available at https://datalab.noirlab.edu/}.
+We have developed code to query this mock LSST catalog using Astro Data Lab portal. For a given position on the sky, we
+extract all catalog entries from an area of $\sim$10 deg$^2$ (the size of the LSST Camera's field of view) around it, then bin
+the sample using apparent $r$-band magnitude (27 bin centers from $r=14$ to $r=27$, with a bin width of 1 mag). 
+Given an $r$-band selected subsample, we separate all modeled populations and then bin them using adequate model
+parameters (e.g., $M_r$ and $[Fe/H]$ in case of main-sequence stars and red giants). The resulting 2-dimensional map
+is renormalized so that its integral over all model parameters is unity.
 
-\subsection{Bayesian Model Selection}
+Figure~\ref{fig:priorsAstroLab} shows examples of such $p(M_r, [Fe/H])$ maps for a grid of apparent magnitudes and for
+several characteristic sky positions. Several features are noticeable: i) there are two ``clouds'', at low and high metallicity,
+that correspond to halo and disk stars; ii) the distribution towards the Galactic center (top row) is much more compact in
+the $M_r$ direction (because these subsamples are dominated by the Bulge stars at similar distances); iii) the whole 
+distribution shifts to fainter levels as the bin center ($r$) becomes fainter. We have experimented with different region
+sizes\footnote{We use HEALPix geometry, see https://healpix.jpl.nasa.gov.} and found that $\sim$10 deg$^2$
+(HEALPix $N_{side}$=16 and 32, with 3,072 and 12,288 pixels over the full sky) is sufficiently small to capture the
+variation of $p(M_r, [Fe/H])$ maps across the sky.
 
-\subsection{Implementation: Photo-D pipeline}
+\subsection{An Example of Bayesian Model Parameter Estimation}
 
-optimized grid search, MCMC, refer to NN paper, also in Discussion
-
-Our fitting procedure is also executed on an adaptive grid, a coarse search over the parameter space is performed first in order to establish the layout of the manifold.
-However, care is taken that any possible local minima are not missed by appropriately adjusting the step size \magcom{how?}.
-The located maxima are then explored with a smaller step size (\magcom{adjusted how?}).
-
-In addition to the approach described here, we also tested Markov Chain Monte Carlo and neural network approaches that will be/are described in forthcoming/published papers.
-
-%Under the assumptions that the locus of the color-color diagram is representative of the stellar SEDs, that the shape of the extinction curve can be parametrized with a single parameter \RV
-
-%%% METHODOLOGY
-
-% 2) statistics:
-
-\begin{figure*}[ht!]
-\plotone{bayesPanels_ex1.png}
-\caption{Write caption: prior, likelihood, posterior maps}
+\begin{figure*}[t!]
+\hskip -0.7in
+\includegraphics[width=1.2\textwidth,angle=0]{figures/bayesPanels_ex1.png}
+\caption{Maps of the prior (left), likelihood (middle; computed using eq.~\ref{eq:lnLcolors}) and posterior (right;
+computed using eq.~\ref{eq:BayesFull}) for a simulated main-sequence star with $r=21.2$, $(g-r)=0.48$,
+$M_r= 4.58$, $[Fe/H]=-0.76$ and $A_r=0.37$. The prior in the $A_r$ direction is uniform. The circles mark
+the values of input $M_r$ and $[Fe/H]$. The posterior marginal distributions for all three model parameters
+are shown in Figure~\ref{fig:cornerPlot3ex1}.
+}
 \label{fig:bayesPanels}
 \end{figure*}
+
+We use isochrones shown in the right panels of Figure~\ref{fig:augmLocus} and priors shown
+in the second row of Figure~\ref{fig:priorsAstroLab} to illustrate the Bayesian model parameter estimation. 
+Figure~\ref{fig:bayesPanels} shows prior, likelihood and posterior for a main-sequence star close to the
+turn-off point. Note how the posterior, although much wider than the likelihood map, helps break the
+degeneracy\footnote{Priors can break degeneracies between the giant and dwarf stars because luminous
+stars become strongly disfavored at faint magnitudes (because an apparently faint giant star would imply
+a very large distance, beyond the presumed edge of the Galaxy at $\sim$100 kpc; for example a giant star
+with $M_r$=0 and $r$=22 would imply a distance of $\sim$250 kpc).} in the likelihood map (the two ``islands'').
+
+\begin{figure}[t!]
+  \includegraphics[width=1.0\textwidth,angle=0]{figures/cornerPlot3_ex1.png}
+  \vskip -0.5in
+\caption{Two-parameter covariances and marginal distributions for
+  the posterior map from Figure~\ref{fig:bayesPanels}. The symbols and
+  dashed lines show true input values.}
+\label{fig:cornerPlot3ex1}
+\end{figure}
+
+Figure~\ref{fig:cornerPlot3ex1} shows two-parameter covariances and marginal distributions for this case.
+Note that true values are recovered within expected uncertainties, as well as non-vanishing covariances
+between the parameters. The marginal distributions produced with the prior, likelihood and posterior maps,
+shown in
+% Figure~\ref{fig:margPost3D},   % FOR SOME REASON, THIS REFERENCE DOESN'T WORK
+Figure 9,
+illustrate the improvement in the ``knowledge'' of model
+parameters between prior and posterior, brought by color measurements via the likelihood map.
 
@@
 
-\begin{figure*}[ht!]
-\plotone{cornerPlot3_ex1.png}
-\caption{Write caption: 2-param covariances and marginal distributions }
-\label{fig:cornerPlot3}
-\end{figure*}
+\begin{figure}[t!]
+%\plotone{margPosteriors3D_ex1.png}
+\includegraphics[width=0.99\textwidth,angle=0]{figures/margPosteriors3D_ex1.png}
+\caption{An illustration of the improvement in the ``knowledge'' of model parameters between prior and posterior,
+brought by measurements via the likelihood map. The lines show marginal distributions produced with the
+prior (blue), likelihood (green) and posterior (red) maps from Figure~\ref{fig:bayesPanels}. The dashed line shows the
+expectation value for the marginalized posterior and the vertical solid line marks the true input value.}
+\label{fig:margPost3D}
+\end{figure}
 
-\begin{figure*}[ht!]
-\plotone{margPosteriors3D_ex1.png}
-\caption{Write caption: prior, likelihood, posterior marginal distributions}
-\label{fig:margPosteriors3D}
-\end{figure*}
+In the next Section, we discuss a fast numerical pipeline implementation that can perform this computation for LSST-sized
+catalogs and provide a more quantitative analysis of the method's performance.  We conclude this Section with a brief
+discussion of Bayesian model selection method for assigning posterior probabilities to each stellar population used to interpret
+observations.
 
-% 3) code testing:
+\subsection{Bayesian Model Selection}
 
-\begin{figure*}[ht!]
-\plotone{qpBmeans_chiTest4_3D.png}
-\caption{Write caption: TRILEGAL mean values of input model params in umag vs. g-i }
-\label{fig:qpBmeans_chiTest4_3D.}
-\end{figure*}
+For each star and for each stellar population (that is, a model for color-magnitude tracks), a posterior
+3-dimensional data cube is produced.  These posteriors are used when choosing the best model, as follows.
 
-\begin{figure*}[ht!]
-\plotone{qpBcmd_chiTest4_3D_Mr.png}
-\caption{Write caption: performance for Mr vs. true Mr and FeH}
-\label{fig:qpBcmd_chiTest4_3D_Mr}
-\end{figure*}
+Bayes' theorem as introduced by eq.~\ref{eq:BayesFull} quantifies the posterior
+pdf of parameters describing a single model, with that model {\it assumed to be true}.
+To find out which of two models, say $M_1$ and $M_2$, is better supported by data, we
+compare their posterior probabilities via the {\it odds ratio} in favor of model $M_2$ over
+model $M_1$ as 
+\begin{equation}
+                  O_{21} \equiv { p(M_2|D,I) \over p(M_1|D,I) }.
+\end{equation}
 
-\begin{figure*}[ht!]
-\plotone{qpBcmd_chiTest4_3D_FeH.png}
-\caption{Write caption: performance for FeH vs. true Mr and FeH}
-\label{fig:qpBcmd_chiTest4_3D_FeH}
-\end{figure*}
+The posterior probability for model $M$ ($M_1$ or $M_2$) given data $D$, $p(M|D,I)$  in this
+expression, can be obtained from the posterior pdf $p(M,\vec{\theta}|D,I)$ in eq.~\ref{eq:BayesFull}
+using marginalization (integration) over the model parameter space spanned by $\vec{\theta}$.
+The posterior probability that the model $M$ is correct given data $D$ (a number between 0~and~1)
+can be derived using eqs.~\ref{eq:BayesFull} and \ref{eq:BayesPriorExpand} as
+\begin{equation}
+             p(M|D,I) =  {p(D|M,I) \, p(M|I) \over p(D|I)},
+\end{equation}           
+where
+\begin{equation}
+\label{eq:evidence}
+       E(M) \equiv p(D|M,I) = \int p(D|M,\vec{\theta},I) \, p(\vec{\theta}|M,I) \, d \vec{\theta}
+\end{equation}
 
-\begin{figure*}[ht!]
-\plotone{qpBcmd_chiTest4_3D_Ar.png}
-\caption{Write caption: performance for Ar vs. true Mr and FeH}
-\label{fig:qpBcmd_chiTest4_3D_Ar}
-\end{figure*}
+is called the {\it marginal likelihood} (or {\it the evidence}) for model $M$ and it quantifies the 
+probability that the data $D$ would be observed {\it if} the model $M$ were the correct model.
+Since the marginal likelihood $E(M)$ involves integration of the data likelihood $p(D|M,\vec{\theta},I)$,
+it is also called the {\it global likelihood} for model $M$.
 
-\begin{figure*}[ht!]
-\plotone{qpB_chiTest4_3D_Mr.png}
-\caption{Write caption: performance for Mr vs. estimated Mr and FeH}
-\label{fig:qpB_chiTest4_3D_Mr}
-\end{figure*}
+The global likelihood is a {\it weighted average} of the likelihood function, with the prior for model parameters
+acting as the weighting function. Alternatively, $E(M)$ is simply the integral over allowed parameter space of
+the posterior pdf before its renormalization to set this integral to unity (e.g., the integral of the posterior shown
+in the right panel in Figure~\ref{fig:bayesPanels}). If the chosen model color tracks cannot explain the observed
+colors, the likelihood (eq.~\ref{eq:lnLcolors}) will never be very high and the resulting $E(M)$ will be low.
+We note that in the limit of Gaussian posterior and flat priors, Bayesian evidence-based model selection becomes
+equivalent to $\chi^2$ selection from the frequentist statistical framework.
 
-% *** data-based comparisons ***
-% Stripe 82 vs. Gaia
-% Bulge BDSD sample?
+The probability of data, $p(D|I)$, cancels out when the odds ratio is considered: 
+\begin{equation}
+       O_{21} =  {E(M_2) \, p(M_2|I) \over E(M_1) \, p(M_1|I)} =
+       B_{21} \, {p(M_2|I) \over p(M_1|I)}. 
+\end{equation}
+In practice, the values of the odds ratio are interpreted using Jeffreys' scale; in particular, $O_{21}>10$ represents
+``strong'' evidence in favor of $M_2$ ($M_2$ is ten times more probable than $M_1$).
+
+The ratio of global likelihoods, $B_{21}\equiv E(M_2)/E(M_1)$, is called the {\it Bayes factor}, and
+is equal to 
+\begin{equation}
+     B_{21} = {\int p(D|M_2,\vec{\theta}_2,I) \, p(\vec{\theta}_2|M_2,I) \, d \vec{\theta}_2 \over
+                    \int p(D|M_1,\vec{\theta}_1,I) \, p(\vec{\theta}_1|M_1,I) \, d \vec{\theta}_1}.
+\end{equation}
+The vectors of parameters, $\vec{\theta_1}$ and $\vec{\theta_2}$, are explicitly indexed to emphasize
+that the two models may span vastly different parameter spaces (including the  number of
+parameters per model).
+
+The prior model probabilities, $p(M_1|I)$ and $p(M_2|I)$, are determined using TRILEGAL simulated catalog.
+For example, if model 1 is main-sequence stars and model 2 is white dwarfs, we estimate the $p(M_2|I)/p(M_1|I)$
+ratio by simply counting main-sequence stars and white dwarfs in the corresponding $r$ magnitude bin.
+
+In case of $N_M$ models, we assume that they represent an exhaustive model set and estimate the posterior
+probability of model $k$, $k=1...N_M$, as
+\begin{equation}
+                 p(k) = { O_{k1}  \over \sum_{j=1}^{j=N_M}  O_{kj}},
+\end{equation}
+where the model $k=1$ is chosen arbitrarily without a loss of generality ($O_{11}$ = 1). We will illustrate model
+selection with multiple stellar populations in the next Section.

### 2024-01-26 paper/methodology.tex
Minor additions to the paper.

@@
 % Figure~\ref{fig:margPost3D},   % FOR SOME REASON, THIS REFERENCE DOESN'T WORK
-Figure 9,
-illustrate the improvement in the ``knowledge'' of model
+%Figure 9,
+Figure~\ref{fig:margPost3D} illustrate the improvement in the ``knowledge'' of model
 parameters between prior and posterior, brought by color measurements via the likelihood map.

### 2024-01-26 paper/paper.tex
first draft of Section 2
Minor additions to the paper.

@@
-\documentclass{aastex63} 
+\documentclass{aastex631} 
 % \documentclass[12pt,preprint]{aastex}
@@
 
-\author[0000-0000-0000-0000]{Tomislav Jurki\'{c}}
+\author[0000-0002-4993-2939]{Tomislav Jurki\'{c}}
 \affiliation{Faculty of Physics, University of Rijeka, Radmile Matej\v{c}i\'{c} 2, 51000 Rijeka, Croatia}
@@
 
-\acknowledgements
-\v{Z}.I. acknowledges funding by the Fulbright Foundation and thanks the Ru\d er Bo\v{s}kovi\'{c} Institute for hospitality.
+\begin{acknowledgements}
+
+This work is financed within the Tenure Track Pilot Programme of the Croatian Science Foundation and the Ecole Polytechnique Fédérale de Lausanne and the Project TTP-2018-07-1171 Mining the variable sky, with the funds of the Croatian-Swiss Research Programme.
+
+\v{Z}.I. acknowledges funding by the Fulbright Foundation and thanks the Ru\dj er Bo\v{s}kovi\'{c} Institute for hospitality.
 
@@
 
+\end{acknowledgements}
+
 \vspace{5mm}

### 2024-01-30 paper/methodology.tex
Paper  comments.

@@
 
-Therefore, $\vec{D}$ = ($r$, $\vec{c}$) and $\vec{\theta}$ = ($M_r$, $[Fe/H]$, $A_r$). Data $\vec{D}$ and
+\redcom{LP: "In summary", instead of "Therefore", or am I missing something?
+Equations could be moved to the top of the paragraph, this is somewhat repetitive.} Therefore, $\vec{D}$ = ($r$, $\vec{c}$) and $\vec{\theta}$ = ($M_r$, $[Fe/H]$, $A_r$).
+Data $\vec{D}$ and
 model parameters $\vec{\theta}$ are related via the Bayes theorem (see, e.g., Chapter 5 in \citealt{2020sdmm.book.....I}),
@@
 is simply renormalized so that
-its integral over all model parameters $\vec{\theta}$ is unity. The integral of the prior $p(\vec{\theta}|M,I)$
+its integral over all model parameters $\vec{\theta}$ is unity. \redcom{LP: A zašto?} The integral of the prior $p(\vec{\theta}|M,I)$
 over all parameters should also be unity, but for the same reason, calculations of the posterior pdf are
 often done with an arbitrary normalization. An important exception is model selection discussed further
-below, where the correct normalization of the product $p(D|M,\vec{\theta},I) \, p(\vec{\theta}|M,I)$ is crucial.
+below, where the correct normalization of the product $p(D|M,\vec{\theta},I) \, p(\vec{\theta}|M,I)$ is crucial. \redcom{LP: OK, ovo mi baš nije bilo jasno...}
 
@@
 \caption{An illustration of multiple stellar populations. The red dots in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
-(variable stars are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
+(variable stars (\redcom{LP: are QSO excluded too?
+If yes then "stars" should be replaced with "sources".}) are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
 A subset of 415,000 stars with $r < 22$ and $u<22$ are overplotted as blue dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ (dominated by main-sequence stars and red giants) are overplotted as cyan dots.
@@
 \caption{The left panel shows SDSS-based empirical absolute magnitude vs. color parametrization for main-sequence stars and red giants.
-The symbols are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
+The symbols \redcom{"ribbons"?} are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
 metallicity: $[Fe/H] = -2, -1, 0$ (dot-dashed, solid and dashed, respectively). The middle panel shows a sample of 63,000 stars that have
 signal-to-noise ratio for Gaia’s parallax measurements of at least 20 (white dwarfs can be seen in the lower left corner; same sample is
-shown by green dots in Figure~\ref{fig:3HRdiags}). Their absolute magnitude are derived from parallax measurements. The dot-dashed,
+shown by green dots in Figure~\ref{fig:3HRdiags}). Their absolute magnitudes are derived from parallax measurements. The dot-dashed,
 solid and dashed black lines are the same as in the left panel. For comparison, the essentially identical yellow dotted lines were computed
-using eqs. A2 and A7 from Ivezi\'{c} et al. (2008). Note the discrepancy between these parametrizations and data for sub-giant stars ($M_r \sim 3-4$
+using eqs. A2 and A7 from Ivezi\'{c} et al. (2008). \redcom{very difficult to discern...} Note the discrepancy between these parametrizations and data for sub-giant stars ($M_r \sim 3-4$
 and $g-i \sim 0.8-1.1$).
-The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots (shown by blue dots in Figure~\ref{fig:3HRdiags}), and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots.
+The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots (shown by blue dots in Figure~\ref{fig:3dataDiags}(\redcom{corrected to Fig 1; was this the intention?}), and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots.
 Their absolute magnitudes were computed
@@
 same as in the left and middle panels. About 10,000 outliers (about 2.5\% of the full
-sample) seen at $g-i = 0.4$ and $Mr > 7$ are predominantly found at the faint end ($r>20$).}
+sample) seen at $g-i = 0.4$ and $Mr > 7$ are predominantly found at the faint end ($r>20$).(\redcom{There is some interesting structure in the right plot...but I don't get exactly što je pjesnik želio reći s ovim plotom?})}
 \label{fig:3HRdiags}
@@
 We address all three problems by switching from the SDSS-based empirical isochrones that correspond to old
-populations to model-based isochrones that span a range of ages and a wider range of metallicities.
+populations to model-based isochrones that span a range of ages and a wider range of metallicities \redcom{where the former fail to properly explain the observations?}
 
@@
 
+\redcom{Tu sam bio malo zbunjen: u zadnjem odlomku iz 2.2.1. piše: da se prebacimo na model-based izokrone; čitanjem 2.2.2. ispada da su modeli samo nadoštukani? }
+
 \begin{figure*}[b!]
@@
   to most luminous and evolved high-metallicity stars. 
-}
+(\redcom{Why is the row 3 funky?})}
 \label{fig:augmLocus}
@@
 with extincted colors generated using different values of $C_m$.
+\redcom{If we are constraining $A_r$ from data that are fitted to the models, what does this loop look like...?
+For each LoS we fit different models and stop when some convergence criteria is satisfied?
+Is this better than taking a prior from TRILEGAL, or at the least using TRILEGAL to constrain the possible models?
+OK, only a small area is possibly affected by this but IIRC,in the central regions there are clumps so the results may look too "smooth".}
 
@@
 \end{equation}
-and $a=1.3$, $b=0.1$. Therefore, $p(A_r) = 1/A_r^{max}$ for $0 \le A_r \le A_r^{max}$ and $p(A_r) = 0$ for $A_r > A_r^{max}$.
+and $a=1.3$, $b=0.1$. Therefore, $p(A_r) = 1/A_r^{max}$ for $0 \le A_r \le A_r^{max}$ and $p(A_r) = 0$ for $A_r > A_r^{max}$. \redcom{Why those exact numbers?}
 

### 2024-03-07 paper/notebooks/paperPlots.py
added WD and BHB models

@@ -6,7 +6,7 @@ from scipy.stats import norm
 from astroML.stats import binned_statistic_2d
 
 
-def plot3diagsBobAbel(df1, df2, df3, df4, L0, L1, L2, WD, WDMD, zoom=False):
+def plot3diagsBobAbel(df1, df2, df3, df4, L0, L1, L2, WD, WDMD, BHB, zoom=False):
 
     fig=plt.figure(2,figsize=(12,4))
     fig.subplots_adjust(wspace=0.2,hspace=0.2,top=0.97,bottom=0.1,left=0.09,right=0.95)
@@ -35,9 +35,11 @@ def plot3diagsBobAbel(df1, df2, df3, df4, L0, L1, L2, WD, WDMD, zoom=False):
     ax=fig.add_subplot(131)
     ax.scatter(df4['gi'], df4['Mr'], s=3.5, c='green') 
     if zoom:
-        ax.text(-1.8, 14, 'DA: yellow')
-        ax.text(-1.8, 14.8, 'DB: red')
-        ax.text(-1.8, 15.6, 'DC: cyan')
+        #ax.text(-1.8, 14, 'DA: yellow')
+        #ax.text(-1.8, 14.8, 'DB: red')
+        #ax.text(-1.8, 15.6, 'DC: cyan')
+        ax.text(-1.8, 14, 'H: yellow')
+        ax.text(-1.8, 14.8, 'He: cyan') 
     if (1):
         giGrid = np.linspace(0.2, 4.0, 300)
         Mr0 = getMr(giGrid, 0.0)
@@ -47,9 +49,9 @@ def plot3diagsBobAbel(df1, df2, df3, df4, L0, L1, L2, WD, WDMD, zoom=False):
         ax.plot(giGrid, Mr1, ls='dotted', c='yellow')
         ax.plot(giGrid, Mr2, ls='dotted', c='yellow')
     ax.set_xlim(-1.8,4.1)
-    ax.set_ylim(16.1, 1.0)
+    ax.set_ylim(16.1, 0.0)
     if zoom:
-        ax.set_xlim(-2.0,1.0)
+        ax.set_xlim(-2.0, 1.0)
         ax.set_ylim(16.1, 8.0)
        
     ax.set_xlabel('g-i')
@@ -62,7 +64,8 @@ def plot3diagsBobAbel(df1, df2, df3, df4, L0, L1, L2, WD, WDMD, zoom=False):
[…]

### 2024-03-07 paper/intro.tex
Paper  comments.
new sections 2.2.3 and 2.2.4, and comments from Lovro and Bob

@@
 structure seen in the Hertzsprung-Russell diagram. In addition to colors and luminosity, these correlations
-also involve metallicity in case of main-sequence and red giant stars, and surface gravity in case of white dwarfs.
+also involve metallicity in the case of main-sequence and red giant stars, and surface gravity in the case of white dwarfs.
 For the youngest main-sequence stars, stellar age may play a role, too. Measured stellar colors are also affected

### 2024-03-07 paper/methodology.tex
new sections 2.2.3 and 2.2.4, and comments from Lovro and Bob

--- 2024-01-25 paper/methodology.tex
@@
 Data, or observations, include multi-band photometry that is used
-to construct colors, $\vec{c}$. In case of SDSS, colors include $u-g$, $g-r$, $r-i$, $i-z$, and in case of LSST also
+to construct colors, $\vec{c}$. In the case of SDSS, colors include $u-g$, $g-r$, $r-i$, $i-z$, and in the case of LSST also
 $z-y$. In addition to colors, $\vec{D}$ also includes an apparent magnitude and hereafter we choose the $r$-band
@@
 ($\sim$1 \%) include absolute magnitude (here chosen in the $r$ band, $M_r$), metallicity ($[Fe/H]$) and surface
-gravity. In case of main-sequence stars and red giants, the color tracks can be expressed as functions of $M_r$
+gravity. In the case of main-sequence stars and red giants, the color tracks can be expressed as functions of $M_r$
 and $[Fe/H]$ along an isochrone, without having to explicitly specify surface gravity. With other populations,
@@
 $A_r$ is extinction in the $r$ band and extinction in other bands is proportional to $A_r$, with known constants
-of proportionality (this assumption can be relaxed, see below). Once $M_r$ and $A_r$ are constrained, distance
+of proportionality (this assumption can be relaxed, see below). Once $M_r$ and $A_r$ are constrained, the distance
 modulus $\Delta$ can be computed from
@@
 Note that in practice the uncertainty of the observed magnitude, $r^{obs}$, is always significantly smaller than
-the width of $p(M_r + A_r)$ (at the bright end by an order of magnitude, $\sim$0.01 mag vs. $\sim$0.1 mag).
+the width of $p(M_r + A_r)$ distribution (at the bright end by an order of magnitude, $\sim$0.01 mag vs. $\sim$0.1 mag).
 
-Therefore, $\vec{D}$ = ($r$, $\vec{c}$) and $\vec{\theta}$ = ($M_r$, $[Fe/H]$, $A_r$). Data $\vec{D}$ and
-model parameters $\vec{\theta}$ are related via the Bayes theorem (see, e.g., Chapter 5 in \citealt{2020sdmm.book.....I}),
+In summary, $\vec{D}$ = ($r$, $\vec{c}$) and $\vec{\theta}$ = ($M_r$, $[Fe/H]$, $A_r$).
+Data $\vec{D}$ and model parameters $\vec{\theta}$ are related via the Bayes theorem (see, e.g., Chapter 5 in \citealt{2020sdmm.book.....I}),
 \begin{equation}
@@
 The term $p(D|I)$ is the {\it probability of data}, or the prior predictive probability for $D$. 
-It provides proper normalization for the posterior  pdf and usually it is not explicitly computed
+It provides proper normalization for the posterior  pdf; for simplicity, it is usually not explicitly computed
 when estimating model parameters: rather, $p(M,\vec{\theta}|D,I)$ for a given $M$
-is simply renormalized so that
+is renormalized so that
 its integral over all model parameters $\vec{\theta}$ is unity.  The integral of the prior $p(\vec{\theta}|M,I)$
@@
 often done with an arbitrary normalization. An important exception is model selection discussed further
-below, where the correct normalization of the product $p(D|M,\vec{\theta},I) \, p(\vec{\theta}|M,I)$ is crucial.
+below in \S\ref{sec:ModSel}, where the correct normalization of the product $p(D|M,\vec{\theta},I) \, p(\vec{\theta}|M,I)$ is crucial.
 
@@
 \caption{An illustration of multiple stellar populations. The red dots in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
-(variable stars are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
+(variable sources are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
 A subset of 415,000 stars with $r < 22$ and $u<22$ are overplotted as blue dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ (dominated by main-sequence stars and red giants) are overplotted as cyan dots.
@@
 by main-sequence stars. As a result, a simple functional relationship, $M_r = f(g-i)$, or its generalized version that
-accounts for the shift of $M_r$ as a function of metallicity, \citep{2008ApJ...684..287I} can be used to estimate distance
+accounts for the shift of $M_r$ as a function of metallicity \citep{2008ApJ...684..287I}, can be used to estimate distance
 in a straightforward manner.}
@@
 
-Given chosen model (i.e., a stellar population) $M$, the likelihood $p(D|M,\vec{\theta},I)$ can be explicitly written as
+Given a chosen model (i.e., a stellar population) $M$, the likelihood $p(D|M,\vec{\theta},I)$ can be explicitly written as
 \begin{equation}
@@
 \end{equation}
-where summation is over all colors (for example, $N=4$ for SDSS, and $N=5$ for LSST), \ensuremath{c^{obs}_i} are the
+where the summation is over all colors (for example, $N=4$ for SDSS, and $N=5$ for LSST), \ensuremath{c^{obs}_i} are the
 observed colors and \ensuremath{c^{mod}_i} are the model colors (they are functions of $M_r$, $[Fe/H]$, and $A_r$
@@
 The likelihood function can be thought of as a  3-dimensional data cube that, for a given set of observed colors $c^{obs}_i$,
-maps the triplet ($M_r, [Fe/H], A_r$) to 1-dimensional scalar, that is, $ln (\mathcal{L})$ is a scalar function of
+maps the triplet ($M_r, [Fe/H], A_r$) to a 1-dimensional scalar, that is, $ln (\mathcal{L})$ is a scalar function of
 $M_r$, $[Fe/H]$, and $A_r$.
 
-The existance of multiple stellar populations in an SDSS photometric catalog is illustrated in Figure~\ref{fig:3dataDiags}.
+The existence of multiple stellar populations in an SDSS photometric catalog is illustrated in Figure~\ref{fig:3dataDiags}.
 With accurate multi-band photometry that includes a UV band (here SDSS $u$), main-sequence stars, red giants,
@@
 \caption{The left panel shows SDSS-based empirical absolute magnitude vs. color parametrization for main-sequence stars and red giants.
-The symbols are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
+The date are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
 metallicity: $[Fe/H] = -2, -1, 0$ (dot-dashed, solid and dashed, respectively). The middle panel shows a sample of 63,000 stars that have
-signal-to-noise ratio for Gaia’s parallax measurements of at least 20 (white dwarfs can be seen in the lower left corner; same sample is
-shown by green dots in Figure~\ref{fig:3HRdiags}). Their absolute magnitude are derived from parallax measurements. The dot-dashed,
+signal-to-noise ratio for Gaia’s parallax measurements of at least 20 (white dwarfs can be seen in the lower left corner.
+Their absolute magnitudes are derived from parallax measurements. The dot-dashed,
 solid and dashed black lines are the same as in the left panel. For comparison, the essentially identical yellow dotted lines were computed
 using eqs. A2 and A7 from Ivezi\'{c} et al. (2008).  Note the discrepancy between these parametrizations and data for sub-giant stars ($M_r \sim 3-4$
-and $g-i \sim 0.8-1.1$).
-The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots (shown by blue dots in Figure~\ref{fig:3HRdiags}), and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots.
-Their absolute magnitudes were computed
-using the so-called “photo-geometric” distances from Bailer-Jones et al. (2021). The dot-dashed, solid and dashed black lines are the
+and $g-i \sim 0.8-1.1$). The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots (shown by blue dots in
+Figure~\ref{fig:3dataDiags}, and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots. Their absolute magnitudes were computed
+using the so-called “photogeometric” distances from Bailer-Jones et al. (2021). The dot-dashed, solid and dashed black lines are the
 same as in the left and middle panels. About 10,000 outliers (about 2.5\% of the full
@@
 for main-sequence stars the agreement is excellent. Furthermore, the right panel in Figure~\ref{fig:3HRdiags} demonstrates
-that SDSS-based photometric distances and Gaia-based ``photo-geometric'' distances from \cite{bailer-jones_estimating_2021}
+that SDSS-based photometric distances and Gaia-based ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}
 for main-sequence stars are on the ``same scale''.  XXX ZI: provide numerical results for offsets!
 
-We address all three problems by switching from the SDSS-based empirical isochrones that correspond to old
-populations to model-based isochrones that span a range of ages and a wider range of metallicities.
+We address all three problems by augmenting the SDSS-based empirical isochrones that correspond to old
+populations with model-based isochrones that span a range of ages and a wider range of metallicities, which
+results in better agreement with the observations.
 
 \subsubsection{Model-based Isochrones for Main-sequence Stars and Red Giants}
+
+\begin{figure*}[ht!]
+\plotone{compare2isochrones_augmentedSDSS_1Gyr_10Gyr.png}
+\caption{Augmented SDSS color tracks for two isochrone ages (left: 1 Gyr; right: 10 Gyr) and the full metallicity range
+  ($-2.5 < [Fe/H] < +0.5$). The black dots show the same sample of 63,000 stars from the middle panel in
+  Figure~\ref{fig:3HRdiags}. As can be seen in the top right panel, sub-giant stars that could not be fit with empirical
+  SDSS-based tracks (see the middle panel in Figure~\ref{fig:3HRdiags}) can be explained with tracks for old stars and
+  intermediate-range metallicity. The sharp feature protruding from the main locus in the middle two panels corresponds
+  to most luminous and evolved high-metallicity stars. Note that diagrams in the bottom row, which do not include
+  the $u$ band, show very little dependence on metallicity.}  
+\label{fig:augmLocus}
+\end{figure*}
 
@@
 
-Examples of resulting color tracks for two representative ages are shown in Figure~\ref{fig:augmLocus}. Sub-giant stars that could not
+Examples of the resulting color tracks for two representative ages are shown in Figure~\ref{fig:augmLocus}. Sub-giant stars that could not
 be fit with empirical SDSS-based tracks can now be explained with tracks for old stars and intermediate-range metallicity. 
@@
 
-These color tracks for main-sequence stars and red giants account for an overwhelming majority of stars 
+These color tracks for main-sequence stars,  blue horizontal branch stars, and red giants account for an overwhelming majority of stars 
 expected in SDSS and LSST catalogs (approximately $>$95\% but the fraction varies with apparent magnitude and sky position). 
-Nevertheless, we also explicitly account for a few additional populations: white dwarfs, unresolved binary stars and
-blue horizontal branch stars.
-
-\begin{figure*}[b!]
-\plotone{compare2isochrones_augmentedSDSS_1Gyr_10Gyr.png}
-\caption{Augmented SDSS color tracks for two isochrone ages (left: 1 Gyr; right: 10 Gyr) and the full metallicity range
-  ($-2.5 < [Fe/H] < +0.5$). The black dots show the same sample of 63,000 stars from the middle panel in
-  Figure~\ref{fig:3HRdiags}. As can be seen in the top right panel, sub-giant stars that could not be fit with emprical
-  SDSS-based tracks (see the middle panel in Figure~\ref{fig:3HRdiags}) can be explained with tracks for old stars and
-  intermediate-range metallicity. The sharp feature protruding from the main locus in the middle two panels corresponds
-  to most luminous and evolved high-metallicity stars. 
-}
-\label{fig:augmLocus}
-\end{figure*}
+Nevertheless, we also explicitly account for a few additional populations: white dwarfs and unresolved binary stars.
 
@@
 
-There are two model families of tracks, for H and He white dwarfs, by \cite{1995PASP..107.1047B}.
-Bob Abel will provide text and plots.
+High-precision SDSS photometry clearly shows two white dwarf (WD) sequences in the $g-r$ vs. $u-g$ color-color diagram
+(see, e.g., Figures 23 and 24 in \citealt{2007AJ....134..973I}, as well as the middle panel in Figure~\ref{fig:3dataDiags}).
+A comparison with models from \cite{1995PASP..107.1047B}, as well as with SDSS spectra, reveals that the two sequences
+correspond to H and He white dwarfs, with the mean log(g) = 8.0 for H sequence and log(g) = 8.5 for He sequence.
+Upper limits for the scatter of log(g) around these mean sequences appears as no more than 0.5. 
+We use three modern white dwarf catalogs: the Montreal White Dwarf
+Database\footnote{https://www.montrealwhitedwarfdatabase.org} \citep{2017ASPC..509....3D}, the Gaia EDR3 White Dwarf Catalog\footnote{https://warwick.ac.uk/fac/sci/physics/research/astro/research/catalogues/} \citep{2021MNRAS.508.3877G},
+and the Gaia-based White Dwarf Database\footnote{https://cdsarc.cds.unistra.fr/viz-bin/cat/J/A+A/679/A127} \citep{2023A&A...679A.127G}
+to validate the \cite{1995PASP..107.1047B} models and derive small color offsets that bring models in perfect agreement with these three datasets.
 
-Figure~\ref{fig:locusWDs} shows preliminary results.
+We consider only white dwarfs with cataloged DA, DB or DC spectral class, at galactic latitudes further than 10 degrees from the plane,
+with SDSS photometry, apparent magnitudes $5 \le m \le 22$ in any band $m$, and $15 \le r \le 19$, where $r$ is the SDSS $r$-band magnitude.
+Observed magnitudes are corrected for interstellar dust extinction using maps from \cite{schlegel_maps_1998} and per-band extinction coefficients
+discussed in \S\ref{sec:ismdust} below. We group DB and DC spectral classes as He dwarfs (1,939 objects), while DA spectral class corresponds
+to H white dwarfs (9,307 objects). Their absolute magnitudes were calculated using ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}.
+
+For each WD type (H and He) and SDSS color, we bin the data into 20 $M_r$ bins and compute the median value of a given color in each bin.
+The color vs. $M_r$ sequences from \cite{1995PASP..107.1047B} are then slightly (up to 0.1 mag) shifted in color so that the mean offset for all bins
+vanishes. The only case where a small (up to 0.1 mag) linear adjustment of the model track was needed is the $u-g$ color for H models at $g-r<-0.3$. 
+After these color adjustments are applied, all color vs. $M_r$ model sequences were linearly interpolated to a common $M_r$ grid ($8.5 < M_r < 14.5$,
+with a step of 0.02 mag). The resulting two model tracks for H and He white dwarfs are shown in Figure~\ref{fig:locusWDs}.
 
@@
 \plotone{plot3diagsBAzoom.png}
-\caption{Preliminary results for white dwarf luminosity-color
-  tracks. When updating this figure with Bob's improved tracks, add also tracks for BHB stars. }
+\caption{Luminosity-color tracks for H and He white dwarfs based on models from \cite{1995PASP..107.1047B}, with
+  colors slightly shifted to bring them in agreement with SDSS photometry. Data, shown as symbols, are the same as in
+  Figure~\ref{fig:3dataDiags} and in the middle panel in Figure~\ref{fig:3HRdiags}. Note the crucial role of the $u-g$ color
+  for distinguishing H and He tracks.}
 \label{fig:locusWDs}
@@
 
-Following \cite{2004ApJ...615L.141S}, we generate luminosity-color tracks for unresolved binaries
-consisting of a main-sequence star and white dwarf.
+Following \cite{2004ApJ...615L.141S}, who discovered the so-called ``second stellar locus'' of unresolved binary
+stars in SDSS dataset, we generate luminosity-color tracks for unresolved binaries consisting of an M dwarf and a white dwarf.
+We limit models to M dwarfs because white dwarfs would have negligible impact on colors of more luminous main-sequence stars. 
+M dwarfs are parametrized with $M_r$ and $[Fe/H]$, and H/He white dwarfs with $M_r$. Therefore, model tracks for unresolved binary
+stars would have three model parameters. After numerical experimentation that revealed small model sensitivity to metallicity, we
+decided to consider only M dwarfs with two metallicity values corresponding to mean metallicity for disk and halo populations:
+$[Fe/H]=0$ and $[Fe/H]=-1.5$, respectively. For the remaining two model parameters, we selected $M_r$ corresponding to the
+total system luminosity and the component luminosity ratio in the $r$ band.
 
-Bob Abel will provide text and plots. Figure~\ref{fig:locusBinaries}
-shows preliminary results.
+We sample the M dwarf luminosity in the range $8.5 \le M_r \le 14.5$, with a step of 0.1 mag, and for each value
+generate a track by adding luminosity of a white dwarf, where the track is sampled on the same grid of $M_r$ but
+using white dwarf models (separately for H and He models). There are four model families (for two families of white dwarf
+models and two values of $[Fe/H]$ for M dwarfs), each with 3,721 $M_r$ vs. color entries. The color tracks for H white dwarfs
+and $[Fe/H]=-1.5$ M dwarfs care shown in Figure~\ref{fig:locusBinaries}. Analogous tracks for binaries composed of an M dwarf with
+$[Fe/H]=0$ and either an H or He white dwarf look very similar though not identical (differences are at most a few tenths
+of a magnitude). In practice, it will be nearly impossible to constrain M dwarf metallicity, and very hard to distinguish
+H and He white dwarfs. It may turn out that it will be sufficient to use a single model family to fit LSST data (e.g., a
+combination of H white dwarf and $[Fe/H]=-1.5$ M dwarf).
 
 \begin{figure*}[ht!]
-\plotone{plot3diagsBA.png}
-\caption{Preliminary results for unresolved binaries composed of a main-sequence star and white dwarf.}
+\plotone{plot3diagsBA_DAh.png}
+\caption{Luminosity-color tracks, shown with gray symbols, for unresolved binaries composed of an M dwarf with
+  $[Fe/H]=-1.5$ and an H white dwarf. Analogous tracks for binaries composed of an M dwarf with
+  $[Fe/H]=0$ and either an H or He white dwarf look very similar though not identical. Data, shown as symbols,
+  are the same as in Figure~\ref{fig:3dataDiags} and in the middle panel in Figure~\ref{fig:3HRdiags}.} 
 \label{fig:locusBinaries}
 \end{figure*}
-
-\subsubsection{Luminosity-Color Tracks for Blue Horizontal Branch Stars}
-
-\v{Z}I has data files from the \cite{2004AJ....127..899S} paper and needs to write a few words
-(and add tracks to Figure~\ref{fig:locusWDs}).
 
@@
 they are variable \citep[e.g., see][]{2007AJ....134.2236S}. The remaining non-variable sources
-that are not well fit with models can be recognized as ``bad fits'' (see the last subsection on
+that are not well fit with available models can be recognized as ``bad fits'' (see the last subsection on
 model selection below).
@@
 Roman surveys). While they will require specialized studies, based on TRILEGAL simulations
-their fraction in the sample should be very small (most likely $<1$\%).
+we expect that their fraction in the sample should be very small (most likely $<1$\%).
 
-\subsubsection{Accounting for Interstellar Dust Extinction}
+\subsubsection{Accounting for Interstellar Dust Extinction \label{sec:ismdust}}
 
@@
 by definition). This choice does not preclude the use of our framework to fit for $R_V$ as the fourth free
-model parameter in Galactic plane regions with very large $A_r$; we  can simply add multiple models
-with extincted colors generated using different values of $C_m$.
+model parameter in Galactic plane regions with very large $A_r$ ($R_V$ is poorly constrained when $A_r$ is small);
+we  can simply add multiple models with extincted colors generated using different values of $C_m$.
 
 In our implementation, constraints on $A_r$ for two nearby stars, even if they have similar distances,
-are independent. We note that one could use the so-called hierarchical Bayesian modeling  and specify prior
+are independent. We note that one could use the so-called hierarchical Bayesian modeling  and specify the prior
 for the line-of-sight extinction profile: nearby stars would then jointly constrain it. For more
@@
 In addition to specifying likelihood $p(\vec{c}|M_r, [Fe/H], A_r)$ for a given model $M$, we need to specify
-prior probability distribution for model parameters, $p(M_r, [Fe/H],A_r)$ (see eq.~\ref{eq:BayesPriorExpand};
+the prior probability distribution for model parameters, $p(M_r, [Fe/H],A_r)$ (see eq.~\ref{eq:BayesPriorExpand};
 we address the model probability $p(M|I)$ further below). First, we assume that the stellar model parameters
@@
 \end{equation}
-and $a=1.3$, $b=0.1$. Therefore, $p(A_r) = 1/A_r^{max}$ for $0 \le A_r \le A_r^{max}$ and $p(A_r) = 0$ for $A_r > A_r^{max}$.
+and $a=1.3$, $b=0.1$, where the choice of these parameters allows for plausible upper limits for additive and
+multiplicative errors in the dust extinction maps. Therefore, $p(A_r) = 1/A_r^{max}$ for $0 \le A_r \le A_r^{max}$
+and $p(A_r) = 0$ for $A_r > A_r^{max}$.
 
@@
 extract all catalog entries from an area of $\sim$10 deg$^2$ (the size of the LSST Camera's field of view) around it, then bin
-the sample using apparent $r$-band magnitude (27 bin centers from $r=14$ to $r=27$, with a bin width of 1 mag). 
+the sample by apparent $r$-band magnitude (27 bin centers from $r=14$ to $r=27$, with a bin width of 1 mag). 
 Given an $r$-band selected subsample, we separate all modeled populations and then bin them using adequate model
-parameters (e.g., $M_r$ and $[Fe/H]$ in case of main-sequence stars and red giants). The resulting 2-dimensional map
+parameters (e.g., $M_r$ and $[Fe/H]$ in the case of main-sequence stars and red giants). The resulting 2-dimensional map
 is renormalized so that its integral over all model parameters is unity.
@@
 
-\subsection{Bayesian Model Selection}
+\subsection{Bayesian Model Selection \label{sec:ModSel}}
 
@@
 \end{equation}
-In practice, the values of the odds ratio are interpreted using Jeffreys' scale; in particular, $O_{21}>10$ represents
-``strong'' evidence in favor of $M_2$ ($M_2$ is ten times more probable than $M_1$).
+In practice, the values of the odds ratio are interpreted using Jeffreys' scale (see, e.g., Chapter 5 in \citealt{2020sdmm.book.....I});
+in particular, $O_{21}>10$ represents ``strong'' evidence in favor of $M_2$ ($M_2$ is ten times more probable than $M_1$).
 

### 2024-03-07 paper/notebooks/paperPlots.py
changes in plot3diagsBobAbel

@@ -49,7 +49,7 @@ def plot3diagsBobAbel(df1, df2, df3, df4, L0, L1, L2, WD, WDMD, BHB, zoom=False)
         ax.plot(giGrid, Mr1, ls='dotted', c='yellow')
         ax.plot(giGrid, Mr2, ls='dotted', c='yellow')
     ax.set_xlim(-1.8,4.1)
-    ax.set_ylim(16.1, 0.0)
+    ax.set_ylim(16.1, 1.0)
     if zoom:
         ax.set_xlim(-2.0, 1.0)
         ax.set_ylim(16.1, 8.0)

### 2024-03-09 paper/methodology.tex
minor edits from Bob

@@
 A comparison with models from \cite{1995PASP..107.1047B}, as well as with SDSS spectra, reveals that the two sequences
-correspond to H and He white dwarfs, with the mean log(g) = 8.0 for H sequence and log(g) = 8.5 for He sequence.
-Upper limits for the scatter of log(g) around these mean sequences appears as no more than 0.5. 
+correspond to H and He white dwarfs, with the mean log(g) = 8.0 for the H sequence and log(g) = 8.5 for the He sequence.
+Upper limits for the scatter of log(g) around these mean sequences appear as no more than 0.5. 
 We use three modern white dwarf catalogs: the Montreal White Dwarf
@@
 Observed magnitudes are corrected for interstellar dust extinction using maps from \cite{schlegel_maps_1998} and per-band extinction coefficients
-discussed in \S\ref{sec:ismdust} below. We group DB and DC spectral classes as He dwarfs (1,939 objects), while DA spectral class corresponds
+discussed in \S\ref{sec:ismdust} below. We group the DB and DC spectral classes as He dwarfs (1,939 objects), while the DA spectral class corresponds
 to H white dwarfs (9,307 objects). Their absolute magnitudes were calculated using ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}.
@@
 models and two values of $[Fe/H]$ for M dwarfs), each with 3,721 $M_r$ vs. color entries. The color tracks for H white dwarfs
-and $[Fe/H]=-1.5$ M dwarfs care shown in Figure~\ref{fig:locusBinaries}. Analogous tracks for binaries composed of an M dwarf with
+and $[Fe/H]=-1.5$ M dwarfs are shown in Figure~\ref{fig:locusBinaries}. Analogous tracks for binaries composed of an M dwarf with
 $[Fe/H]=0$ and either an H or He white dwarf look very similar though not identical (differences are at most a few tenths
@@
 H and He white dwarfs. It may turn out that it will be sufficient to use a single model family to fit LSST data (e.g., a
-combination of H white dwarf and $[Fe/H]=-1.5$ M dwarf).
+combination of an H white dwarf and a $[Fe/H]=-1.5$ M dwarf).
 

### 2024-03-09 paper/notebooks/paperPlots.py

@@ -110,9 +110,10 @@ def plot3diagsBobAbel(df1, df2, df3, df4, L0, L1, L2, WD, WDMD, BHB, zoom=False)
     plt.tight_layout()
     if zoom:
         plt.savefig('plot3diagsBAzoom.png')
+        print('made plot: plot3diagsBAzoom.png')
     else:
         plt.savefig('plot3diagsBA.png')
-    print('made plot: plot3diagsBA.png')
+        print('made plot: plot3diagsBA.png')
     plt.show() 
     plt.close("all")
     return

### 2024-03-09 src/BayesTools.py

@@ -70,7 +70,7 @@ def get2Dmap(sample, labels, metadata):
     Z = kde.evaluate(np.vstack([Xgrid.ravel(), Ygrid.ravel()]))
     return (Xgrid, Ygrid, Z)
 
-def dumpPriorMaps(sample, fileRootname, show2Dmap=True, verbose=True):
+def dumpPriorMaps(sample, fileRootname, show2Dmap=True, verbose=True, NrowMax=200000):
 
     ## data frame called "sample" here must have the following columns: 'FeH', 'Mr', 'rmag'
     labels = ['FeH', 'Mr', 'rmag']
@@ -94,7 +94,7 @@ def dumpPriorMaps(sample, fileRootname, show2Dmap=True, verbose=True):
     # summary file 
     outsumfile = fileRootname + '-SummaryStats.txt'
     foutsum = open(outsumfile, "w")
-    foutsum.write(" rMin    Amin    Amed    Amax     Ntotal \n")
+    foutsum.write(" rMin    rMax   Ntotal     Amin   Amed    Amax     pMS      ppMS     pAGB      pWD        pA1       pA2      pA3      pA4       ptnD     ptkD       pH        pB        pMC \n")
     for rind, r in enumerate(rGrid):
         # r magnitude limits for this subsample
         rMin = r - rmagBinWidth
@@ -102,11 +102,20 @@ def dumpPriorMaps(sample, fileRootname, show2Dmap=True, verbose=True):
         # select subsample
         tS = sample[(sample[labels[2]]>rMin)&(sample[labels[2]]<rMax)]
         if verbose: print('r=', rMin, 'to', rMax, 'N=', len(sample), 'Ns=', len(tS))
+        if (len(tS)>NrowMax):
+            tSmap = tS.sample(n=NrowMax)
+            if verbose: print('subsampled for 2D map to n=', NrowMax)
+        else:
+            tSmap = tS
         # this is work horse, where data are binned and map arrays produced
         # it takes about 2 mins on Mac OS Apple M1 Pro
         # so about 70 hours for 2,000 healpix samples 
         # maps per healpix are about 2 MB, total 4 GB
-        xGrid, yGrid, Z = get2Dmap(tS, labels, metadata)
+        if (len(tSmap) < 1):
[…]

### 2024-03-09 src/MakePriors.py
first import, tools for computing TRILEGAL-based priors

@@ -1,0 +1,163 @@
+### tools to support the production of metadata and maps with priors based on TRILEGAL simulations
+
+## import supporting local tools: 
+import LocusTools as lt
+import BayesTools as bt
+import numpy as np
+from dl import authClient as ac, queryClient as qc
+from dl.helpers.utils import convert
+from getpass import getpass
+
+## define query with RA/Dec boundaries and a list of quantities to retrieve from lsst_sim.simdr2 table at Astro Lab 
+# TRILEGAL NOTE:
+# gc depends on which Galactic component the star belongs to: 
+#    1 → thin disk; 2 → thick disk; 3 → halo; 4 → bulge; 5 → Magellanic Clouds.
+# label depends on the evolutionary phase of the star, as:
+# 0 pre-main sequence
+# 1 main sequence
+# 2 Hertzsprung gap
+# 3 red giants
+# 4–6 core He burning 
+# 7 EAGB
+# 8 TPAGB
+# 9 Post-AGB
+# 10 CO-WD 
+# 21 22 23 24 25 26 27 He, NS, BH
+def makeQueryString(minRA, maxRA, minDec, maxDec, umagMax=99.99):
+    query1 = """
+    SELECT ra, dec, gall, galb, gc, logage, mass, label, logg, m_h, av, mu0, umag, gmag, rmag, imag, zmag, ymag
+    FROM lsst_sim.simdr2
+    WHERE"""
+    query2 = f" {minRA} < ra and ra < {maxRA}"
+    query3 = f" AND {minDec} < dec and dec < {maxDec}"
+    query4 = f" AND umag < {umagMax}"
+    query = query1 + query2 + query3 + query4 + " \n"
+    return query
+
+## given a query, retrieve data from the database
+def retrievePatch(query, verbose=True):
+    # similar to processPatch() below but adds ugiz magnitudes
+    if verbose: print('querying...')
+    result = qc.query(sql=query)
+    if verbose: print('converting...')
+    trilegal = convert(result)
+    trilegal['Ar'] = 2.75 * trilegal['av'] / 3.10 
+    trilegal['Mr'] = trilegal['rmag'] - trilegal['Ar'] - trilegal['mu0']
+    trilegal2 = trilegal.rename(columns={"m_h": "FeH"})
+    df = trilegal2.rename(columns={"mu0": "DM"})
[…]

### 2024-03-10 src/MakePriors.py
minor changes

@@ -25,7 +25,7 @@ from getpass import getpass
 # 21 22 23 24 25 26 27 He, NS, BH
 def makeQueryString(minRA, maxRA, minDec, maxDec, umagMax=99.99):
     query1 = """
-    SELECT ra, dec, gall, galb, gc, logage, mass, label, logg, m_h, av, mu0, umag, gmag, rmag, imag, zmag, ymag
+    SELECT ra, dec, gall, galb, gc, logage, mass, label, logg, m_h, av, mu0, umag, gmag, rmag, imag, zmag, ymag, nest4096
     FROM lsst_sim.simdr2
     WHERE"""
     query2 = f" {minRA} < ra and ra < {maxRA}"

### 2024-03-11 src/LocusTools.py
minor changes

@@ -286,8 +286,8 @@ def readAllDSED(DSEDdatadir):
     for age in ages:
         DSEDFeHlist = []
         for DSED in DSEDlist:
-            if (age > 5.0):
-                DSED = fixLocus10Gyr(DSED)
+            # if (age > 5.0):
+                # DSED = fixLocus10Gyr(DSED)
             DSEDFeHlist.append(DSED[DSED['ageGyr']==age])
         DSEDiso.append(DSEDFeHlist)             
 
@@ -390,9 +390,10 @@ def DSEDlist2locusAugmented(DSEDlist):
         locus.rename_column('col0', 'Mr')     
         locus.rename_column('col1', 'FeH')     
         for c in colors:
-            locus[c] = interpolateColorMr(d['Mr'], d[c], MrGrid)
+            locus[c], distanceAlongLocus = interpolateColorMr(d['Mr'], d[c], MrGrid)
+            locus['Dalong_'+c] = distanceAlongLocus 
             # locus[c] = interpolateColorMrLinear(d['Mr'], d[c], MrGrid)
-        locus['gi'] = locus['gr'] + locus['ri'] 
+        locus['gi'] = locus['gr'] + locus['ri']
         if (i==0):
             DSEDlocus = locus
         else:

### 2024-03-11 src/MakePriors.py

@@ -40,11 +40,16 @@ def retrievePatch(query, verbose=True):
     if verbose: print('querying...')
     result = qc.query(sql=query)
     if verbose: print('converting...')
-    trilegal = convert(result)
-    trilegal['Ar'] = 2.75 * trilegal['av'] / 3.10 
-    trilegal['Mr'] = trilegal['rmag'] - trilegal['Ar'] - trilegal['mu0']
-    trilegal2 = trilegal.rename(columns={"m_h": "FeH"})
-    df = trilegal2.rename(columns={"mu0": "DM"})
+    #trilegal = convert(result)
+    #trilegal['Ar'] = 2.75 * trilegal['av'] / 3.10 
+    #trilegal['Mr'] = trilegal['rmag'] - trilegal['Ar'] - trilegal['mu0']
+    #trilegal2 = trilegal.rename(columns={"m_h": "FeH"})
+    #df = trilegal2.rename(columns={"mu0": "DM"})
+    df = convert(result)
+    df['Ar'] = 2.75 * df['av'] / 3.10 
+    df['Mr'] = df['rmag'] - df['Ar'] - df['mu0']
+    df.rename(columns={"m_h": "FeH"})
+    df.rename(columns={"mu0": "DM"})
     # colors corrected for extinction
     C = lt. extcoeff()
     for b in ['u', 'g', 'r', 'i', 'z']:

### 2024-03-12 paper/notebooks/paperPlots.py
added code for stitching DSED isochrones

@@ -470,6 +470,71 @@ def compare2isochrones(dataDF, isoDF1, isoDF2, alpha1=0.05, alpha2=0.8, title=""
 
 
 
+def compare2isochrones2(dataDF, isoDF1, isoDF2, alpha1=0.05, alpha2=0.8, title="", plotname='compare2isochrones.png'):
+
+    fig,ax = plt.subplots(3,2,figsize=(12,12))
+
+    plt.rcParams.update({'font.size': 12})
+    plt.rc('axes', labelsize=14)    
+
+    ax[0,0].scatter(dataDF['gi'], dataDF['Mr'], s=0.1, c='black')
+    ax[0,0].scatter(isoDF1['gi'], isoDF1['Mr'], s=0.1, c=isoDF1['Mr'], cmap=plt.cm.jet, alpha=alpha1)
+    ax[0,0].set_xlim(-1.0, 4.2)
+    ax[0,0].set_ylim(16.5, -3.2)
+    ax[0,0].set_xlabel('g-i')
+    ax[0,0].set_ylabel('Mr')
+
+    ax[0,1].scatter(dataDF['gi'], dataDF['Mr'], s=0.1, c='black')
+    ax[0,1].scatter(isoDF2['gi'], isoDF2['Mr'], s=0.1, c=isoDF2['Mr'], cmap=plt.cm.jet, alpha=alpha2)
+    ax[0,1].set_xlim(-1.0, 4.2)
+    ax[0,1].set_ylim(16.5, -3.2)
+    ax[0,1].set_xlabel('g-i')
+    ax[0,1].set_ylabel('Mr')
+    ax[0,0].set_title(title, loc='center')
+
+    ax[1,0].scatter(dataDF['ug'], dataDF['gr'], s=0.1, c='black')
+    if (1):
+        flag = (isoDF1['Mr']>0.5)
+        isoDF1x = isoDF1[flag]
+        ax[1,0].scatter(isoDF1x['ug'], isoDF1x['gr'], s=0.1, c=isoDF1x['Mr'], cmap=plt.cm.jet, alpha=alpha1)
+    else:
+        ax[1,0].scatter(isoDF1['ug'], isoDF1['gr'], s=0.1, c=isoDF1['Mr'], cmap=plt.cm.jet, alpha=alpha1)
+    ax[1,0].set_xlim(-1.0,4.2)
+    ax[1,0].set_ylim(-0.5, 2.2)
[…]

### 2024-03-12 src/LocusTools.py

@@ -366,12 +366,12 @@ def DSEDlist2locus(DSEDlist):
     # given a list of DSED tables (e.g. 31 for different FeH values)
     # make a table with (Mr, FeH, {colors}), such as SDSS master locus
     d = DSEDlist[0]
-    sublist = (d['Mr'], d['FeH'], d['ug'], d['gr'], d['ri'], d['iz'], d['gi'])
+    sublist = (d['tLoc'], d['Mr'], d['FeH'], d['ug'], d['gr'], d['ri'], d['iz'], d['gi'])
     DSEDlocus = Table(sublist, copy=True)
     # loop 
     for i in range(1,len(DSEDlist)):
         d = DSEDlist[i]
-        sublist = (d['Mr'], d['FeH'], d['ug'], d['gr'], d['ri'], d['iz'], d['gi'])
+        sublist = (d['tLoc'], d['Mr'], d['FeH'], d['ug'], d['gr'], d['ri'], d['iz'], d['gi'])
         locus = Table(sublist, copy=True)
         DSEDlocus = vstack([DSEDlocus, locus])
     return DSEDlocus 
@@ -390,8 +390,7 @@ def DSEDlist2locusAugmented(DSEDlist):
         locus.rename_column('col0', 'Mr')     
         locus.rename_column('col1', 'FeH')     
         for c in colors:
-            locus[c], distanceAlongLocus = interpolateColorMr(d['Mr'], d[c], MrGrid)
-            locus['Dalong_'+c] = distanceAlongLocus 
+            locus[c] = interpolateColorMr(d['Mr'], d[c], MrGrid)
             # locus[c] = interpolateColorMrLinear(d['Mr'], d[c], MrGrid)
         locus['gi'] = locus['gr'] + locus['ri']
         if (i==0):
@@ -496,15 +495,19 @@ def getDSEDages(DSEDdatadir, DSEDfile):
 def assignAges(DSED, ages):
     k = 0
     DSED['ageGyr'] = DSED['logTeff']
+    DSED['tLoc'] = DSED['logTeff']
     for i in range(0, len(DSED)):
         if i==0: 
             DSED['ageGyr'][i] = ages[k]
+            DSED['tLoc'][i] = 0.0
         else:
             if DSED['EEP'][i]>DSED['EEP'][i-1]:
                 DSED['ageGyr'][i] = ages[k]
+                DSED['tLoc'][i] =  DSED['tLoc'][i-1] + 1.0 
[…]

### 2024-03-12 src/LocusTools.py
changed default  fixForStripe82 to False

@@ -911,7 +911,7 @@ def LSSTsimsLocus(fixForStripe82=True, datafile=""):
 
 # similar to LSSTsimsLocus, but with tLoc running variable 
 
-def readSDSSDSEDlocus(datafile, fixForStripe82=True): 
+def readSDSSDSEDlocus(datafile, fixForStripe82=False): 
         ## Mr, as function of [Fe/H], along the SDSS/LSST stellar 
         ## for more details see the file header
         colnames = ['tLoc', 'Mr', 'FeH', 'ug', 'gr', 'ri', 'iz']

### 2024-04-08 src/makeSimPhotometry.py
packaged code for simulating photometry

@@ -1,0 +1,221 @@
+import numpy as np
+from scipy.interpolate import griddata
+import matplotlib.pyplot as plt
+from matplotlib.colors import LogNorm
+from scipy.stats import gaussian_kde
+import LocusTools as lt 
+import BayesTools as bt 
+import PlotTools as pt
+from astropy.table import Table, vstack
+
+
+def makeSimLSSTcatFromTRILEGAL(infile, MSRGlocus, WDlocusH, WDlocusHe, outfile, verbose=False):
+
+    ## 1) given rmag, Ar and DM from TRILEGAL, generate Mr 
+    ## 2) with this Mr and FeH, generate colors from locus data  
+    ## 3) with original r magnitude, generate other magnitudes from colors
+    ## 4) given magnitudes, generate errors for LSST
+    ## 5) generate observed mags
+    ## 6) regenerate "observed" errors from "observed" magnitudes
+    ## 7) and return data frame 
+
+    ## 1)
+    ## read catalog produced at Astro Lab, using dumpCatalogFromPatch in makePriors.py
+    ## magnitudes (umag, gmag...) are not corrected for extinction, but colors are
+    cat = lt.readAstroLabCatalog(infile, verbose=verbose)
+
+    # save original TRILEGAL magnitudes
+    for b in ['u', 'g', 'r', 'i', 'z']:  
+        cat[b+'magTL'] = cat[b+'mag']
+        ## and this is a hack to "fix" TRILEGAL [Fe/H] distribution to be more similar to SDSS measurements
+    if (1):
+        # shift thin disk stars to lower [Fe/H] by 0.3 dex
+        cat['FeH'] = np.where(cat['GC']==1, cat['FeH']-0.3, cat['FeH'])
+        # shift halo stars to higher [Fe/H] by 0.1 dex
+        cat['FeH'] = np.where(cat['GC']==3, cat['FeH']+0.1, cat['FeH'])
+
+    ## separate MS/RGs from white dwarfs 
+    catMSRG = cat[cat['pop']<9]
+    catWD = cat[cat['pop']==9]
+    if verbose: print('MSRG:', np.size(catMSRG), 'WD:', np.size(catWD))
+
+    ## 2)     
+    ## locus data
[…]

### 2024-04-08 src/BayesTools.py
progress...

@@ -5,6 +5,7 @@ from matplotlib.colors import LogNorm
 from scipy.stats import gaussian_kde
 import LocusTools as lt 
 import PlotTools as pt
+from astropy.table import Table
 
 
 ### numerical model specifications and constants for Bayesian PhotoD method
@@ -24,11 +25,28 @@ def getBayesConstants():
     BayesConst['rmagMax'] = 27
     BayesConst['rmagNsteps'] = 27
     # parameters for dust extinction grid when fitting 3D (Mr, FeH, Ar)
-    # prior for Ar is flat from 0 to (1.3*Ar+0.1) with a step of 0.01 mag ***
+    # prior for Ar is flat from (0*Ar+0.0) to (1.3*Ar+0.1) with a step of 0.01 mag ***
     # where 1.3 <= ArCoef0, 0.1 <= ArCoef1 and 0.01 mag <= ArCoef2 are set here (or user supplied)
+    # ArMin: ArCoeff3*Ar + ArCoeff4 
     BayesConst['ArCoeff0'] = 1.3
     BayesConst['ArCoeff1'] = 0.1
     BayesConst['ArCoeff2'] = 0.01
+    if False:
+        # allow only 0.01 mag around the provided (e.g. SFD) Ar value
+        print('bt.getBayesConstants: DONT FORGET TO UNDO HACK THAT limits Ar range!!!') 
+        BayesConst['ArCoeff3'] =  0.5
+        BayesConst['ArCoeff4'] = -0.01
+    else:
+        # allow Ar as small as 0
+        BayesConst['ArCoeff3'] = 0.0
+        BayesConst['ArCoeff4'] = 0.0
+        
+    if False:
+        ## for testing with fitting Ar: 
+        print('bt.getBayesConstants: DONT FORGET TO UNDO HACK THAT SETS Ar=0!!!') 
+        BayesConst['ArCoeff0'] = 0.0
+        BayesConst['ArCoeff1'] = 0.01
+        BayesConst['ArCoeff2'] = 0.01
     
     return BayesConst   
 
@@ -205,6 +223,7 @@ def readPriors(rootname, locusData, MrColumn='Mr'):
         values = Zval.flatten()
         # actual (linear) interpolation
         priorGrid[rind] = griddata(points, values, (locusData['FeH'], locusData[MrColumn]), method='linear', fill_value=0)
+ 
     return priorGrid
  
         
[…]

### 2024-04-08 src/LocusTools.py

@@ -222,7 +222,7 @@ def GClocusGiants(Mr, FeH):
     color['FeH'] = 0*color['iz'] + FeH
     return color
 
-def CoveyMSlocus(): 
+def CoveyMSlocus(datafile=''): 
         ## STELLAR LOCUS IN THE SDSS-2MASS PHOTOMETRIC SYSTEM
         ## read and return colors from
         ## http://faculty.washington.edu/ivezic/sdss/catalogs/tomoIV/Covey_Berry.txt
@@ -230,7 +230,8 @@ def CoveyMSlocus():
         ## according to Yuan et al (2015, ApJ 799, 134), this locus approximately corresponds to FeH = -0.5 
         colnames = ['Nbin', 'gi', 'Ns', 'ug', 'Eug', 'gr', 'Egr', 'ri', 'Eri', 'iz', 'Eiz', \
                      'zJ', 'EzJ', 'JH', 'EJH', 'HK', 'EHK']
-        coveyMS = Table.read('./Covey_Berry.txt', format='ascii', names=colnames)
+        if datafile=='': datafile='./Covey_Berry.txt'
+        coveyMS = Table.read(datafile, format='ascii', names=colnames)
         return coveyMS 
 
 
@@ -674,7 +675,25 @@ def smoothLocus(locus):
     return locusSmooth
 
 
+### read catalog produced at Astro Lab by dumpCatalogFromPatch() from MakePriors.py
+def readAstroLabCatalog(infile, verbose=False):
+        if verbose: print('reading from:', infile)
+        colnames = ['ra', 'dec', 'glon', 'glat', 'GC', 'logage', 'mass', 'pop', 'logg', 'FeH', 'Mr', 'Ar', 'DM',
+                     'umag',  'gmag', 'rmag', 'imag', 'zmag', 'ymag', 'ug', 'gr', 'ri', 'iz', 'gi']
 
+        # GC: Galactic component the star belongs to: 1 → thin disk; 2 → thick disk; 3 → halo; 4 → bulge; 5 → Magellanic Clouds.
+        # colors are corrected for dust reddening (magnitudes are not) 
+        catalog = Table.read(infile, format='ascii', names=colnames)
+        return catalog 
+
+
+### read White Dwarf locus produced from Bob Abel's inputs
+def readWDlocus(infile, verbose=False):
+        if verbose: print('reading from:', infile)
[…]

### 2024-04-08 src/MakePriors.py

@@ -40,16 +40,11 @@ def retrievePatch(query, verbose=True):
     if verbose: print('querying...')
     result = qc.query(sql=query)
     if verbose: print('converting...')
-    #trilegal = convert(result)
-    #trilegal['Ar'] = 2.75 * trilegal['av'] / 3.10 
-    #trilegal['Mr'] = trilegal['rmag'] - trilegal['Ar'] - trilegal['mu0']
-    #trilegal2 = trilegal.rename(columns={"m_h": "FeH"})
-    #df = trilegal2.rename(columns={"mu0": "DM"})
     df = convert(result)
     df['Ar'] = 2.75 * df['av'] / 3.10 
     df['Mr'] = df['rmag'] - df['Ar'] - df['mu0']
-    df.rename(columns={"m_h": "FeH"})
-    df.rename(columns={"mu0": "DM"})
+    df2 = df.rename(columns={"m_h": "FeH"})
+    df = df2.rename(columns={"mu0": "DM"})
     # colors corrected for extinction
     C = lt. extcoeff()
     for b in ['u', 'g', 'r', 'i', 'z']:

### 2024-04-08 src/PlotTools.py

@@ -46,6 +46,8 @@ def plot2Dmap(x, y, xMin, xMax, nXbin, yMin, yMax, nYbin, xLabel, yLabel, addMed
     ygrid = np.linspace(yMin, yMax, nYbin)
     Xgrid, Ygrid = np.meshgrid(xgrid, ygrid)
     Z = kde.evaluate(np.vstack([Xgrid.ravel(), Ygrid.ravel()]))
+
+    print('nXbin=', nXbin)
     
     # plot the result as an image
     if (logScale):
@@ -75,7 +77,7 @@ def plot2Dmap(x, y, xMin, xMax, nXbin, yMin, yMax, nYbin, xLabel, yLabel, addMed
          plt.plot([2.5, 10.5], [0, 0], ls='--', lw=1, c='red')
    
     if (addMedians):     
-        xBin, nPts, medianBin, sigGbin = lt.fitMedians(x, y, xMin, xMax, 60)
+        xBin, nPts, medianBin, sigGbin = lt.fitMedians(x, y, xMin, xMax, nXbin)
         plt.scatter(xBin, medianBin, s=40, c='red')
         plt.plot(xBin, medianBin+sigGbin*np.sqrt(nPts), c='yellow')
         plt.plot(xBin, medianBin-sigGbin*np.sqrt(nPts), c='yellow')
@@ -461,8 +463,8 @@ def qpB(df, Qname, Dname='Mr', cmd=False, estQ=False):
             xAxis = 'MrEst'
             yAxis = 'FeHEst'
             yTicks = [-2.0, -1.5, -1.0, -0.5, 0.0, 0.5] 
-            xlabel = '$\mathrm{Mr estimate}$'
-            ylabel = '$\mathrm{[Fe/H] estimate}$'
+            xlabel = 'Mr estimate'
+            ylabel = '[Fe/H] estimate'
         else:
             xAxis = 'Mr'
             yAxis = 'FeH'
@@ -493,7 +495,7 @@ def qpB(df, Qname, Dname='Mr', cmd=False, estQ=False):
     plt.ylabel(ylabel)
     if (cmd):
         plt.xlim(0.0, 4.0)
-        plt.ylim(28, 15)
+        plt.ylim(27, 15)
     else:
         plt.xlim(xedges[0], xedges[-1])
         plt.ylim(yedges[0], yedges[-1])
@@ -510,7 +512,7 @@ def qpB(df, Qname, Dname='Mr', cmd=False, estQ=False):
             plt.clim(-0.15, 0.15)
[…]

### 2024-04-08 paper/analysis.tex
progress...

--- 2023-11-06 paper/analysis.tex
@@
 % -- kuglasti skupovi
+
+\section{Implementation: Photo-D pipeline and Tests}
+
+First describe pipeline and implementation choices, then its performance
+
+% Our method is basically brute-force fitting with some intelligent tricks leveraged to obtain faster execution that will be required for 10B LSST stars.
+
+\subsection{Accelerated Exhaustive Grid Search Method}
+optimized grid search, MCMC, refer to NN paper, also in Discussion
+
+Our fitting procedure is also executed on an adaptive grid, a coarse search over the parameter space is performed first in order to establish the layout of the manifold.
+However, care is taken that any possible local minima are not missed by appropriately adjusting the step size \magcom{how?}.
+The located maxima are then explored with a smaller step size (\magcom{adjusted how?}).
+
+\subsection{Markov Chain Monte Carlo Method}
+
+In addition to the approach described here, we also tested Markov Chain Monte Carlo and neural network approaches that will be/are described in forthcoming/published papers.
+
+ XXX Green et al. (2014) used MCMC
+
+\subsection{Neural Network Method}
+
+\subsection{LSDB and Healpix Implementation}
+
+% 2) statistics:
+
+The priors are established by partitioning the TRILEGAL galaxy model
+\citep{dal_tio_simulating_2022} in healpixels,
+and each of the pixel in one-magnitude wide bins in apparent
+magnitude.
+
+\subsection{Performance Testing}
+
+Motivate three parts: simulated photometry, SDSS-Gaia sample and DECam bulge sample.
+
+\subsubsection{Simulated Data Set}
+
+\begin{figure*}[ht!]
+\hskip -0.3in
+\includegraphics[width=1.08\textwidth,angle=0]{figures/qpBmeans_SDSSpatchRA340-350-simLSST_FeH.png}
+\vskip -0.3in
+\caption{The mean per-pixel values of absolute magnitude $M_r$,
+  metallicity $[Fe/H]$, and interstellar dust extinction along the line of sight in
+  the $r$ band, $A_r$, for a TRILEGAL-simulated sample of 283,000 main
+  sequence and red giants stars with $r<26$, $340^{\degree} < {\rm R.A.} < 350^{\degree}$
+  and $-1.3^{\degree} < \delta < 1.3^{\degree}$
+  (a small patch from the SDSS Stripe 82 region). The mean values are
+  color-coded according to legends below each panel. Contours
+  visualize the sample distribution in each diagram. The strong variation
+  of $M_r$ with the $g-i$ color is seen in the left panel. The
+  diagonal iso-metallicity boundaries in the middle panel closely correspond to distance
+  (in the range from about 1 kpc to about 100 kpc). The values of
+  dust extinction, shown in the right panel, are not large because the
+  selected field is at high Galactic latitudes (centered on $b = -52^{\degree}$).}
+\label{fig:qpBmeans}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{qpB_SDSSpatchRA340-350-simLSST_Mr.png}
+\plotone{qpB_SDSSpatchRA340-350-simLSST_FeH.png}
+\plotone{qpB_SDSSpatchRA340-350-simLSST_Ar.png}
+\caption{Performance analysis for estimates of model parameters $M_r$
+  (top row),  $[Fe/H]$ (middle row), and $A_r$ (bottom row) in the
+  space of $M_r$ and $[Fe/H]$, for the same simulated sample as shown
+  in Figure~\ref{fig:qpBmeans}. 
+  The left column shows the mean difference per pixel between the true
+  and estimated values, the middle column shows scatter per pixel and
+  the right column shows the scatter normalized by estimated
+  uncertainties. Contours visualize the sample distribution in each
+  diagram. Note that for main sequence stars $M_r>4$ and for most
+  stars $4 < M_r < 10$. 
+\label{fig:perfVStrueParams}}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{qpB_estQ_SDSSpatchRA340-350-simLSST_Mr.png}
+\plotone{qpB_estQ_SDSSpatchRA340-350-simLSST_FeH.png}
+\plotone{qpB_estQ_SDSSpatchRA340-350-simLSST_Ar.png}
+\caption{Analogous to Figure~\ref{fig:perfVStrueParams}, except that
+  the results are binned in the space of estimated $M_r$ and $[Fe/H]$ parameters.
+\label{fig:perfVSestParams}}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{qpBcmd_SDSSpatchRA340-350-simLSST_Mr.png}
+\plotone{qpBcmd_SDSSpatchRA340-350-simLSST_FeH.png}
+\plotone{qpBcmd_SDSSpatchRA340-350-simLSST_Ar.png}
+\caption{Analogous to Figures~\ref{fig:perfVStrueParams} and \ref{fig:perfVSestParams}
+except that the results are binned in the space of 
+  of observables: $u$ band magnitude and the $g-i$ color
+\label{fig:perfVScmd}}
+\end{figure*}
+
+Basic tests of numerics, intrinsic model degeneracies and performance
+as a function of photometric signal-to-noise ratio...
+
+Simulated photometry using color sequences for 10 Gyr isochrone shown in Figure~\ref{fig:augmLocus}. 
+and photometric errors expected for LSST coadded depths (Bianco et al. )
+
+Main summary in Table 1 (need proper latex), more details in figures below.
+
+No trend with u mag, only increase in scatter towards the faint end.
+
+The distribution of model parameters (absolute magnitude $M_r$,
+metallicity $[Fe/H]$, and interstellar dust extinction along the line
+of sight in the $r$ band, $A_r$) for the simulated sample in the space
+of observables: $u$ band magnitude and the $g-i$ color, is shown in
+Figure~\ref{fig:perfVScmd}.
+
+Performance for estimates of model parameters $M_r$, $[Fe/H]$ and
+$A_r$ is illustrated in the space of model parameters in Figures~\ref{fig:perfVStrueParams}
+and \ref{fig:perfVSestParams}, and in the space of observables in
+Figure~\ref{fig:perfVScmd}.
+
+\newpage
+
+{\bf Quantitative summary of the performance of Bayesian method on simulated sample:}
+Each column shows the median error, robust standard deviation and the subsample size. 
+``mag selected'' refers to a subsample selected by $u < 24$. Low and high $[Fe/H]$ subsamples are
+separated at $[Fe/H] = -1$. ``Giants'' are selected by $M_r<4$ and blue/red (i.e., bright/faint) main
+sequence stars are separated by $M_r=8$.
+
+\begin{verbatim}
+-----------------------------------------------------------------------------------------------
+    FULL SAMPLE:           dMr                      dFeH                         Ar
+           all: [ '0.010', '0.28', 275863] [ '0.000', '0.25', 275863] ['0.000', '0.04',275863]
+  mag selected: [ '0.000', '0.24',  94161] ['-0.010', '0.14',  94161] ['0.000', '0.04', 94161]
+-----------------------------------------------------------------------------------------------
+     low [FeH]:            dMr                      dFeH                         Ar
+           all: ['-0.020', '0.28',  43470] [ '0.010', '0.14',  43470] ['0.000', '0.06', 43470]
+        giants: ['-0.430', '0.51',   9165] ['-0.060', '0.15',   9165] ['0.020', '0.07',  9165]
+       blue MS: [ '0.030', '0.23',  33609] [ '0.020', '0.13',  33609] ['0.000', '0.05', 33609]
+        red MS: [ '0.040', '0.14',    696] ['-0.060', '0.22',    696] ['0.000', '0.05',   696]
+    high [FeH]:            dMr                      dFeH                         Ar
+           all: [ '0.020', '0.21',  50691] ['-0.020', '0.14',  50691] ['0.000', '0.04', 50691]
+        giants: ['-0.570', '0.91',   5721] ['-0.050', '0.20',   5721] ['0.000', '0.06',  5721]
+       blue MS: [ '0.040', '0.20',  32841] [ '0.000', '0.13',  32841] ['0.000', '0.04', 32841]
+        red MS: [ '0.060', '0.14',  12129] ['-0.070', '0.16',  12129] ['0.000', '0.02', 12129]
+-----------------------------------------------------------------------------------------------
+\end{verbatim}
+
+% 3) code testing:
+
+% *** data-based comparisons ***
+\subsubsection{SDSS-based Comparison to Gaia's Distance Scale}
+
+\begin{figure*}[ht!]
+\plotone{gaiaStripe82plot1.png}
+\caption{Analysis of the difference between Gaia's geometric (trigonometric) absolute magnitudes and photometric magnitudes
+  derived from SDSS colors. Symbols show a sample of 33,186 stars with the parallax signal-to-noise ratio above 10, absolute
+  magnitude MrGeo$>$4 and MrPho$>$4 , and SDSS-based $u<21$ and $0.2 < g-r < 0.6$. The median magnitude difference is
+  $-0.01$ mag, with a scatter of 0.28 mag. The four panel illustrate variation of the binned median (red circles) and scatter (red
+  lines) with the $u$-band magnitude, the parallax signal-to-noise ratio, SDSS $g-r$ color, and photometric metallicity.} 
+\label{fig:gaiasdss1}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{gaiaStripe82plot2.png}
+\caption{Analysis of the difference between Gaia's photometric (color-based) absolute magnitudes and photometric magnitudes
+  derived from SDSS colors. Symbols show a sample of 165,834 stars with absolute magnitude MrPho$>$4, and SDSS-based $u<21$
+  and $0.2 < g-r < 0.6$. The median magnitude difference is $-0.05$ mag, with a scatter of 0.32 mag. The four panel illustrate
+  variation of the binned median (red circles) and scatter (red lines) with the $u$-band magnitude, SDSS $u-g$ and $g-r$ colors,
+  and photometric metallicity.}
+\label{fig:gaiasdss2}
+\end{figure*}
+
+Stripe 82 vs. Gaia...  It works great for $0.2 < g-r < 0.6$ stars using directly photometric $[Fe/H]$ and $M_r$ expressions from the
+Tomography papers. See Figures~\ref{fig:gaiasdss1} and \ref{fig:gaiasdss2}. Does it work within the Bayesian framework? Does it work outside this color range?
+
+\subsubsection{Performance in Bulge Fields}
+
+Bulge BDSD sample...  Different photometric system, differences compared to model colors. Should we descope it?

### 2024-04-08 paper/methodology.tex
progress...

@@
 \caption{Augmented SDSS color tracks for two isochrone ages (left: 1 Gyr; right: 10 Gyr) and the full metallicity range
-  ($-2.5 < [Fe/H] < +0.5$). The black dots show the same sample of 63,000 stars from the middle panel in
+  ($-2.5 < [Fe/H] < +0.5$, color-coded linearly from blue to red). The black dots show the same sample of 63,000 stars from the middle panel in
   Figure~\ref{fig:3HRdiags}. As can be seen in the top right panel, sub-giant stars that could not be fit with empirical
@@
 \label{fig:augmLocus}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{RGdegeneracy.png}
+\caption{Analysis of color degeneracies between giants and main  sequence stars. The left panel shows minimum color distance in
+  4-dimensional color space ($u-g$, $g-r$, $r-i$, $i-z$) between a position on the locus with $M_r<4$ (giants) and any position on the
+  main-sequence $M_r>4$ locus (in magnitudes). Symbols are color-coded by metallicity, linearly from blue to red for the range $-2.5$ to 0.5).
+  The middle and right panels show errors in absolute magnitude and metallicity when a giant is misidentified as a main sequence star
+  closest to it in color space.} 
+\label{fig:RGdegeneracy}
 \end{figure*}
@@
 a fine age grid can be easily executed.
+
+We note strong degeneracies in color space between giants and main sequence stars (see Figure~\ref{fig:RGdegeneracy}). 
+They are especially strong for sub-giant stars with $2 < M_r < 4$, for which it is always possible to find a matching main sequence star
+closer in 4-dimensional color distance than 0.02 mag (and for $3 < M_r < 4$ even closer than 0.01 mag). Due to photometric scatter, 
+such subgiant stars can be misidentified as a main sequence star and absolute magnitude errors can range up to several magnitudes.
 

### 2024-04-08 paper/notebooks/paperPlots.py

@@ -4,6 +4,9 @@ import pylab as plt
 import scipy.stats as stats
 from scipy.stats import norm
 from astroML.stats import binned_statistic_2d
+import LocusTools as lt 
+import BayesTools as bt 
+import PlotTools as pt
 
 
 def plot3diagsBobAbel(df1, df2, df3, df4, L0, L1, L2, WD, WDMD, BHB, zoom=False):
@@ -404,6 +407,7 @@ def show4diagsDSED(age, ages, FeHlist, DSEDiso, FeHlocus3vals, Lcomparison, data
     showRIZ(dataDF, Lcomparison, DSEDs)
     return
 
+ 
 
 
 def compare2isochrones(dataDF, isoDF1, isoDF2, alpha1=0.05, alpha2=0.8, title="", plotname='compare2isochrones.png'):
@@ -740,3 +744,427 @@ def compare2isochronesColorMrAlongLocus2(df1, df2, alpha=0.8, title="", plotname
     plt.show() 
     plt.close("all")
     return
+
+
+
+
+
+def compare2isochronesColorDistance(df1, df2, alpha=0.8, title="", plotname='compare2isochronesColorMrAlongLocus.png'):
+
+    fig,ax = plt.subplots(7,2,figsize=(12,12))
+
+    def plotPanel(xP, yP, df, xName, yName, cName, xMin, xMax, yMin, yMax, xLabel, yLabel, alpha):
+        ax[xP,yP].scatter(df[xName], df[yName], s=0.3, c=df[cName], cmap=plt.cm.jet, alpha=alpha)
+        ax[xP,yP].set_xlim(xMin, xMax)
+        ax[xP,yP].set_ylim(yMin, yMax)
+        ax[xP,yP].set_xlabel(xLabel)
+        ax[xP,yP].set_ylabel(yLabel)
+
+#   tLoc    Mr    FeH     ug      gr      ri      iz   
+#   3.46   3.00 -1.50   0.959   0.316   0.116   0.030
+    ugT = 0.959
+    grT = 0.316
+    riT = 0.116
+    izT = 0.030 
+
+    df = df1
[…]

### 2024-05-07 paper/analysis.tex
last changes before distributing first draft

--- 2024-01-19 paper/analysis.tex
@@
+% -- choice of grid search vs. MCMC, and NN-based acceleration
+% -- method validation with simulated photometry
+% -- distance scale validation with SDSS-Gaia Stripe 82 sample 
+% -- 3D dust map derived from SDSS data in Galactic plane 
+% 
+%   for spinoff papers: 
+%      -- kuglasti skupovi
+%      -- aginst Rich et al. BDSD survey ugriz podaci
 
-% -- aginst Rich et al. BDSD survey ugriz podaci
-% -- Bailer-Jones i S82
-% -- Trilegal simulacije s SDSS SED modelima (validacija)
-% -- kuglasti skupovi
+\section{Numerical Implementation and Performance Tests \label{sec:tests}}
 
-XXX moved from Methods, use when discussing 2-D posteriors...
+We first discuss our numerical implementation choices and then test the method's performance using simulated TRILEGAL catalogs. We
+validate adopted luminosity-color models and implied distance scale using SDSS and Gaia observations. An all-sky pipeline implementation
+and its performance are discussed in the next Section.
 
-Another important degeneracy arises from the fact that even for a fixed \teff and \feh, the \logg and thus the luminosity are not uniquely determined by the colors: a degeneracy may exist between the giant branch and the main sequence as the colors constructed from \textit{ugrizy} bands are not sensitive to \logg.
+\subsection{Accelerated Exhaustive Grid Search Method}
+
+Luminosity-color models introduced in \S\ref{sec:method} are expressed as lookup tables, rather than as analytic functions. They are initially defined
+on a two-dimensional grid, and after accounting for dust extinction, on a three-dimensional grid. We use equidistant grids for all three
+model parameters: 0.01 mag for $M_r$, 0.05 dex for $[Fe/H]$ and a range of steps from 0.005 mag to 0.02 mag for $A_r$, depending on its
+prior range. These grids have sufficient resolution for numerical analysis of the posterior probability distributions and typically
+result (depending on the maximum allowed $A_r$ value) in several million grid points. For a given star, exhaustive grid search method simply iterates
+over all these points and for each computes likelihood and ultimately the posterior pdf.
+
+Significant acceleration can be achieved by using a two-step iteration, with degraded grids in the first step, but still fine
+enough to sufficiently resolve posterior pdf to approximately estimate its behavior around its maximum. We find that degrading $M_r$ step to 0.05 mag and
+$[Fe/H]$ step to 0.1 dex works well in practice and results in about 10 times shorter run time. After the first iteration, the range $\pm 3
+\sigma$ around the posterior  maximum, where $\sigma$ is standard deviation, is estimated using marginal distribution for each model
+parameter. The calculation of posterior is then repeated
+using the finest grid but only in the $\pm 3 \sigma$  neighborhood (cube) around posterior maximum, which includes 100 to 1000 times fewer grid locations.
+
+With this acceleration trick, runtime per star is typically about 10 millisec per star  (it varies nearly proportionally to the maximum allowed $A_r$ value). 
+While exceedingly simple, this method is sufficiently fast and robust to offer a convenient practical solution for processing samples of billions of stars.
+
+\subsubsection{Markov Chain Monte Carlo Method}
+
+Markov Chain Monte Carlo methods have been successfully used in this context (e.g., \citealt{2014ApJ...783..114G}). 
+Given that we aim to process LSST photometry for about 10$^{10}$ stars, it is worthwhile to compare MCMC runtime
+to runtime for our accelerated exhaustive grid search method. We considered two implementations, based on {\it PyMC3} \citep{pymc2023} and
+{\it emcee} \citep{2013PASP..125..306F}. The latter was about 10-20 times faster than the former; for example, with 4 ``walkers'' and 500
+iterations, it took on average about 0.1 sec per star. Therefore, in this low-dimensional case (there are only three model parameters) the
+grid search method described above is about an order of magnitude faster than {\it emcee} implementation.
+
+\subsubsection{Neural Network Method}
+
+Further acceleration can be achieved using Neural Network method. In a companion paper by Mrakov\v{c}i\'{c} et al., we describe
+how to use a small subsample of stars to train a neural network model and then estimate model parameters and their uncertainty for the remaining
+stars using trained network. This approach yields runtimes of about 1 milisec per star, that is, shorter by an order of magnitude than when
+the grid search method is used for the full sample. With this additional speed up, a sample of 10 billion stars can be processed in
+about 10 hours using a 300-core machine and pipeline described in the next Section.
+
+\subsection{Performance Testing}
+
+We first validate numerical implementation of our method using a simulated dateset, where the true model parameters are known.
+The luminosity-color model tracks are then validated using SDSS photometry from the so-called Stripe 82 region and trigonometric
+distances obtained using Gaia data. Finally, we demonstrate
+performance in the large $A_r$ regime and three-dimensional mapping of interstellar dust distribution using SDSS
+scans that cross high-extinction regions in the Galactic plane.
+
+\subsubsection{Simulated Data Set}
+
+Our simulated catalogs are based on TRILEGAL simulations of LSST stellar content by \cite{2022ApJS..262...22D}, already
+mentioned in the context of priors in \S\ref{sec:priors}. We use these simulations to generate distributions in the
+$M_r$--$[Fe/H]$--$A_r$--magnitude--sky position space. Given these quantities for a sample of simulated TRILEGAL stars,
+we generate simulated LSST photometry for each star as follows: 
+\begin{enumerate}
+\item
+Given relevant model parameters (e.g., $M_r$ and $[Fe/H]$ for main sequence and red giant stars, $M_r$ and log(g) for
+white dwarfs, the total system luminosity and the component luminosity ratio in the $r$ band for unresolved binaries),
+we use luminosity-color tracks (see Figure~\ref{fig:augmLocus}) to
+assign photometric noise-free colors. We do not use original TRILEGAL
+colors because for proper statistical tests colors must be consistent
+with luminosity-color tracks used for likelihood computations. 
+\item
+Using apparent $r$ band magnitude from TRILEGAL simulation, corresponding to simulated distance and $M_r$, and 
+colors from the previous step, we generate magnitudes in all the remaining bands.
+\item
+We generate per-band photometric uncertainties, $\sigma_b$, using eqs. 4 and 5 from \cite{2019ApJ...873..111I} and
+per-band 5-$\sigma$ depths (the so-called $m_5$) expected for co-added LSST data from \cite{2022ApJS..258....1B}.
+\item
+We draw Gaussian noise from N(0, $\sigma_b$) and add it to each magnitude to obtain ``observed'' magnitudes.   
+\item
+Finally, we recompute photometric uncertainties to be treated as ``observational'' uncertainties using ``observed'' magnitudes
+(because the first set of photometric uncertainties are derived using true noise-free magnitudes). 
+\end{enumerate}
+
+For numerical tests, we select a region from the so-called SDSS Stripe 82, defined by $340^{\deg} < {\rm R.A.} < 350^{\deg}$ and
+$-1.3^{\deg} < \delta < 1.3^{\deg}$, which contains 280,000 simulated main sequence and red giants stars that satisfy
+$r<26$ and $u<27$. Their distributions of model parameters $M_r$, $[Fe/H]$, and interstellar dust extinction along the line of sight in
+the $r$ band, $A_r$,  in the $u$ vs. $g-i$ color-magnitude diagram are shown in Figure~\ref{fig:qpBmeans}.
+
+\begin{figure*}[ht!]
+\hskip -0.3in
+\includegraphics[width=1.08\textwidth,angle=0]{figures/qpBmeans_SDSSpatchRA340-350-simLSST_FeH.png}
+\vskip -0.3in
+\caption{The mean per-pixel values of absolute magnitude $M_r$, metallicity $[Fe/H]$, and interstellar dust extinction along the
+  line of sight in the $r$ band, $A_r$, for a TRILEGAL-simulated sample of 280,000 main
+  sequence and red giants stars with $r<26$, $u<27$, $340^{\degree} < {\rm R.A.} < 350^{\degree}$
+  and $-1.3^{\degree} < \delta < 1.3^{\degree}$ (a small patch from the SDSS Stripe 82 region) in the $u$ vs. $g-i$
+  color-magnitude diagram. The mean values are color-coded according to legend below each panel. Contours
+  visualize the sample distribution in each diagram. The strong variation of $M_r$ with the $g-i$ color is seen in the left panel,
+  except for the small patch with $u<16.5$ and $1.0 < g-i < 1.5$ where red giant stars dominate. The
+  diagonal iso-metallicity boundaries in the middle panel closely correspond to distance
+  (in the range from about 1 kpc to about 100 kpc in the lower left corner). The values of
+  dust extinction, shown in the right panel, are not large ($<0.2$) because the
+  selected field is at high Galactic latitudes (centered on $b = -52^{\degree}$).}
+\label{fig:qpBmeans}
+\end{figure*}
+
+Figures~\ref{fig:perfVSestParamsMr} and \ref{fig:perfVSestParamsFeH} show summarized statistical performance of the Bayesian
+method (bias, scatter, $\chi^2$) in the $u$ vs. $g-i$ color-magnitude diagram. In this test based on a high-galactic
+latitude field, we assumed that all stars are beyond the dust layer and fixed $A_r$ to its known true value (in practice provided by dust
+maps such as those from \citealt{schlegel_maps_1998}). When $A_r$ is considered as a free parameter in sky regions with small $A_r$,
+the best-fit $A_r$ values are often underestimated for stars in the blue part of the stellar locus (because the reddening vector and locus
+are nearly parallel; for more details, see \S2.7.1 in \citealt{2012ApJ...757..166B}). We note that at faint magnitudes probed by LSST, this
+assumption will only break down very close to the Galactic plane.
+
+\begin{figure*}[ht!]
+\hskip -0.2in
+\includegraphics[width=1.06\textwidth,angle=0]{figures/qpB_estQ_SDSSpatchRA340-350-simLSST_Mr.png}
+\vskip -0.3in
+\caption{Statistical performance analysis for the estimates of model
+  parameter $M_r$ (absolute magnitude), as a function of observed $u$
+  band magnitude and the $g-i$ color, for the same simulated sample as shown
+  in Figure~\ref{fig:qpBmeans}. 
+  The left column shows the mean difference per pixel between the true
+  and estimated values, the middle column shows scatter per pixel and
+  the right column shows the scatter normalized by estimated
+  uncertainties ($\chi^2$). Contours visualize the sample distribution in each
+  diagram. Note that for main sequence stars $M_r>4$ and for most
+  stars $4 < M_r < 10$. 
+\label{fig:perfVSestParamsMr}}
+\end{figure*}
+
+\begin{figure*}[h!]
+\hskip -0.2in
+\includegraphics[width=1.06\textwidth,angle=0]{figures/qpB_estQ_SDSSpatchRA340-350-simLSST_FeH.png}
+\vskip -0.3in
+\caption{Analogous to Figure~\ref{fig:perfVSestParamsMr}, except for estimates of $[Fe/H]$ (metallicity) instead of $M_r$. 
+\label{fig:perfVSestParamsFeH}}
+\end{figure*}
+
+As discussed in detail in \cite{2008ApJ...684..287I}, the photometric metallicity is best constrained for blue main-sequence stars with
+$g-i<1$ (equivalently, $0.2 < g − r < 0.6$).
+Therefore, we expect the best statistical behavior (smallest bias and scatter, $\chi^2 \sim 1$) in this color range, which is consistent with
+the behavior seen in Figures~\ref{fig:perfVSestParamsMr} and \ref{fig:perfVSestParamsFeH}. An effective depth limit is about
+$u=25$ (about 1 mag brighter than the 5-$\sigma$ depth in the $u$ band for coadded LSST photometry,  \citealt{2022ApJS..258....1B}); 
+at fainter magnitudes the scatter between true and estimated values
+for both $M_r$ and $[Fe/H]$ rapidly increases.
+
+Figure~\ref{fig:perfVSestParams} shows a more quantitative illustration of the bias and scatter for best-fit $M_r$ and $[Fe/H]$ as
+functions of the $u$ band magnitude, which sets the effective signal-to-noise ratio. In the bright limit ($u < 22$; approximately
+$r<21$ for blue stars), $M_r$ and $[Fe/H]$ can be estimated with uncertainties of 0.10 mag and 0.07 dex, respectively, and essentially
+negligible biases (0.02 mag for $M_r$ and 0.03 dex for $[Fe/H]$). These uncertainties are consistent with simulated photometric errors
+and intrinsic properties of photometric parallax and photometric metallicity methods (for detailed discussion, see \citealt{2008ApJ...684..287I}).
+
+This $M_r$ uncertainty implies distance estimates for blue stars accurate to about 5\% to a distance limit of 25 kpc, with coadded LSST
+photometry.  Such unprecedented accuracy for photometric distance estimates is due
+to the ability of the $u$ band to constrain $[Fe/H]$. For a subsample with $u\sim24.73$, one magnitude brighter than 5-$\sigma$ limit,
+the scatter increases to 0.27 mag and 0.38 dex, and further to 0.36 mag and 0.54 dex for a subsample at 5-$\sigma$ limit in the $u$ band
+(where the $u$ band photometric uncertainty is 0.2 mag).
+
+As the right panel in Figure~\ref{fig:perfVSestParams} demonstrates, estimates of $M_r$ and $[Fe/H]$ are highly anti-correlated (this is essentially a
+consequence of eq. A2 from \citealt{2008ApJ...684..287I}, also visible as the shift of stellar main sequence with metallicity in the top
+panels in Figure~\ref{fig:augmLocus}). An $[Fe/H]$ uncertainty of 0.3 dex, approximately the width of the halo metallicity distribution, 
+would induce an additional uncertainty of estimated photometric $M_r$ of 0.3 mag (as would be the case for faint stars with weak or no $[Fe/H]$ constraint
+coming from their $u$ band measurement).
+
+The bias and scatter for both $M_r$ and $[Fe/H]$ are large in a small patch with $u<16.5$ and $1.0 < g-i < 1.5$ where red giant stars dominate,
+due to strong degeneracies in color space between giants and main sequence stars (recall Figure~\ref{fig:RGdegeneracy}). 
+About 2/3 of all red giants are mis-identified as main sequence stars (due to priors, as they are by and large indistinguishable by colors). 
+Nevertheless, these simulation-based tests suggest that it will be possible to select highly pure samples of red giants using best-fit
+$M_r$ and a simple $M_r < 3$ cut. This selection criterion selects 28\% of all red giants in the sample (which correspond to about 1\% of the full
+sample), with a purity of 99.9\%. With LSST data, it will be possible to select such red giant candidates, if they exist, to
+distances of several hundred kpc (note that low-$[Fe/H]$ main sequence stars at the blue edge of the stellar locus and $u=25$, with
+LSST distance uncertainties of about 20\%, will be detected to distances of about 100 kpc).
+
+\begin{figure*}[h]
+\hskip 0.05in
+\includegraphics[width=0.35\textwidth,angle=0]{figures/dMrVSumag_SDSSpatchRA340-350-simLSST.png}
+\hskip -0.12in
+\includegraphics[width=0.35\textwidth,angle=0]{figures/dFeHVSumag_SDSSpatchRA340-350-simLSST.png} 
+\hskip -0.12in
+\includegraphics[width=0.35\textwidth,angle=0]{figures/dMrVSdFeH_SDSSpatchRA340-350-simLSST.png} 
+\caption{The variation of the difference between the true and estimated values (left: $M_r$, middle: $[Fe/H]$) with $u$
+  band magnitude. The binned background map shows the same simulated sample as in Figure~\ref{fig:qpBmeans}. 
+  The symbols show binned medians and the dashed lines show robust standard deviation around the medians.
+  At about $u=23$, the scatter for both $M_r$ and $[Fe/H]$ starts
+  increasing due to increasing $u$ band measurement error. The dataset
+  was generated assuming $u$ band 5-$\sigma$ limiting depth of 25.73
+  for coadded LSST photometry  \citep{2022ApJS..258....1B}. 
+  The right panel illustrates strong anti-correlation between the differences. 
+\label{fig:perfVSestParams}}
+\end{figure*}
+
+\subsubsection{SDSS-based Comparison to Gaia's Distance Scale \label{sec:SDSSGaia}}
+
+\begin{figure*}[ht!]
+\plotone{gaiaStripe82plot1.png}
+\caption{Analysis of the difference between Gaia's geometric (trigonometric) absolute magnitudes and photometric magnitudes
+  derived from SDSS colors. Symbols show a sample of 33,186 stars with the parallax signal-to-noise ratio above 10, absolute
+  magnitude MrGeo$>$4 and MrPho$>$4 , and SDSS-based $u<21$ and $0.2 < g-r < 0.6$. The median magnitude difference is
+  $-0.01$ mag, with a scatter of 0.28 mag. The four panel illustrate variation of the binned median (red circles) and scatter (red
+  lines) with the $u$-band magnitude, the parallax signal-to-noise ratio, SDSS $g-r$ color, and photometric metallicity.} 
+\label{fig:gaiasdss1}
+\end{figure*}
+
+\begin{figure*}[ht!]
+\plotone{gaiaStripe82plot2.png}
+\caption{Analysis of the difference between Gaia's photometric (color-based) absolute magnitudes and photometric magnitudes
+  derived from SDSS colors. Symbols show a sample of 165,834 stars with absolute magnitude MrPho$>$4, and SDSS-based $u<21$
+  and $0.2 < g-r < 0.6$. The median magnitude difference is $-0.05$ mag, with a scatter of 0.32 mag. The four panel illustrate
+  variation of the binned median (red circles) and scatter (red lines) with the $u$-band magnitude, SDSS $u-g$ and $g-r$ colors,
+  and photometric metallicity.}
+\label{fig:gaiasdss2}
+\end{figure*}
+
+Analysis of the method's performance in the preceding subsection was based on a simulated sample with ``observed'' colors generated using
+exactly the same luminosity-color model tracks as those used in fitting. Therefore, that analysis cannot test validity of those tracks in an absolute sense.
+While these empirical luminosity-color model tracks were derived using SDSS observations of
+globular clusters with known distances and metallicities, here we validate them further using trigonometric distances recently obtained by Gaia.
+
+For validation, we used SDSS Stripe 82 Standard Star Catalog, recalibrated by \cite{2021MNRAS.505.5941T},
+which provides the most precise available photometry in SDSS bands (due to averaging of many repeated SDSS observations). 
+For stars listed in the catalog, we extracted Gaia measurements and ``photogeometric'' distances from 
+\cite{bailer-jones_estimating_2021}. Out of 841,000 stars listed in the Standard Star Catalog, there are 415,000 stars with $r < 22$, $u<22$ and a Gaia
+match within 0.15 arcsec (after correcting for proper motion using Gaia measurements). Their distribution in color-color and
+color-magnitude diagrams is shown in Figures~\ref{fig:3dataDiags} and \ref{fig:3HRdiags}. 
+We estimated $M_r$ and $[Fe/H]$ for these stars used the same procedure and assumptions as for the simulated sample from the
+preceding subsection. In particular, we assumed that all stars are beyond the dust layer and fixed $A_r$ to its value provided by the dust
+maps from \cite{schlegel_maps_1998}.
+
+We first analyze results for a subsample of stars that have $0.2 < g − r < 0.6$, a color range where photometric metallicity estimator is best constrained,
+and a signal-to-noise ratio for Gaia's trigonometric
+parallax measurement above 10.  We also required $u<21$ but that requirement is less stringent than the implied flux limit due to
+Gaia's signal-to-noise ratio limit (approximately, $u<19$). There are about 18,000 stars in this subsample.
+We find that their $M_r$ is estimated using SDSS photometry with a bias of 0.06 mag and root-mean-square scatter of 0.30 mag, compared
+to Gaia's measurements. Equivalently, photometric distances  are estimated with a scatter of 15\% and a bias of 3\%.
+
+The median contribution of Gaia's parallax measurement uncertainty to Gaia's absolute
+magnitude uncertainty for this subsample is about 0.12 mag. This estimate implies that the median uncertainty of SDSS-based $M_r$
+estimates is 0.27 mag (corresponding to about 14\% distance uncertainty). On the other hand, when only stars with signal-to-noise
+ratio for Gaia's trigonometric parallax measurement above 50 are considered, the implied uncertainty of SDSS-based $M_r$
+estimates is 0.21 mag (corresponding to about 10\% distance uncertainty).  Recalling that the statistical uncertainties for $M_r$
+estimates using simulated sample in the preceding section are 0.10 mag at the bright end, these somewhat larger uncertainties may
+imply additional statistical effects not accounted for in our analysis, and/or imperfect luminosity color-tracks.
+Figure~\ref{fig:gaiasdss1} demonstrates that there are no concerningly large systematic errors in photometric $M_r$ estimates with
+respect to photometric noise ($u$ band magnitude), $g-r$ color and photometric metallicity estimate (though note that the metallicity
+range corresponds only to disk stars).
+
+We extended our analysis to fainter magnitude limit by replacing Gaia's magnitudes based on trigonometric distance with those based on
+the so-called ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}. This comparison goes about 2
+magnitudes deeper and also extends to low halo-like metallicity range. The scatter between photometric $M_r$ based on SDSS data and
+$M_r$ based on ``photogeometric'' distances is 0.38 mag. As shown in Figure~\ref{fig:gaiasdss2}, the systematic error
+increases to about 0.2-0.3 at the blue and low-metallicity edge of the stellar locus. When stars with $0.2 < g-r < 0.3$ are excluded,
+systematic errors at the low-metallicity ($[Fe/H]<-1$) end disappear, indicating that it is the $M_r$ as a function of color relation that fails at the blue end,
+rather than the shift of $M_r$ as a function of $[Fe/H]$. Such edge effects will have to be recalibrated when LSST photometry becomes available
+(both using Gaia data and globular clusters).
+
+\subsubsection{Performance in the Galactic Plane}
+
+\begin{figure*}[ht!]
+\hskip -0.2in
+\includegraphics[width=1.06\textwidth,angle=0]{figures/qArSEGUEl110.png}
+\vskip -1.4in
+\caption{Color-coded maps show the best-fit $A_r$ (dust
+  extinction along the line of sight) per pixel for the SDSS-SEGUE strip centered on $l = 110^{\degree}$
+  (modeled after Figure~25 from \citealt{2012ApJ...757..166B}). Each pixel, defined in Galactic coordinates, shows the median Ar
+  according to the legend below each panel.  The three panels  correspond to distance slices 0.3–-1 kpc (left), 1-–1.5 kpc (middle),
+  and 2–-2.5 kpc (right). Note that the median extinction increases rapidly with distance. In the right panel, there are no stars in
+  white regions in the center because extinction makes them fainter than the sample magnitude limit.
+\label{fig:SEGUEdust}}
+\end{figure*}
+
+As the final test, we analyze the method's performance in regions with large interstellar dust extinction ($A_r$). We use SDSS-SEGUE data
+collated and described by \cite{2012ApJ...757..166B}. For illustration, we use a single strip perpendicular to the Galactic
+plane and centered on longitude $l=110^{\degree}$. Figure~\ref{fig:SEGUEdust} shows best-fit dust extinction along the
+line of sight for three distance slices. Clear three-dimensional structure of the dust distribution is evident and consistent with
+Figure~25 from \cite{2012ApJ...757..166B}.
+
+These tests validate numerical implementation of the Bayesian algorithm and demonstrate its performance on a single patch of data, where priors
+can be assumed uniform. An all-sky pipeline implementation is described in the next Section.

### 2024-05-07 paper/discussion.tex
first draft of Section 2
last changes before distributing first draft

--- 2023-11-06 paper/discussion.tex
@@
-% -- popraviti modele
-% -- dovući model s mladim zvijezdama
-% -- dodati modele za WD, BHB, unresolved binaries, RRL i druge populacije
+\section{Discussion and Conclusions \label{sec:disc}}
 
-% Here we provide a layout of our method that simultaneously provides distance estimates as well as the three-dimensional estimation of the dust distribution.
-The distance modulus of a star can be estimated given its absolute and apparent magnitudes and the extinction along the line of sight.
-Therefore, given the observed colors it is necessary to simultaneously estimate the effective temperature, metallicity and extinction.
+Distances to stars are a crucial ingredient in our quest to better understand the formation and evolution
+of the Milky Way galaxy. Anticipating photometric catalogs with tens of billions of stars from Rubin's LSST,
+here we presented PhotoD, a pipeline for computing Bayesian distance estimates that can handle LSST-sized datasets.
+
+The Bayesian procedure implemented in the PhotoD pipeline builds on previous work \citep[e.g.,][]{2011MNRAS.411..435B, 2012ApJ...757..166B,
+2014ApJ...783..114G, green_3d_2019, bailer-jones_estimating_2021} and improves it in several important ways:
+i) the use of multiple stellar populations (in addition to the dominant main-sequence stars)  
+ii) improved color tracks for main-sequence stars and (especially) red giants, including the use of very young ($<$1 Gyr) populations 
+               and an extended $[Fe/H]$ range, and
+iii) priors based on sophisticated TRILEGAL simulations \citep{2022ApJS..262...22D} that include multiple stellar populations and
+       account for all principal structural components of the Galaxy.
+
+Extensive, although still preliminary, testing demonstrates the expected statistical behavior of implemented computations and
+that SDSS-based luminosity-color sequences are supported by more recent direct (trigonometric) distance measurements by Gaia.
+Tests of pipeline performance show that a sample of 10$^{10}$ stars can be processed in a few days using a moderate-size cluster.
+We intend to process each LSST Data Release as they become available and make outputs accessible via Rubin Science Platform.
+
+We anticipate that the accuracy of resulting distance estimates, as well as estimates of metallicity and interstellar dust extinction
+along the line of sight, will improve as LSST advances because:
+\begin{itemize}
+\item As more LSST data are collected, not only that photometric depth will improve but photometric calibration will improve as well.
+\item Photometric light curves will improve identification of variable populations such as quasars, RR Lyrae stars and eclipsing
+           binary stars. When proper motion measurements become available, separation of nearby stars from distant stars will improve, too. 
+\item Luminosity-color sequences will be recalibrated in LSST's photometric system and improved using LSST's own globular cluster data and Gaia's parallax measurements.
+\item The extension of photometric coverage to longer IR wavelengths (e.g., using Euclid and Roman Space Telescope survey photometry)
+          will improve constraints on model parameters, especially in Galactic plane regions with large dust extinction (for more details,
+          see \citealt{2012ApJ...757..166B}).
+\item Improvements in our understanding of the Milky Way structure will iteratively lead to improvements of Galaxy models
+         such as TRILEGAL, and in turn to improvements of Bayesian priors.
+\item In the context of interstellar dust studies, it may be worthwhile to consider hiearchical Bayesian modeling (e.g., setting priors
+             for the line-of-sight reddening profile, as discussed by \citealt{2014ApJ...783..114G}). 
+\end{itemize}
+
+These improvements will require substantial additional work, but given the implied transformative impact of resulting
+distance estimates on our understanding  of the formation and evolution of the Milky Way, it seems well justified.

### 2024-05-07 paper/methodology.tex
last changes before distributing first draft

@@
 \begin{figure*}[t!]
-\plotone{plot3diagsData.png}
-\caption{An illustration of multiple stellar populations. The red dots in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
+\includegraphics[width=1.0\textwidth,angle=0]{figures/plot3diagsData.png} 
+\caption{An illustration of multiple stellar populations. The dots (of
+any color) in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
 (variable sources are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
@@
 in a straightforward manner.}
-by, e.g., \cite{2011MNRAS.411..435B},  \cite{2014ApJ...783..114G},  \cite{green_3d_2019} and \cite{bailer-jones_estimating_2021}.
+by, e.g., \cite{2011MNRAS.411..435B},  \cite{2014ApJ...783..114G}, \cite{lallement_3d_2014}, 
+\cite{gordon_panchromatic_2016}, \cite{queiroz_starhorse_2018}, \cite{green_3d_2019} and \cite{bailer-jones_estimating_2021}.
 The main differences compared to these works include: 
@@
 \begin{figure*}[b!]
-\plotone{plot3HRdiags.png}
+\includegraphics[width=1.0\textwidth,angle=0]{figures/plot3HRdiags.png}  
 \caption{The left panel shows SDSS-based empirical absolute magnitude vs. color parametrization for main-sequence stars and red giants.
-The date are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
+The data are color-coded by metallicity, ranging from $[Fe/H] = -2.5$ to 0 (blue to red). The three lines correspond to three values of
 metallicity: $[Fe/H] = -2, -1, 0$ (dot-dashed, solid and dashed, respectively). The middle panel shows a sample of 63,000 stars that have
@@
 and $g-i \sim 0.8-1.1$). The right panel shows a sample of 415,000 stars with $r < 22$ and $u<22$ as red dots (shown by blue dots in
-Figure~\ref{fig:3dataDiags}, and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots. Their absolute magnitudes were computed
+Figure~\ref{fig:3dataDiags}), and 409,000 of those that also have  $0.2 < g-i < 3.5$ as cyan dots. Their absolute magnitudes were computed
 using the so-called “photogeometric” distances from Bailer-Jones et al. (2021). The dot-dashed, solid and dashed black lines are the
@@
 that SDSS-based photometric distances and Gaia-based ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}
-for main-sequence stars are on the ``same scale''.  XXX ZI: provide numerical results for offsets!
+for main-sequence stars are on the ``same scale''.  We discuss these
+distance scales in more detail in \S\ref{sec:SDSSGaia}.
 
@@
 \begin{figure*}[ht!]
-\plotone{RGdegeneracy.png}
+\includegraphics[width=1.0\textwidth,angle=0]{figures/RGdegeneracy.png}
 \caption{Analysis of color degeneracies between giants and main  sequence stars. The left panel shows minimum color distance in
@@
 \begin{figure*}[ht!] 
-\plotone{plot3diagsBAzoom.png}
+\includegraphics[width=1.0\textwidth,angle=0]{figures/plot3diagsBAzoom.png}
 \caption{Luminosity-color tracks for H and He white dwarfs based on models from \cite{1995PASP..107.1047B}, with
@@
 \begin{figure*}[ht!]
-\plotone{plot3diagsBA_DAh.png}
+\includegraphics[width=1.0\textwidth,angle=0]{figures/plot3diagsBA_DAh.png}  
 \caption{Luminosity-color tracks, shown with gray symbols, for unresolved binaries composed of an M dwarf with
   $[Fe/H]=-1.5$ and an H white dwarf. Analogous tracks for binaries composed of an M dwarf with
-  $[Fe/H]=0$ and either an H or He white dwarf look very similar though not identical. Data, shown as symbols,
+  $[Fe/H]=0$ and either an H or He white dwarf look very similar though not identical. Data, shown as colored symbols,
   are the same as in Figure~\ref{fig:3dataDiags} and in the middle panel in Figure~\ref{fig:3HRdiags}.} 
@@
 
-\subsection{TRILEGAL-based Priors}
+\subsection{TRILEGAL-based Priors \label{sec:priors}}
 
@@
 \end{equation}
-where the model $k=1$ is chosen arbitrarily without a loss of generality ($O_{11}$ = 1). We will illustrate model
-selection with multiple stellar populations in the next Section.
+where the model $k=1$ is chosen arbitrarily without a loss of generality ($O_{11}$ = 1).
+% We will illustrate model selection with multiple stellar populations in the next Section.

### 2024-05-07 paper/paper.tex
new sections 2.2.3 and 2.2.4, and comments from Lovro and Bob
progress...
last changes before distributing first draft

@@
 \usepackage{booktabs}
+\usepackage{gensymb}
+
 \addto\extrasenglish{%
@@
 % Croatian letter "dj"
-\def\d   {{d $\mkern-14.3mu \mathchar'26 $}}
+\def\dj   {{d $\mkern-14.3mu \mathchar'26 $}}
 
@@
 
-\author[0000-0012-3245-123X]{\v{Z}eljko Ivezi\'{c}}
+\author[0000-0001-5250-2633]{\v{Z}eljko Ivezi\'{c}}
 \affiliation{Department of Astronomy and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}
@@
 \affiliation{Faculty of Physics, University of Rijeka, Radmile Matej\v{c}i\'{c} 2, 51000 Rijeka, Croatia}
+
+\author[0000-0000-0000-0000]{Neven Caplar}
+\affiliation{Department of Astronomy and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}
 
@@
 
-%\include{analysis}
+\include{analysis}
 
-%\include{discussion}
+\include{pipeline}
+
+\include{discussion}
 
 \begin{acknowledgements}
-
-This work is financed within the Tenure Track Pilot Programme of the Croatian Science Foundation and the Ecole Polytechnique Fédérale de Lausanne and the Project TTP-2018-07-1171 Mining the variable sky, with the funds of the Croatian-Swiss Research Programme.
-
+This work is financed within the Tenure Track Pilot Programme of the Croatian Science Foundation and the Ecole Polytechnique Fédérale
+de Lausanne and the Project TTP-2018-07-1171 Mining the variable sky, with the funds of the Croatian-Swiss Research Programme. 
 \v{Z}.I. acknowledges funding by the Fulbright Foundation and thanks the Ru\dj er Bo\v{s}kovi\'{c} Institute for hospitality. 
+B.A. acknowledges helpful discussions with Douglas Tucker and the members of the Rubin Observatory Stack Club.
 
@@
 institutions participating in the Gaia Multilateral Agreement.
-
 \end{acknowledgements}
@@
 \end{document}
+
+dust extinction plot in: 
+BayesMethod3D_SEGUElong.ipynb
+
+BDBS:
+BDBS/analyzeBDBS_field_044_TRILEGALcatalog.ipynb
+PhotoD/notebooks/BayesMethod3D_BDBSfield044.ipynb
+
+IBJ2012: fig. 23 
+star with g-i = 3.0 and r =20: DM = 7 mag so d=250 pc
+Mr = 13
+
+10% distance from both Gaia and LSST: fig. 21: 
+sigmaPi = 0.5 mas for both LSST and Gaia
+SNR=10, so pi = 5 mas or D=200 pc OK
+
+within 150 pc: 10% dist for DM=6 and r = 21.5 => Mr = 15.5  (g-i = 3.6) 
+LSST sigmaPi at r=21.5: 0.7 mas so pi = 7 mas or D = 150 pc!
+
+blue star with g-i = 0.3 and r = 20: DM = 16, d=15 kpc
+so Mr = 4
+parallax = 1/15 mas so SNR=0.13
+
+DM = 12: r = 16, d=2.5 kpc, pi = 0.4 
+Gaia sigmaPi = 0.05 mas, SNR = 10 
+LSST pi limited by systematics: sigmaPi = 0.5 so SNR~1
+
+LSST at r = 24: sigmaPi = 3
+for SNR=10, pi = 30 mas and d=30pc DM=2.5 mag
+Mr = 24-2.5 -> Mr = 21.5
+
+For stars with g-i > 3.2 (M5 and later), LSST has better parallax distance
+estimates for r>20
+
+For stars of any color at distances beyond 3 kpc, LSST photo distances
+are better than Gaia distances
+For blue stars at 3 kpc, r = 16 and Gaia's parallax are already 10\%
+between 16 and 20, LSST benefits from FeH which photogeo don't have
+LSST at blue end goes to 100 kpc with 10% distance, correct?

### 2024-05-07 paper/pipeline.tex
for Lovro...

\section{All-sky PhotoD Pipeline Implementation \label{sec:LSDBpipe}}

{\bf Lovro:} will here describe LSDB implementation (with Neven's help) and pipeline performance on the full TRILEGAL catalog...

LSDB = Large Survey Database

\subsection{LSDB and Healpix Implementation}

Implementation of the Bayesian algorithm described in preceding sections assumes that priors can be assumed constant
over some sky region, hereafter a patch (see \S\ref{sec:priors}). In practice,  patches need to be at least several sq. deg.
large in order to have enough stars to compute priors. At the same time, they should not be too large to satisfy the assumption
of constancy. Both of these size limits vary across the sky due to Galactic structure effects.

One of recent tools developed to efficiently loop across the sky patches and perform some per-patch calculations
is the  Large Survey Database...

### 2024-05-28 paper/analysis.tex
comments from coauthors

@@
 prior range. These grids have sufficient resolution for numerical analysis of the posterior probability distributions and typically
-result (depending on the maximum allowed $A_r$ value) in several million grid points. For a given star, exhaustive grid search method simply iterates
+result (depending on the maximum allowed $A_r$ value) in several
+million grid points. For a given star, the exhaustive grid search method simply iterates
 over all these points and for each computes likelihood and ultimately the posterior pdf.
@@
 Significant acceleration can be achieved by using a two-step iteration, with degraded grids in the first step, but still fine
-enough to sufficiently resolve posterior pdf to approximately estimate its behavior around its maximum. We find that degrading $M_r$ step to 0.05 mag and
-$[Fe/H]$ step to 0.1 dex works well in practice and results in about 10 times shorter run time. After the first iteration, the range $\pm 3
-\sigma$ around the posterior  maximum, where $\sigma$ is standard deviation, is estimated using marginal distribution for each model
-parameter. The calculation of posterior is then repeated
-using the finest grid but only in the $\pm 3 \sigma$  neighborhood (cube) around posterior maximum, which includes 100 to 1000 times fewer grid locations.
+enough to sufficiently resolve the posterior pdf to approximately estimate its behavior around its maximum. We find that degrading the
+$M_r$ step to 0.05 mag and the $[Fe/H]$ step to 0.1 dex works well in practice and results in a about ten times shorter run time. After the first iteration, the range $\pm 3
+\sigma$ around the posterior  maximum, where $\sigma$ is standard deviation, is estimated using the marginal distribution for each model
+parameter. The calculation of the posterior is then repeated using the finest grid but only in the $\pm 3 \sigma$  neighborhood
+(cube) around the posterior maximum, which includes 100 to 1000 times fewer grid locations.
 
@@
 Markov Chain Monte Carlo methods have been successfully used in this context (e.g., \citealt{2014ApJ...783..114G}). 
-Given that we aim to process LSST photometry for about 10$^{10}$ stars, it is worthwhile to compare MCMC runtime
-to runtime for our accelerated exhaustive grid search method. We considered two implementations, based on {\it PyMC3} \citep{pymc2023} and
+Given that we aim to process LSST photometry for about 10$^{10}$ stars, it is worthwhile to compare the MCMC runtime
+to the runtime for our accelerated exhaustive grid search method. We considered two implementations, based on {\it PyMC3} \citep{pymc2023} and
 {\it emcee} \citep{2013PASP..125..306F}. The latter was about 10-20 times faster than the former; for example, with 4 ``walkers'' and 500
 iterations, it took on average about 0.1 sec per star. Therefore, in this low-dimensional case (there are only three model parameters) the
-grid search method described above is about an order of magnitude faster than {\it emcee} implementation.
+grid search method described above is about an order of magnitude faster than the {\it emcee} implementation.
 
@@
 
-Further acceleration can be achieved using Neural Network method. In a companion paper by Mrakov\v{c}i\'{c} et al., we describe
+Further acceleration can be achieved using the Neural Network method. In a companion paper by Mrakov\v{c}i\'{c} et al., we describe
 how to use a small subsample of stars to train a neural network model and then estimate model parameters and their uncertainty for the remaining
-stars using trained network. This approach yields runtimes of about 1 milisec per star, that is, shorter by an order of magnitude than when
+stars using the trained network. This approach yields runtimes of about 1 milisec per star, that is, shorter by an order of magnitude than when
 the grid search method is used for the full sample. With this additional speed up, a sample of 10 billion stars can be processed in
-about 10 hours using a 300-core machine and pipeline described in the next Section.
+about 10 hours using a 300-core machine and the pipeline described in the next Section.
 
@@
 we use luminosity-color tracks (see Figure~\ref{fig:augmLocus}) to
-assign photometric noise-free colors. We do not use original TRILEGAL
+assign photometric noise-free colors. We do not use the original TRILEGAL
 colors because for proper statistical tests colors must be consistent
@@
 \item
-Using apparent $r$ band magnitude from TRILEGAL simulation, corresponding to simulated distance and $M_r$, and 
+Using the apparent $r$ band magnitude from the TRILEGAL simulation, corresponding to simulated distance and $M_r$, and 
 colors from the previous step, we generate magnitudes in all the remaining bands.
@@
 
-For numerical tests, we select a region from the so-called SDSS Stripe 82, defined by $340^{\deg} < {\rm R.A.} < 350^{\deg}$ and
-$-1.3^{\deg} < \delta < 1.3^{\deg}$, which contains 280,000 simulated main sequence and red giants stars that satisfy
+For numerical tests, we select a region from the so-called SDSS Stripe 82, defined by $340^{\degree} < {\rm R.A.} < 350^{\degree}$ and
+$-1.3^{\degree} < \delta < 1.3^{\degree}$, which contains 280,000 simulated main sequence and red giants stars that satisfy
 $r<26$ and $u<27$. Their distributions of model parameters $M_r$, $[Fe/H]$, and interstellar dust extinction along the line of sight in
@@
   and $-1.3^{\degree} < \delta < 1.3^{\degree}$ (a small patch from the SDSS Stripe 82 region) in the $u$ vs. $g-i$
-  color-magnitude diagram. The mean values are color-coded according to legend below each panel. Contours
+  color-magnitude diagram. The mean values are color-coded according to the legend below each panel. Contours
   visualize the sample distribution in each diagram. The strong variation of $M_r$ with the $g-i$ color is seen in the left panel,
@@
 
-Figures~\ref{fig:perfVSestParamsMr} and \ref{fig:perfVSestParamsFeH} show summarized statistical performance of the Bayesian
+Figures~\ref{fig:perfVSestParamsMr} and \ref{fig:perfVSestParamsFeH} show the summarized statistical performance of the Bayesian
 method (bias, scatter, $\chi^2$) in the $u$ vs. $g-i$ color-magnitude diagram. In this test based on a high-galactic
@@
 Analysis of the method's performance in the preceding subsection was based on a simulated sample with ``observed'' colors generated using
-exactly the same luminosity-color model tracks as those used in fitting. Therefore, that analysis cannot test validity of those tracks in an absolute sense.
+exactly the same luminosity-color model tracks as those used in fitting. Therefore, that analysis cannot test the validity of those tracks in an absolute sense.
 While these empirical luminosity-color model tracks were derived using SDSS observations of
@@
 
-For validation, we used SDSS Stripe 82 Standard Star Catalog, recalibrated by \cite{2021MNRAS.505.5941T},
+For validation, we used the SDSS Stripe 82 Standard Star Catalog, recalibrated by \cite{2021MNRAS.505.5941T},
 which provides the most precise available photometry in SDSS bands (due to averaging of many repeated SDSS observations). 
@@
 color-magnitude diagrams is shown in Figures~\ref{fig:3dataDiags} and \ref{fig:3HRdiags}. 
-We estimated $M_r$ and $[Fe/H]$ for these stars used the same procedure and assumptions as for the simulated sample from the
+We estimated $M_r$ and $[Fe/H]$ for these stars using the same procedure and assumptions as for the simulated sample from the
 preceding subsection. In particular, we assumed that all stars are beyond the dust layer and fixed $A_r$ to its value provided by the dust
@@
 Figure~\ref{fig:gaiasdss1} demonstrates that there are no concerningly large systematic errors in photometric $M_r$ estimates with
-respect to photometric noise ($u$ band magnitude), $g-r$ color and photometric metallicity estimate (though note that the metallicity
+respect to photometric noise ($u$ band magnitude), $g-r$ color and photometric metallicity estimates (though note that the metallicity
 range corresponds only to disk stars).
 
-We extended our analysis to fainter magnitude limit by replacing Gaia's magnitudes based on trigonometric distance with those based on
+We extended our analysis to fainter magnitude limits by replacing Gaia's magnitudes based on trigonometric distance with those based on
 the so-called ``photogeometric'' distances from \cite{bailer-jones_estimating_2021}. This comparison goes about 2

### 2024-05-28 paper/methodology.tex
comments from coauthors

@@
 
-In most general terms, our aim is to estimate for each star an array (a vector) of model parameters $\vec{\theta}$,
+In the most general terms, our aim is to estimate for each star an array (a vector) of model parameters $\vec{\theta}$,
 for some model $M$ (e.g., main-sequence stars), using data vector $\vec{D}$ and priors for $M$ and $\vec{\theta}$.
@@
 $A_r$ is extinction in the $r$ band and extinction in other bands is proportional to $A_r$, with known constants
-of proportionality (this assumption can be relaxed, see below). Once $M_r$ and $A_r$ are constrained, the distance
-modulus $\Delta$ can be computed from
+of proportionality (this assumption can be relaxed, see below). Once $M_r$ and $A_r$ are constrained, that is, the
+posterior probability distribution $q(Q_r)$, where $Q_r=M_r+A_r$ is known, the distance modulus $\Delta$ can be
+computed using relationship
 \begin{equation}
   \label{eq:distmod}
-                   r = M_r + A_r + \Delta,
+                 r = M_r + A_r + \Delta.
 \end{equation}
-or in a probabilistic form
+In practice the uncertainty of the observed magnitude, $r^{obs}$, is always much smaller than the width of $q(Q_r)$
+distribution (at the bright end by an order of magnitude, $\sim$0.01 mag vs. $\sim$0.1 mag). Therefore,
 \begin{equation}
   \label{eq:distmodpdf}
-                       p(\Delta) = p(r) - p(M_r + A_r). 
+                         p(\Delta) = q(r^{obs} - \Delta),
 \end{equation}
-Note that in practice the uncertainty of the observed magnitude, $r^{obs}$, is always significantly smaller than
-the width of $p(M_r + A_r)$ distribution (at the bright end by an order of magnitude, $\sim$0.01 mag vs. $\sim$0.1 mag).
+with the mean values related as $\mean{\Delta}$ = $r^{obs} - \mean{Q_r}$ and the uncertainty of $\Delta$
+approximately equal to the standard deviation of $q(Q_r)$.
 
@@
 \includegraphics[width=1.0\textwidth,angle=0]{figures/plot3diagsData.png} 
-\caption{An illustration of multiple stellar populations. The dots (of
-any color) in the left panel show color-magnitude diagram for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
-(variable sources are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper motion using Gaia measurements).
-A subset of 415,000 stars with $r < 22$ and $u<22$ are overplotted as blue dots, and 409,000 of those that also have $0.2 < g-i < 3.5$ (dominated by main-sequence stars and red giants) are overplotted as cyan dots.
-Finally, 63,000 stars that have signal-to-noise ratio for Gaia’s parallax measurements of at least 20 are shown as green dots (these stars can be used for the calibration of luminosity-color relations).
-The same symbol color scheme is used in other two panels.
-The three yellow lines in the middle panel show stellar locus parametrization used by Green et al.
+\caption{An illustration of multiple stellar populations.
+  The left panel is a color-magnitude diagram  for 841,000 stars from the SDSS Stripe 82 Standard Star Catalog
+(variable sources are excluded, \citealt{2021MNRAS.505.5941T}) that have Gaia matches within 0.15 arcsec (after correcting for proper
+motion using Gaia measurements). A subset of 415,000 stars with $r < 22$ and $u<22$ are overplotted as blue dots, and 409,000 of those that
+also have $0.2 < g-i < 3.5$ (dominated by main-sequence stars and red giants) are overplotted as cyan dots. Finally, 63,000 stars that have
+a signal-to-noise ratio for Gaia’s parallax measurements of at least
+20 are shown as green dots (these stars can be used for the
+calibration of luminosity-color relations). The same symbol color
+scheme is used in the other two panels. The three yellow lines in the
+middle panel show the stellar locus parametrization used by Green et al.
 (2014) for three values of metallicity (left to right): $[Fe/H] = -2, -1, 0$.
@@
 \item Priors based on sophisticated TRILEGAL simulations by \cite{2022ApJS..262...22D} that include multiple stellar populations and
-               also account for the Galaxy's bulge component. 
+               also account for the Galaxy's bulge component (see
+               \S\ref{sec:priors} below for more details). 
 \end{enumerate}
@@
 using the so-called “photogeometric” distances from Bailer-Jones et al. (2021). The dot-dashed, solid and dashed black lines are the
-same as in the left and middle panels. About 10,000 outliers (about 2.5\% of the full
-sample) seen at $g-i = 0.4$ and $Mr > 7$ are predominantly found at the faint end ($r>20$).}
+same as in the left and middle panels. About 10,000 stars below the main sequence (about 2.5\% of the full sample) seen at $g-i = 0.4$ and
+$Mr > 7$ are predominantly found at the faint end ($r>20$) and may be outliers due to the low photometric signal-to-noise ratio.}
 \label{fig:3HRdiags}
@@
   For $M_r < M_{TO}$, where turn-off absolute magnitude $M_{TO}$ depends on age and ranges from $M_{TO}$=4 for age = 1 Gyr
-  to $M_{TO}$ = 5 for age = 10 Gyr, we ``attach'' model-based isochrones (we use DSED isochrones hereafter) to empirical
+  to $M_{TO}$ = 5 for age = 10 Gyr, we ``attach'' model-based
+  isochrones (we use DSED isochrones hereafter) to the empirical
   SDSS-based isochrones.
@@
 We sample the M dwarf luminosity in the range $8.5 \le M_r \le 14.5$, with a step of 0.1 mag, and for each value
-generate a track by adding luminosity of a white dwarf, where the track is sampled on the same grid of $M_r$ but
+generate a track by adding the luminosity of a white dwarf, where the track is sampled on the same grid of $M_r$ but
 using white dwarf models (separately for H and He models). There are four model families (for two families of white dwarf
@@
 
-Recent studies, such as \cite{2012ApJ...757..166B} based on SDSS data (including the Galactic plane), 
+Studies, such as \cite{2012ApJ...757..166B}, based on SDSS data (including the Galactic plane), 
 find that the scatter of $R_V$ around its mean value $R_V=3.1$ is very small. For this reason, we adopt
@@
 catalogs and provide a more quantitative analysis of the method's performance.  We conclude this Section with a brief
-discussion of Bayesian model selection method for assigning posterior probabilities to each stellar population used to interpret
+discussion of the Bayesian model selection method for assigning posterior probabilities to each stellar population used to interpret
 observations.

### 2024-05-28 paper/paper.tex
comments from coauthors

@@
 \newcommand{\AV}{\ensuremath{A{\mathrm{_V}}}\xspace}
+\newcommand\mean[1]{\overline{#1}}
 
@@
 
-\author[0000-0000-0000-0000]{Dani Chao}
-\affiliation{Ru{\dj}er Bo\v{s}kovi\'{c} Institute, Bijeni\v{c}ka cesta 54, 10000 Zagreb, Croatia }
+% no confirmation of authorship 
+%\author[0000-0000-0000-0000]{Dani Chao}
+%\affiliation{Ru{\dj}er Bo\v{s}kovi\'{c} Institute, Bijeni\v{c}ka cesta 54, 10000 Zagreb, Croatia }
 
@@
 star on a single core for both optimized grid search and MCMC methods; we show in a companion paper by Mrakov\v{c}i\'{c} et al. how to utilize neural
-networks to accelerate this performance by an order of magnitude.  We validate our pipeline, named
+networks to accelerate this performance by up to an order of magnitude.  We validate our pipeline, named
 PhotoD (in analogy with photo-z, photometric redshifts of galaxies)
 using both simulated catalogs, and SDSS, DECam and Gaia photometry. We intend to make LSST-based value-added PhotoD catalogs publicly
-available via Rubin Science Platform with every LSST Data Release. 
+available via the Rubin Science Platform with every LSST Data Release. 
 \end{abstract}
@@
 This work is financed within the Tenure Track Pilot Programme of the Croatian Science Foundation and the Ecole Polytechnique Fédérale
-de Lausanne and the Project TTP-2018-07-1171 Mining the variable sky, with the funds of the Croatian-Swiss Research Programme. 
+de Lausanne and the Project TTP-2018-07-1171 ``Mining the Variable Sky'', with the funds of the Croatian-Swiss Research Programme. 
 \v{Z}.I. acknowledges funding by the Fulbright Foundation and thanks the Ru\dj er Bo\v{s}kovi\'{c} Institute for hospitality. 

### 2024-06-11 paper/analysis.tex
added a few references to the CBJ paper.

@@
 increases to about 0.2-0.3 at the blue and low-metallicity edge of the stellar locus. When stars with $0.2 < g-r < 0.3$ are excluded,
-systematic errors at the low-metallicity ($[Fe/H]<-1$) end disappear, indicating that it is the $M_r$ as a function of color relation that fails at the blue end,
+systematic errors at the low-metallicity ($[Fe/H]<-1$) end disappear\footnote{The so-called ``photogeometric'' distances from \cite{bailer-jones_estimating_2021} do not utilize [Fe/H] information.
+Since the [Fe/H] distribution varies with apparent magnitude, one might expect biased distances (20\% or more for halo vs.
+disk comparison).
+We find no significant bias compared to our estimates, which probably indicates that their distance priors absorb the impact of unknown [Fe/H].
+Hence, the only net effect of unknown [Fe/H] is increased uncertainty of estimated distances (for halo stars, with a metallicity scatter of 0.3 dex, the expected contribution to $M_r$ scatter is about 0.2 mag).}, indicating that it is the $M_r$ as a function of color relation that fails at the blue end,
 rather than the shift of $M_r$ as a function of $[Fe/H]$. Such edge effects will have to be recalibrated when LSST photometry becomes available

### 2024-06-11 paper/methodology.tex
added a few references to the CBJ paper.

@@
 with the mean values related as $\mean{\Delta}$ = $r^{obs} - \mean{Q_r}$ and the uncertainty of $\Delta$
-approximately equal to the standard deviation of $q(Q_r)$.
+approximately equal to the standard deviation of $q(Q_r)$. For related discussion, please see \S2.4 in \cite{bailer-jones_estimating_2021}.
 
@@
 constraints, we opted to utilize recent simulations that implement constraints from a variety of modern surveys (some
-include the Galactic plane and Bulge regions).
+include the Galactic plane and Bulge regions). An analogous approach was taken by \cite{bailer-jones_estimating_2021}.
 
@@
 distribution shifts to fainter levels as the bin center ($r$) becomes fainter. We have experimented with different region
-sizes\footnote{We use HEALPix geometry, see https://healpix.jpl.nasa.gov.} and found that $\sim$10 deg$^2$
+sizes\footnote{Following \cite{bailer-jones_estimating_2021}, we use HEALPix geometry, see https://healpix.jpl.nasa.gov.} and found that $\sim$10 deg$^2$
 (HEALPix $N_{side}$=16 and 32, with 3,072 and 12,288 pixels over the full sky) is sufficiently small to capture the

### 2024-08-27 paper/analysis.tex
Section 4 draft.

@@
 As discussed in detail in \cite{2008ApJ...684..287I}, the photometric metallicity is best constrained for blue main-sequence stars with
-$g-i<1$ (equivalently, $0.2 < g − r < 0.6$).
+$g-i<1$ (equivalently, $0.2 < g - r < 0.6$).
 Therefore, we expect the best statistical behavior (smallest bias and scatter, $\chi^2 \sim 1$) in this color range, which is consistent with
@@
 
-We first analyze results for a subsample of stars that have $0.2 < g − r < 0.6$, a color range where photometric metallicity estimator is best constrained,
+We first analyze results for a subsample of stars that have $0.2 < g - r < 0.6$, a color range where photometric metallicity estimator is best constrained,
 and a signal-to-noise ratio for Gaia's trigonometric

### 2024-08-27 paper/pipeline.tex
comments from coauthors
Section 4 draft.

@@
 
-Implementation of the Bayesian algorithm described in preceding sections assumes that priors can be assumed constant
-over some sky region, hereafter a patch (see \S\ref{sec:priors}). In practice,  patches need to be at least several sq. deg.
-large in order to have enough stars to compute priors. At the same time, they should not be too large to satisfy the assumption
-of constancy. Both of these size limits vary across the sky due to Galactic structure effects.
+Implementation of the Bayesian algorithm described in preceding sections assumes that priors can be assumed constant over some sky region, hereafter a patch (see \S\ref{sec:priors}).
+In practice,  patches need to be at least several sq.
+deg.
+large in order to have enough stars to compute priors.
+At the same time, they should not be too large to satisfy the assumption of constancy.
+Both of these size limits vary across the sky due to Galactic structure effects.
 
-One of recent tools developed to efficiently loop across the sky patches and perform some per-patch calculations
-is the  Large Survey Database...
+One of the recent tools developed to efficiently loop across the sky patches and perform some per-patch calculations is the \href{https://lsdb.readthedocs.io/}{Large Survey Database}\footnote{https://lsdb.readthedocs.io/}.
+
+The combination of the HiPSCat and LSDB workflow are particularly efficient and useful for a science case tackled by the PhotoD.
+HiPSCat effectively partitions the sky in pixels based on the density of the sources, therefore creating pixels that contain roughly the same number of stars.
+This strategy is appropriate both scientifically, as the resolution of the map (area subtended by the pixel) depends on the number of stars within (e.g.
+smaller pixels and higher resolution in the galactic plane and lower in the halo), and computationally because the size of the each chunk that is sent to processing is approximately the same, which therefore optimizes the execution of the code.
+In our particular case, we are using HiPSCat pixels that are a few hundred megabytes in size.
+
+\subsection{Pipeline Performance}
+
+We have performed several benchmarks to estimate the performance of the PhotoD code.
+In our initial tests based on the SDSS Stripe 82 data we found that the distances for a billion stars can be estimated in about 100 days on a Mac M1 Pro laptop using one core (10 milliseconds per star).
+
+In further testing, we used outputs from the TRILEGAL simulation and found that the parallel processing performed with LSDB scaled well with respect to the single core values.
+On a local cluster node with dual AMD EPYC 9474F 48-core processors we were able to achieve per-star figures similar to single-core processing, therefore reducing the required processing time to about a week.
+Our ongoing tests are showing that when all 10 nodes (400 cores) of the Ruđer Bošković Institute's Narval cluster are used, we may be able to estimate distances for about 10 billion stars in about two days.
+
+The output catalog containing LSST photometry (with errors) and minimalistic auxiliary metadata for 10 billion stars are would amount to about 1 TB of data.
+The code outputs, including a covariance matrix of the three fitted parameters, would have approximately the same volume.

### 2024-08-28 paper/pipeline.tex
Mario's LSDB blurbs added.

@@
 
-{\bf Lovro:} will here describe LSDB implementation (with Neven's help) and pipeline performance on the full TRILEGAL catalog...
-
-LSDB = Large Survey Database
-
-\subsection{LSDB and Healpix Implementation}
+% {\bf Lovro:} will here describe LSDB implementation (with Neven's help) and pipeline performance on the full TRILEGAL catalog...
 
@@
 
-One of the recent tools developed to efficiently loop across the sky patches and perform some per-patch calculations is the \href{https://lsdb.readthedocs.io/}{Large Survey Database}\footnote{https://lsdb.readthedocs.io/}.
+The scale of the dataset expected from LSST -- O(10Bn) objects -- makes this
+computation challenging to perform on a single core or computer. Ideally,
+one would wish to execute the computation in parallel on many (hundreds) of cores,
+likely housed on a number of separate nodes. This requires a distributed
+processing framework capable of orchestrating such a solution.
+
+\subsection{LSDB and Healpix Implementation}
+
+For our work, we've adopted the \href{https://lsdb.readthedocs.io/}{\em Large Survey Database} (LSDB; \cite{2024AAS...24326109W})
+framework.  LSDB is a distributed computing framework written in Python,
+which enables multi-node and multi-core computation on large (PB-scale),
+partitioned, catalog datasets.  It extends the concepts of the LSD toolkit
+\citep{2011AAS...21743319J}
+to astronomical catalogs where the distribution of objects on the sky is
+very uneven -- such as is the case for Galactic star counts.  Leveraging
+broadly adopted community libraries such as astropy, Pandas, and Dask, LSDB
+presents a user-friendly API on which it is possible to build our pipeline.
+
+Specifically, LSDB allows us to define a piece of computation (the PhotoD
+function, written in Python) that should be executed either on every object in
+the catalog, or on a per-patch basis (where the patch has a potentially
+dynamically sized solid angle). For example, we compute priors for each of the 
+patches (\S\ref{sec:priors}). LSDB then ensures that these functions are successfully run over the entire (arbitrarily sized) catalog. LSDB handles the data distribution and
+movement between multiple nodes in the cluster, automatically handles
+transient failures, and writes the output (in parallel) to another
+partitioned, searchable, table.
+
+% Using this approach, we were able to process simulated catalogs up to XX in
+% size, using YY nodes, in XX wall-clock hours. Extrapolating to expectations
+% for Rubin's DR1, we calculate the same cluster will be able to repeat this
+% computation on real data in ZZ hours.
+
+%One of the recent tools developed to efficiently loop across the sky patches and perform some per-patch calculations is the \href{https://lsdb.readthedocs.io/}{Large Survey Database}\footnote{https://lsdb.readthedocs.io/}.
 

### 2024-08-28 paper/pipeline.tex
Streamlined Section 4.

@@
 
-The combination of the HiPSCat and LSDB workflow are particularly efficient and useful for a science case tackled by the PhotoD.
-HiPSCat effectively partitions the sky in pixels based on the density of the sources, therefore creating pixels that contain roughly the same number of stars.
-This strategy is appropriate both scientifically, as the resolution of the map (area subtended by the pixel) depends on the number of stars within (e.g.
-smaller pixels and higher resolution in the galactic plane and lower in the halo), and computationally because the size of the each chunk that is sent to processing is approximately the same, which therefore optimizes the execution of the code.
-In our particular case, we are using HiPSCat pixels that are a few hundred megabytes in size.
+This workflow is particularly efficient and well-suited for the scientific objectives addressed by {\em PhotoD}, as it leverages the variable resolution of the map based on the star density within each region.
+For instance, in the galactic plane where star density is higher, smaller pixels provide higher resolution, while in the halo, where star density is lower, larger pixels suffice.
+
+This approach is also computationally advantageous because it maintains consistent chunk sizes for processing.
+Each data chunk, regardless of its spatial resolution, is approximately the same size, optimizing the execution of the code.
+In our specific case, the data chunks correspond to pixels that are a few hundred megabytes in size.
 

### 2024-08-28 paper/paper.tex
Minor edits.

@@
 
-\author[0000-0012-3245-123X]{Karlo Mrakov\v{c}i\'{c}}
+\author[0009-0009-8154-3827]{Karlo Mrakov\v{c}i\'{c}}
 \affiliation{Faculty of Physics, University of Rijeka, Radmile Matej\v{c}i\'{c} 2, 51000 Rijeka, Croatia}
@@
 
-\author[0000-0000-0000-0000]{Connor Yablonski}
+\author[0009-0000-0990-8339]{Connor Yablonski}
 \affiliation{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195, USA}
@@
 
-\author[0000-0000-0000-0000]{Neven Caplar}
+\author[ 	
+0000-0003-3287-5250]{Neven Caplar}
 \affiliation{Department of Astronomy and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}
 
-\author[0000-0000-0000-0000]{Mario Juri\'{c}}
+\author[0000-0003-1996-9252]{Mario Juri\'{c}}
 \affiliation{Department of Astronomy and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}
@@
 \end{document}
-
-dust extinction plot in: 
-BayesMethod3D_SEGUElong.ipynb
-
-BDBS:
-BDBS/analyzeBDBS_field_044_TRILEGALcatalog.ipynb
-PhotoD/notebooks/BayesMethod3D_BDBSfield044.ipynb
-
-IBJ2012: fig. 23 
-star with g-i = 3.0 and r =20: DM = 7 mag so d=250 pc
-Mr = 13
-
-10% distance from both Gaia and LSST: fig. 21: 
-sigmaPi = 0.5 mas for both LSST and Gaia
-SNR=10, so pi = 5 mas or D=200 pc OK
-
-within 150 pc: 10% dist for DM=6 and r = 21.5 => Mr = 15.5  (g-i = 3.6) 
-LSST sigmaPi at r=21.5: 0.7 mas so pi = 7 mas or D = 150 pc!
-
-blue star with g-i = 0.3 and r = 20: DM = 16, d=15 kpc
-so Mr = 4
-parallax = 1/15 mas so SNR=0.13
-
-DM = 12: r = 16, d=2.5 kpc, pi = 0.4 
-Gaia sigmaPi = 0.05 mas, SNR = 10 
-LSST pi limited by systematics: sigmaPi = 0.5 so SNR~1
-
-LSST at r = 24: sigmaPi = 3
-for SNR=10, pi = 30 mas and d=30pc DM=2.5 mag
-Mr = 24-2.5 -> Mr = 21.5
-
-For stars with g-i > 3.2 (M5 and later), LSST has better parallax distance
-estimates for r>20
-
-For stars of any color at distances beyond 3 kpc, LSST photo distances
-are better than Gaia distances
-For blue stars at 3 kpc, r = 16 and Gaia's parallax are already 10\%
-between 16 and 20, LSST benefits from FeH which photogeo don't have
-LSST at blue end goes to 100 kpc with 10% distance, correct?

### 2024-08-29 paper/pipeline.tex
minor fixed in Sec. 4

@@
 computation challenging to perform on a single core or computer. Ideally,
-one would wish to execute the computation in parallel on many (hundreds) of cores,
+one would wish to execute the computation in parallel on many (hundreds of) cores,
 likely housed on a number of separate nodes. This requires a distributed
@@
 
-For our work, we've adopted the \href{https://lsdb.readthedocs.io/}{\em Large Survey Database} (LSDB; \cite{2024AAS...24326109W})
-framework. LSDB is a distributed computing framework written in Python,
+For our work, we have adopted the \href{https://lsdb.readthedocs.io/}{\em Large Survey Database} 
+\citep[LSDB;][]{2024AAS...24326109W} framework. LSDB is a distributed computing framework written in Python,
 which enables multi-node and multi-core computation on large (PB-scale),
@@
 This workflow is particularly efficient and well-suited for the scientific objectives addressed by {\em PhotoD}, as it leverages the variable resolution of the map based on the star density within each region.
-For instance, in the galactic plane where star density is higher, smaller pixels provide higher resolution, while in the halo, where star density is lower, larger pixels suffice.
+For instance, in the Galactic plane where star density is higher, smaller pixels provide higher resolution, while in the halo, where star density is lower, larger pixels suffice.
 
@@
 We have performed several benchmarks to estimate the performance of the PhotoD code.
-In our initial tests based on the SDSS Stripe 82 data we found that the distances for a billion stars can be estimated in about 100 days on a Mac M1 Pro laptop using one core (10 milliseconds per star).
+In our initial tests based on the SDSS Stripe 82 data we found that the distances for a billion stars can be estimated in about 100 days on a Mac M1 Pro laptop using one core (corresponding to 10 milliseconds per star).
 
 In further testing, we used outputs from the TRILEGAL simulation and found that the parallel processing performed with LSDB scaled well with respect to the single core values.
-On a local cluster node with dual AMD EPYC 9474F 48-core processors we were able to achieve per-star figures similar to single-core processing, therefore reducing the required processing time to about a week.
-Our ongoing tests are showing that when all 10 nodes (400 cores) of the Ruđer Bošković Institute's Narval cluster are used, we may be able to estimate distances for about 10 billion stars in about two days.
+On a local cluster node with dual AMD EPYC 9474F 48-core processors, we were able to achieve per-star figures similar to single-core processing, therefore reducing the required processing time to about a week.
+Our ongoing tests are showing that when all 10 nodes (400 cores) of the Ruđer Bošković Institute's Narval cluster are used, we may be able to estimate distances for 10 billion stars in about two days.
 

### 2024-09-01 paper/paper.tex
minor fixed in Sec. 4
Final edits to the TeX.

@@
 \author[0000-0001-5250-2633]{\v{Z}eljko Ivezi\'{c}}
-\affiliation{Department of Astronomy and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}
+\affiliation{DIRAC Institute and the Department of Astronomy, University of Washington, 3910 15th Avenue NE, Seattle, WA 98195, USA}
+
+\author[0000-0003-3287-5250]{Neven \v{C}aplar}
+\affiliation{DIRAC Institute and the Department of Astronomy, University of Washington, 3910 15th Avenue NE, Seattle, WA 98195, USA}
 
@@
 \author[0009-0000-0990-8339]{Connor Yablonski}
-\affiliation{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195, USA}
+\affiliation{DIRAC Institute and the Department of Astronomy, University of Washington, 3910 15th Avenue NE, Seattle, WA 98195, USA}
 
@@
 
-\author[ 	
-0000-0003-3287-5250]{Neven Caplar}
-\affiliation{Department of Astronomy and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}
-
 \author[0000-0003-1996-9252]{Mario Juri\'{c}}
-\affiliation{Department of Astronomy and DiRAC Institute, University of Washington, Box 351580, Seattle, WA 98195, USA}
+\affiliation{DIRAC Institute and the Department of Astronomy, University of Washington, 3910 15th Avenue NE, Seattle, WA 98195, USA}
 
@@
 de Lausanne and the Project TTP-2018-07-1171 ``Mining the Variable Sky'', with the funds of the Croatian-Swiss Research Programme. 
-\v{Z}.I. acknowledges funding by the Fulbright Foundation and thanks the Ru\dj er Bo\v{s}kovi\'{c} Institute for hospitality. 
-B.A. acknowledges helpful discussions with Douglas Tucker and the members of the Rubin Observatory Stack Club.
+Parts were based upon work supported by the National Science Foundation under Grant No. AST-2003196.
 
-Funding for the SDSS and SDSS-II has been provided by the Alfred
+\v{Z}.I.
+acknowledges funding by the Fulbright Foundation and thanks the Ru{\dj}er Bo\v{s}kovi\'{c} Institute for hospitality.
+v{Z}.I, M.J.
+and N.\v{C}.
+acknowledge support from the DiRAC Institute in the Department of Astronomy at the University of Washington.
+B.A.
+acknowledges helpful discussions with Douglas Tucker and the members of the Rubin Observatory Stack Club.
+
+The DiRAC Institute is supported through generous gifts from the Charles and
+Lisa Simonyi Fund for Arts and Sciences and the Washington Research
+Foundation. Funding for the SDSS and SDSS-II has been provided by the Alfred
 P. Sloan Foundation, the Participating Institutions, the National
